---
title: "Lecture 02: Models and Probability"  
date: 'Last updated: `r Sys.Date()`'
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [xaringan-themer.css, custom_style.css]
    seal: TRUE
    nature:
      highlightStyle: magula
      highlightLines: TRUE
      countIncrementalSlides: TRUE
      ratio: '16:9'
---

```{r, include = FALSE}
source(here::here("slides", "before_chunk.R"))
```

## Lecture Outline





---

## A simple experiment

```{r}

data("cars")

set.seed(12345)

cars <- cars %>% 
  sample_n(10) %>% 
  rename("distance" = dist) %>% 
  mutate(
    id = 1:n(),
    speed = speed * 0.44704,
    distance = distance * 0.3048,
    distance = round(distance, digits = 1)
  )

```

.center-vertical[.container[

.pull-left[

```{r}

bob <-
  ggplot(cars) +
  labs(
    x = "Car ID", 
    y = "Distance (m)", 
    title = "Distance to complete stop") +
  scale_x_continuous(breaks = 0:11, limits = c(0, 11.2)
  ) + 
  ylim(-2, max(cars$distance) + 2) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  theme_bw(14) +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(color = "gray75")
  )

bob + 
  geom_segment(
    aes(x = id, xend = id, y = 0, yend = distance),
    linetype = "dashed",
    color = "gray75",
    size = 0.5
  ) +
  geom_point(
    aes(x = id, y = distance),
    size = 2.5
  ) +
  geom_text(
    aes(x = id, y = distance, label = distance),
    vjust = 0,
    nudge_y = 1,
    color = "gray50"
  )

```

]

.pull-right[

We take ten cars, send each down a track, have them brake at the same point, and measure the distance it takes them to stop.  

]

]]


---
count: false

## A simple experiment

.center-vertical[.container[

.pull-left[

```{r}

bob + 
  geom_segment(
    aes(x = id, xend = id, y = 0, yend = distance),
    linetype = "dashed",
    color = "gray75",
    size = 0.5
  ) +
  geom_point(
    aes(x = id, y = distance),
    size = 2.5
    ) +
  geom_text(
    aes(x = id, y = distance, label = distance),
    vjust = 0,
    nudge_y = 1,
    color = "gray50"
  ) +
  annotate(
    geom = "segment", 
    x = 11, y = 0,
    xend = 11, yend = 5,
    size = 0.5,
    linetype = "dashed",
    color = "darkred"
  ) +
  annotate(
    geom = "text", 
    x = 11, y = 6.5,
    label = "?",
    size = 10,
    color = "darkred"
  )

```

]

.pull-right[

We take ten cars, send each down a track, have them brake at the same point, and measure the distance it takes them to stop.  

__Question:__ how far do you think it will take the next car to stop?

]

]]


---
count: false

## A simple experiment

.center-vertical[.container[

.pull-left[

```{r}

bob + 
  geom_point(
    aes(x = id, y = distance),
    size = 2.5
    ) +
  geom_hline(
    aes(yintercept = mean(distance)),
    color = "darkred"
  )

```

]

.pull-right[

We take ten cars, send each down a track, have them brake at the same point, and measure the distance it takes them to stop.  

__Question:__ how far do you think it will take the next car to stop?

This is our __expectation__, $E[y]$, which is equal to the mean, $\mu$.

]

]]



---
count: false

## A simple experiment

.center-vertical[.container[

.pull-left[

```{r}

bob + 
  geom_segment(
    aes(x = id, xend = id, y = distance, yend = mean(distance)),
    linetype = "dashed",
    color = "gray75",
    size = 0.5
  ) +
  geom_point(
    aes(x = id, y = distance),
    size = 2.5
    ) +
  geom_hline(
    aes(yintercept = mean(distance)),
    color = "darkred"
  )

```

]

.pull-right[

We take ten cars, send each down a track, have them brake at the same point, and measure the distance it takes them to stop.  

__Question:__ how far do you think it will take the next car to stop?

This is our __expectation__, $E[y]$, which is equal to the mean, $\mu$.

The difference between what we observe, $y$, and what we expect is the __error__, $\epsilon$.

]

]]


---

## A simple model

.center-vertical[.container[

$$y = E[y] + \epsilon$$

<br>

where

* $E[y] = \mu$, which we estimate using $\bar{y}$  

* $\epsilon \sim N(0, \sigma)$

]]