---
title: "Quantitative Analysis of Archaeological Data"
subtitle: "Lecture 03: Statistical Inference"  
date: 'Last updated: `r Sys.Date()`'
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [xaringan-themer.css, custom_style.css]
    seal: true
    nature:
      highlightStyle: magula
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
---

```{r, include = FALSE, code=xfun::read_utf8(here::here("slides", "before_chunk.R"))}
```

## &#x1F4CB; Lecture Outline


---

## Why statistics?

.pull-rleft[

```{r}

figure("samples_and_populations-0.png")

```

]

.pull-rright[

<br><br>

We want to understand something about a __population__.  

]


---
count: false

## Why statistics?

.pull-rleft[

```{r}

figure("samples_and_populations-1.png")

```

]

.pull-rright[

<br><br>

We want to understand something about a __population__.  

We can never observe the entire population, so we draw a __sample__. 

]



---
count: false

## Why statistics?

.pull-rleft[

```{r}

figure("samples_and_populations-2.png")

```

]

.pull-rright[

<br><br>

We want to understand something about a __population__.  

We can never observe the entire population, so we draw a __sample__. 

We then use a model to __describe__ the sample.

]



---
count: false

## Why statistics?

.pull-rleft[

```{r}

figure("samples_and_populations-3.png")

```

]

.pull-rright[

<br><br>

We want to understand something about a __population__.  

We can never observe the entire population, so we draw a __sample__. 

We then use a model to __describe__ the sample.

By comparing that model to an alternative model, we can __infer__ something about the population.

]

