---
title: "Lecture 05: Visualizing Distributions"  
date: 'Last updated: `r Sys.Date()`'
---

```{r setup}
#| include: false

here::here("slides", "_setup.R") |> source()

penguins <- palmerpenguins::penguins |> na.omit()

titanic <- read_csv("https://raw.githubusercontent.com/wilkelab/SDS375/master/datasets/titanic.csv")

```

## ðŸ“‹ Lecture Outline

- Bar Chart
- Histogram
- Probability Density
- Cumulative Density
- Boxplot

# Bar Chart

## What's it for?

Visualize the amount of some variable across categories, represented using length or height of bars. 

```{r}
#| fig-width:  6

ggplot(penguins, aes(x = fct_infreq(species))) +
  geom_bar(fill = alpha(qcolors("wintergreen"), 0.85)) +
  labs(x = NULL, y = "Count", title = "Palmer Penguins") +
  theme(panel.grid.major.y = element_line(color = "gray95"))

```

## A rule of thumb ðŸ‘

Often easier to read when oriented horizontally.  

```{r}
#| fig-width:  6

ggplot(penguins, aes(y = fct_infreq(species) |> fct_rev())) +
  geom_bar(fill = alpha(qcolors("wintergreen"), 0.85)) +
  labs(x = "Count", y = NULL, title = "Palmer Penguins") +
  theme(panel.grid.major.x = element_line(color = "gray95"))

```

## Grouped data

Grouped bar chart can represent higher dimensional data.

```{r}
#| fig-width:  6

color_values <- c(darken(qcolors("wintergreen"), 0.85), qcolors("wintergreen"))

ggplot(
  penguins, 
  aes(y = fct_infreq(species) |> fct_rev(), fill = sex)
) +
  geom_bar(position = "dodge", linewidth = 1) +
  scale_color_manual(name = NULL, values = color_values) +
  scale_fill_manual(name = NULL, values = alpha(color_values, 0.85)) +
  labs(x = "Count", y = NULL, title = "Palmer Penguins") +
  theme(
    panel.grid.major.x = element_line(color = "gray95"),
    legend.background = element_rect(fill = alpha("white", 0.7)),
    legend.justification = c("right", "bottom"),
    legend.position = c(0.99, 0.05)
  ) +
  guides(fill = guide_legend(reverse = TRUE))

```

_Although this graph is not terribly informative..._

# Histogram

## What's it for?

Visualize the approximate distribution of a continuous random variable.

```{r}
#| fig-width:  6

ggplot(titanic) +
  geom_histogram(
    aes(age),
    binwidth = 5,
    boundary = 0,
    color = darken(qcolors("wintergreen"), 0.2),
    fill = alpha(qcolors("wintergreen"), 0.85)
  ) +
  theme(
    panel.grid.major.y = element_line(color = "gray95")
  ) +
  labs(
    x = "Age (years)",
    y = "Count",
    title = "Passengers of the Titanic"
  )

```

## Bin Width

Obtained by counting the number of observations that fall into each interval or "bin."

```{r}
#| fig-width:  6

counts <- hist(titanic$age, breaks = seq(0, 75, by = 5), plot = FALSE)$counts

ggplot(titanic) +
  geom_histogram(
    aes(age),
    binwidth = 5,
    boundary = 0,
    color = "gray85",
    fill = "gray95",
    alpha = 0.5
  ) +
  geom_point(
    aes(age, runif(nrow(titanic), 0, 15)),
    color = alpha(qcolors("wintergreen"), 0.2),
    size = 4
  ) +
  annotate(
    "segment",
    x = seq(0, 75, by = 5), 
    y = 0,
    xend = seq(0, 75, by = 5), 
    yend = 15,
    linewidth = 1.5,
    color = "gray25"
  ) +
  annotate(
    "segment",
    x = 0, 
    y = 0,
    xend = 75, 
    yend = 0,
    linewidth = 1.5,
    color = "gray25"
  ) +
  annotate(
    "text",
    label = counts,
    x = seq(2.5, 72.5, by = 5),
    y = 20,
    vjust = 0,
    size = 3
  ) +
  theme(
    panel.grid.major.y = element_line(color = "gray95")
  ) +
  labs(
    x = "Age (years)",
    y = "Count",
    title = "Passengers of the Titanic"
  )

```

## âš ï¸ Lookout!

The shape of the distribution depends on the bin width.  

```{r}
#| fig-width: 9
#| fig-asp: 0.4
#| out-width: "100%"

bw2 <- ggplot(titanic) +
  geom_histogram(
    aes(age),
    binwidth = 2,
    boundary = 0,
    color = darken(qcolors("wintergreen"), 0.2),
    fill = alpha(qcolors("wintergreen"), 0.85)
  ) +
  theme(
    panel.grid.major.y = element_line(color = "gray95"),
    axis.title.x = element_blank()
  ) +
  labs(
    x = "Age (years)",
    y = "Count",
    title = "Binwidth = 2"
  )

bw5 <- ggplot(titanic) +
  geom_histogram(
    aes(age),
    binwidth = 5,
    boundary = 0,
    color = darken(qcolors("wintergreen"), 0.2),
    fill = alpha(qcolors("wintergreen"), 0.85)
  ) +
  theme(
    panel.grid.major.y = element_line(color = "gray95"),
    axis.title.y = element_blank()
  ) +
  labs(
    x = "Age (years)",
    y = "Count",
    title = "Binwidth = 5"
  )

bw10 <- ggplot(titanic) +
  geom_histogram(
    aes(age),
    binwidth = 10,
    boundary = 0,
    color = darken(qcolors("wintergreen"), 0.2),
    fill = alpha(qcolors("wintergreen"), 0.85)
  ) +
  theme(
    panel.grid.major.y = element_line(color = "gray95"),
    axis.title = element_blank()
  ) +
  labs(
    x = "Age (years)",
    y = "Count",
    title = "Binwidth = 10"
  )

bw2 + bw5 + bw10 + theme(plot.margin = margin(5,5,5,5))

```

## A rule of thumb ðŸ‘

:::::: {.r-stack}
::: {.fragment .fade-out fragment-index=0 style="margin: 0 auto auto 0;"}

Generally, a bad idea to use stacked or dodged groupings in a single histogram.

```{r}

ggplot(titanic) +
  geom_histogram(
    aes(age, fill = sex),
    binwidth = 5,
    boundary = 0,
    color = darken(qcolors("wintergreen"), 0.2)
  ) +
  scale_fill_manual(name = NULL, values = alpha(color_values, 0.85)) +
  theme(
    panel.grid.major.y = element_line(color = "gray95")
  ) +
  labs(
    x = "Age (years)",
    y = "Count",
    title = "Passengers of the Titanic"
  )

```

:::
::: {.fragment .fade-in fragment-index=0 style="margin: 0 auto auto 0;"}

Better to use facets.

```{r}

ggplot(titanic) +
  geom_histogram(
    aes(age, fill = sex),
    binwidth = 5,
    boundary = 0,
    color = darken(qcolors("wintergreen"), 0.2)
  ) +
  scale_fill_manual(name = NULL, values = alpha(color_values, 0.85)) +
  facet_wrap(vars(sex)) +
  theme(
    panel.grid.major.y = element_line(color = "gray95"),
    strip.background = element_blank(),
    strip.text = element_blank(),
    legend.position = "bottom"
  ) +
  labs(
    x = "Age (years)",
    y = "Count",
    title = "Passengers of the Titanic"
  )

```

:::
::::::

# Probability Density

## What's it for?

Visualize the approximate distribution of a continuous random variable.  

```{r}
#| fig-width:  6

ggplot(titanic) +
  geom_density(
    aes(age),
    color = darken(qcolors("wintergreen"), 0.2),
    fill = alpha(qcolors("wintergreen"), 0.85)
  ) +
  theme(
    panel.grid.major.y = element_line(color = "gray95")
  ) +
  labs(
    x = "Age (years)",
    y = "Density",
    title = "Passengers of the Titanic"
  )

```

## Kernel Density Estimation (KDE)

:::::: {.columns}
::: {.column .mt-3 width="45%"}

Procedure:

1. Define a kernel, often a normal distribution with mean equal to the observation.
2. Define bandwidth for scaling the kernel.
2. Sum the kernels.

:::
::: {.column width="55%"}

```{r}

x <- c(4, 19, 24, 28, 31, 36, 58)

bw <- density(titanic$age)$bw

get_kernel <- function(x, bw, n, size = 50){
  
  kernel <- density(x, bw = bw, n = size)
  
  tibble(
    barx = x,
    x = kernel$x,
    y = kernel$y/n
  )
  
}

# n should technically be the sample size, nrow(titanic), but that's harder to visualize
xy <- lapply(x, get_kernel, bw = bw, n = 20)

xy <- bind_rows(xy)

ggplot() +
  geom_density(
    data = titanic,
    aes(age),
    color = "gray85",
    fill = "gray95",
    alpha = 0.5
  ) +
  geom_line(
    data = xy,
    aes(x, y, group = barx),
    color = darken(qcolors("wintergreen"), 0.2),
    linewidth = 1
  ) +
  geom_point(
    aes(x = x, y = 0.0002),
    shape = "|",
    color = darken(qcolors("wintergreen"), 0.2),
    size = 2
  ) +
  theme(
    panel.grid.major.y = element_line(color = "gray95")
  ) +
  labs(
    x = "Age (years)",
    y = "Density",
    title = "Passengers of the Titanic"
  ) +
  xlim(0, 75)

```

:::{style="width:85%;margin-left:auto;"}
_The kernels in this figure are not to scale._
:::

:::
::::::

## Grouped data

There's not a simple answer for how to plot multiple KDE's, but facets are your friend.

```{r}

ggplot(titanic) +
  geom_density(
    aes(age, fill = sex),
    color = darken(qcolors("wintergreen"), 0.2)
  ) +
  scale_fill_manual(name = NULL, values = alpha(color_values, 0.85)) +
  facet_wrap(vars(sex)) +
  theme(
    panel.grid.major.y = element_line(color = "gray95"),
    strip.background = element_blank(),
    strip.text = element_blank(),
    legend.position = "bottom"
  ) +
  labs(
    x = "Age (years)",
    y = "Density",
    title = "Passengers of the Titanic"
  )

```

# Cumulative Density

## What's it for?

Visualize the approximate distribution of a continuous random variable _without having to specify a bandwidth_.

```{r}
#| fig-width:  6

ggplot(titanic) +
  stat_ecdf(
    aes(age),
    color = darken(qcolors("wintergreen"), 0.2)
  ) +
  theme(
    panel.grid.major.y = element_line(color = "gray95")
  ) +
  labs(
    x = "Age (years)",
    y = "Probability",
    title = "Passengers of the Titanic"
  )

```

## Procedure

:::::: {.columns}
::: {.column width="45%"}

Consider this sample: 

`(0.3, 2.0, 3.4, 1.2, 2.2, 1.9)`. 

To calculate its eCDF, we divide the number of observations that are less than or equal to each unique value by the total sample size. 

```md
0.0 -> 0/6 = 0.00
0.3 -> 1/6 = 0.17
1.2 -> 2/6 = 0.33 
1.9 -> 3/6 = 0.50 
2.0 -> 4/6 = 0.67 
2.2 -> 5/6 = 0.83
3.4 -> 6/6 = 1.00
```

:::
::: {.column width="55%"}

```{r}

example <- tibble(
  x = c(0, 0.3, 1.2, 1.9, 2.0, 2.2, 3.4),
  y = c(0, 0.17, 0.33, 0.5, 0.67, 0.83, 1)
)

ggplot(example) +
  geom_step(
    aes(x, y),
    color = darken(qcolors("wintergreen"), 0.2),
    linewidth = 1
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  labs(
    x = NULL,
    y = "Probability"
  )

```

:::
::::::

## âš ï¸ Lookout!

These are a little bit harder to interpret. Gives the probability of being less than or equal to x. E.g., the probability of being 28 years old or younger is 0.5.

```{r}
#| fig-width:  6

titanic_ecdf <- titanic |> 
  group_by(age) |> 
  summarize(count = n()) |> 
  arrange(age) |> 
  mutate(
    total = sum(count),
    ecd = cumsum(count)/total
  )

ggplot() +
  geom_rect(
    data = titanic_ecdf |> filter(ecd <= 0.53),
    aes(xmin = age,
        xmax = ifelse(is.na(lead(age)), max(age), lead(age)),
        ymin = 0,
        ymax = ecd),
    fill = alpha(qcolors("wintergreen"), 0.85)
  ) +
  geom_rect(
    data = titanic_ecdf |> filter(ecd >= 0.51),
    aes(xmin = age,
        xmax = ifelse(is.na(lead(age)), max(age), lead(age)),
        ymin = 0,
        ymax = ecd),
    fill = alpha("gray90", 0.85)
  ) +
  geom_step(
    data = titanic_ecdf,
    aes(age, ecd),
    color = darken(qcolors("wintergreen"), 0.2)
  ) +
  theme(
    panel.grid.major.y = element_line(color = "gray95")
  ) +
  labs(
    x = "Age (years)",
    y = "Probability",
    title = "Passengers of the Titanic"
  )

```

# Boxplot

## What's it for?

Visualize the approximate distribution of a continuous random variable using its quartiles.  

```{r}
#| fig-width:  6

ggplot(penguins, aes(species, body_mass_g/1000)) +
  stat_boxplot(geom ='errorbar', width = 0.33)  +
  geom_boxplot(
    fill = qcolors("wintergreen"), 
    notch = TRUE,
    notchwidth = 0.75
  ) +
  labs(
    x = NULL,
    y = "Body Mass (kg)",
    title = "Palmer Penguins"
  ) +
  theme(
    panel.grid.major.y = element_line(color = "gray95")
  )

```

Useful for plotting distributions across multiple groups.  

## Quartiles

When the data are ordered from smallest to largest, the quartiles divide them into four sets of more-or-less equal size. The second quartile is the median!

```{r}
#| fig-align: center

set.seed(42)

y <- rnorm(250)

labels <- tibble(
  y = c(boxplot.stats(y)$stats, max(y)),
  x = 0.5,
  label = c("1.5 x IQR", "first quartile", "median", "third quartile", "1.5 x IQR", "outlier")
)

ggplot(tibble(x = 0, y), aes(x, y)) + 
  stat_boxplot(geom ='errorbar', width = 0.33) +
  geom_boxplot(
    fill = qcolors("wintergreen"), 
    width = 0.6,
    outlier.size = 4,
    notch = TRUE,
    notchwidth = 0.75
  ) +
  geom_text(
    data = labels,
    aes(x, y, label = label), 
    hjust = 0,
    size = 11/.pt
  ) +
  geom_point(
    data = tibble(x = runif(length(y), -1.2, -0.5), y = y),
    aes(x, y),
    size = 3,
    color = darken(qcolors("wintergreen"), 0.2),
    alpha = 0.85
  ) +
  coord_cartesian(xlim = c(-2.2, 3)) +
  theme_void()

```

## A rule of thumb ðŸ‘

Sometimes easier to read when oriented horizontally.  

```{r}
#| fig-width:  6

ggplot(penguins, aes(body_mass_g/1000, species)) +
  stat_boxplot(geom ='errorbar', width = 0.33) +
  geom_boxplot(
    fill = qcolors("wintergreen"), 
    notch = TRUE,
    notchwidth = 0.75
  ) +
  labs(
    x = "Body Mass (kg)",
    y = NULL,
    title = "Palmer Penguins"
  ) +
  theme(
    panel.grid.major.x = element_line(color = "gray95")
  )

```