---
title: "Lecture 07: Linear Models 4"  
date: 'Last updated: `r Sys.Date()`'
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [xaringan-themer.css, custom_style.css]
    seal: true
    nature:
      highlightStyle: magula
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
---

```{r, include = FALSE, code=xfun::read_utf8(here::here("slides", "before_chunk.R"))}
```

```{r}

library(broom)

adverts <- here::here("datasets", "advertising.csv") %>% read_csv()

adverts_lm <- lm(sales ~ tv, data = adverts)

estimate <- augment(adverts_lm)

predictors <- tidy(adverts_lm)

beta0 <- predictors %>% filter(term == "(Intercept)") %>% pull(estimate)
beta1 <- predictors %>% filter(term == "tv") %>% pull(estimate)

```

## &#x1F4CB; Lecture Outline

1. Simple Linear Regression
2. Multiple Linear Regression
3. Multicollinearity
    - Pearson's Correlation Matrix
    - Variance Inflation Factor
4. ANOVA
5. Marginal Response Plot

---

## Simple Linear Regression

.pull-left[

```{r, fig.asp = 0.9}

ggplot(estimate) +
  geom_point(
    aes(tv, sales),
    size = 2.5
  ) +
  labs(
    x = "Investment in TV Advertising\n(thousands of dollars)",
    y = "Sales\n(thousands of units)"
  )

```

.center[_From [James et al (2021). ISLR. 2nd Ed.](https://www.statlearning.com/)_]

]

.pull-right[

```{r}

adverts %>% 
  kbl(
    table.attr = "class='table-model table-fullwidth'",
    row.names = 1:nrow(adverts)
  ) %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover")
  ) %>% 
  scroll_box(height = "4in")

```

]



---

## Simple Linear Regression

.pull-left[

```{r, fig.asp = 0.9}

ggplot(estimate) +
  geom_segment(
    aes(tv, sales, xend = tv, yend = .fitted),
    color = "gray60"
  ) +
  geom_abline(
    intercept = beta0,
    slope = beta1,
    color = qaad_colors("rufous_red"),
    size = 2
  ) +
  geom_point(
    aes(tv, sales),
    size = 3
  ) +
  labs(
    x = "Investment in TV Advertising\n(thousands of dollars)",
    y = "Sales\n(thousands of units)"
  )

```

.center[_From [James et al (2021). ISLR. 2nd Ed.](https://www.statlearning.com/)_]

]

.pull-right[

```{r}

predictors %>%
  rename(
        "t.statistic" = statistic
  ) %>% 
  kbl(
    table.attr = "class='table-model table-fullwidth'"
  ) %>% 
  kable_styling(
    bootstrap_options = c("condensed", "hover")
  )

```

__Note the units!__ We interpret the slope coefficient here as saying, "For every $1000 spent on television advertising, sales increased by approximately 50 units."

__Question__ What does the intercept mean?

]

---
## But, muh data...

__Question__ What about the money invested in radio and newspaper advertising? How did that affect sales?

--

_How would we answer this?_  

We could build three separate models:
1. sales $\sim$ tv
2. sales $\sim$ newspaper
3. sales $\sim$ radio

But, now, suppose we invest $30k in television, $10k in newspaper, and $5k in radio advertising. How will this increase sales? Should we just add the models together? But, this assumes that the coefficient estimates are independent of each other... 

--

__Solution__ Make one big model!  

1. sales $\sim$ tv + newspaper + radio



---

## Multiple Linear Regression

```{r}

adverts_lm_list <- list(
  tv = adverts_lm,
  radio = lm(sales ~ radio, data = adverts),
  newspaper = lm(sales ~ newspaper, data = adverts),
  all = lm(sales ~ tv + radio + newspaper, data = adverts)
)

adverts_lm_table <- adverts_lm_list %>% 
  lapply(tidy) %>% 
  bind_rows() %>% 
  mutate(
    model = c(rep("simple", 6), rep("multiple", 4))
  ) %>% 
  rename(
    "t.statistic" = statistic
  ) %>% 
  select(model, everything())

```

.pull-left[

```{r, fig.asp = 0.75}

adverts_lm_table %>% 
  filter(term != "(Intercept)") %>% 
  ggplot(
    aes(estimate, term, color = model)
  ) +
  geom_vline(
    xintercept = 0,
    color = qaad_colors("rufous_red"),
    size = 0.4
  ) +
  geom_errorbarh(
    aes(xmin = estimate - std.error, xmax = estimate + std.error),
    height = 0.25,
    size = 1
  ) +
  geom_point(
    size = 4
  ) +
  scale_color_manual(
    values = c(qaad_colors("dark_purple"), "transparent")
  ) +
  labs(
    x = "Coefficient Estimates"
  ) +
  theme(
    axis.title.y = element_blank(),
    legend.position = "none"
  )

```

]

.pull-right[

Coefficients are estimated in the same way, using OLS to minimize the Residual Sum of Squares.

$$RSS = \sum_{i = 1}^{n} (y_i - \hat{y_i})^2$$

But the __interpretation__ changes! Consider 

$$\beta_{radio}=0.189$$ 

This tells us, "For some amount of `tv` and `newspaper` advertising, spending an additional $1000 on `radio` advertising increases sales by approximately 189 units."

]



---

## Multiple Linear Regression

.pull-left[

```{r, fig.asp = 0.75}

adverts_lm_table %>% 
  filter(term != "(Intercept)") %>% 
  ggplot(
    aes(estimate, term, color = model)
  ) +
  geom_vline(
    xintercept = 0,
    color = qaad_colors("rufous_red"),
    size = 0.4
  ) +
  geom_errorbarh(
    aes(xmin = estimate - std.error, xmax = estimate + std.error),
    height = 0.25,
    size = 1
  ) +
  geom_point(
    size = 4
  ) +
  scale_color_manual(
    values = qaad_colors("dark_purple", "maize_crayola")
  ) +
  labs(
    x = "Coefficient Estimates"
  ) +
  theme(
    axis.title.y = element_blank(),
    legend.position = c(0.85, 0.87),
    legend.box.just = c("right", "top"),
    legend.title = element_blank()
  )
  
```

]

.pull-right[

Notice, when we compare these results to the simple linear models, that estimates are the same for `tv` and `radio` but change dramatically for `newspaper`. In fact, the estimate for `newspaper` is no longer significantly different than zero!

__Question__ Why might this be?

]



---

## Collinearity

.pull-left[

```{r, fig.asp = 1, out.width = "90%"}

pnt_matrix <- lapply(
  seq_along(adverts),
  function(x){
    
    n.var <- ncol(adverts)
    n.obs <- nrow(adverts)
    
    s.var <- names(adverts)[[x]]
    
    var1 <- adverts[x]
    
    names(var1) <- "x"
    
    var1 <- var1 %>% 
      slice(rep(1:n.obs, n.var)) %>% 
      mutate(var1 = s.var)
    
    var2 <- adverts %>% 
      pivot_longer(
        everything(),
        names_to = "var2",
        values_to = "y"
      ) %>% 
      arrange(var2)
    
    bind_cols(var1, var2) %>% 
      select(var1, var2, x, y)
    
  }
) %>% 
  bind_rows() %>% 
  mutate(
    across(c(var1, var2), factor),
    across(c(var1, var2), ~forcats::fct_relevel(.x, "sales", "tv", "radio")),
  )

lbl_matrix <- pnt_matrix %>% 
  group_by(var1, var2) %>% 
  summarize(
    x = 0.5 * (min(x) + max(x)),
    y = 0.5 * (min(y) + max(y))
  ) %>% 
  filter(
    var2 == var1
  )

cor_matrix <- cor(adverts) %>% 
  round(digits = 2) %>% 
  as.data.frame() %>% 
  mutate(
    var1 = rownames(.)
  ) %>% 
  as_tibble() %>% 
  tidyr::pivot_longer(
    -var1, 
    names_to = "var2", 
    values_to = "value"
  ) %>% 
  mutate(
    across(c(var1, var2), factor),
    across(c(var1, var2), ~forcats::fct_relevel(.x, "sales", "tv", "radio")),
    color = ifelse(value > 0.5 | is.na(value), "black", "white")
  )

bob_matrix <- pnt_matrix %>% 
  group_by(var1, var2) %>% 
  summarize(
    xmid = 0.5 * (min(x) + max(x)),
    ymid = 0.5 * (min(y) + max(y)),
    xmin = min(x),
    xmax = max(x),
    ymin = min(y),
    ymax = max(y)
  ) %>% 
  filter(
    !(var2 == var1),
    !(var2 == "sales"),
    !(var2 == "tv" & var1 %in% c("radio", "newspaper")),
    !(var2 == "radio" & var1 == "newspaper")
  )

cor_matrix <- left_join(bob_matrix, cor_matrix, by = c("var1", "var2"))

pnt_matrix <- pnt_matrix %>% 
  filter(
    !(var2 == var1),
    !(var2 == "newspaper"),
    !(var2 == "radio" & var1 %in% c("sales", "tv")),
    !(var2 == "tv" & var1 == "sales")
  )

ggplot() +
  geom_point(
    data = pnt_matrix, 
    aes(x, y),
    size = 1.8,
    shape = 21,
    color = qaad_colors("rufous_red"),
    fill = alpha(qaad_colors("maize_crayola"), 0.7)
  ) +
  geom_smooth(
    data = pnt_matrix, 
    aes(x, y),
    method = "lm",
    se = FALSE,
    color = qaad_colors("celadon_blue")
  ) +
  geom_text(
    data = lbl_matrix,
    aes(x, y, label = var1),
    size = 7
  ) +
  geom_rect(
    data = cor_matrix,
    aes(
      fill = value,
      xmin = xmin,
      xmax = xmax,
      ymin = ymin,
      ymax = ymax
    )
  ) +
  scale_fill_viridis(
    option = "magma",
    na.value = "white"
  ) +
  geom_text(
    data = cor_matrix,
    aes(
      xmid,
      ymid,
      label = value, 
      color = color
    ),
    hjust = 0.5,
    vjust = 0.5,
    size = 7
  ) +
  scale_color_manual(
    values = c("black", "white")
  ) +
  facet_grid(var2 ~ var1, scales = "free") +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    legend.position = "none",
    strip.background = element_blank(),
    strip.text = element_blank()
  ) +
  coord_cartesian(expand = FALSE)

```

]

.pull-right[

We see here that `newspaper` and `radio` advertising are correlated (\\(\rho\\) = 0.35). So, markets with large investments in one tend to have large investments in the other, too.  

This tells us that `newspaper` advertising was never really increasing `sales`. It was just taking "credit" for the contributions that `radio` advertising was making.

]


---
name: bob

## Multicollinearity

.pull-left-70[

__Problem__ It is possible for multiple variables to correlate with each other even if no single pair of variables do. Called _multicollinearity_, this cannot be observed by inspecting the correlation matrix.

__Solution__ Measure the Variance Inflation Factor for each covariate.

$$VIF(\hat{B_j}) = \frac{1}{1-R^{2}_{j}}$$

where $R^{2}_{j}$ is the coefficient of determination for a model of a covariate, $x_j$, regressed against all the other covariates, $x_j \sim x_1 + x_2 + \ldots + x_i$.

_Variance Inflation_ inflates the standard errors, too!

$$se(\hat{B_j}) = \sqrt{Var(\hat{B_j}) \cdot VIF(\hat{B_j})}$$

]



---
template: bob

.pull-right-28[

```{r}

library(car)

vif_vct <- vif(adverts_lm_list$all)

tibble(
  covariate = names(vif_vct),
  VIF = unname(vif_vct)
) %>% 
  kbl(
    table.attr = "class='table-model table-fullwidth'",
    row.names = 1:nrow(adverts)
  ) %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover")
  )

```

VIF close to 1 means no mc.  

__Rule of Thumb__ VIF that exceeds 5 or 10 is too much! 

]


---

## Multicollinearity

.pull-left-70[

Strategies for handling multicollinearity include:

- Removing one or more correlated variables (simplest strategy)
- Aggregating correlated variables (e.g., product or mean)
- Using an interaction term (more later)
- Using an ordination technique (more later)

]

.pull-right-28[

```{r}

library(car)

vif_vct <- vif(adverts_lm_list$all)

tibble(
  covariate = names(vif_vct),
  VIF = unname(vif_vct)
) %>% 
  kbl(
    table.attr = "class='table-model table-fullwidth'",
    row.names = 1:nrow(adverts)
  ) %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover")
  )

```

VIF close to 1 means no mc.  

__Rule of Thumb__ VIF that exceeds 5 or 10 is too much! 

]



---

## ANOVA for model

.pull-left[

We have two models of our data:
- $S:$ sales $\sim$ tv
- $C:$ sales $\sim$ tv + newspaper + radio

__Question:__ Does the complex model (C) explain more variance than the simple model (S)? Does the difference arise by chance?

Here are our  hypotheses:
- $H_0:$ no difference in variance explained.
- $H_1:$ difference in variance explained.

]

.pull-right[

```{r}

adverts_anova <- with(adverts_lm_list, anova(tv, all))

adverts_anova <- adverts_anova %>% 
  tidy() %>% 
  rename(
    "RSS" = rss,
    "F" = statistic
  ) %>% 
  mutate(
    model = c("S", "C"),
    p.value = ifelse(
      p.value < 2.2e-16, 
      "< 2.2e-16", 
      as.character(signif(p.value, 3))
    )
  ) %>% 
  rename(
    "Pr(>F)" = p.value
  ) %>% 
  select(model, everything())

adverts_anova %>% 
  kbl(
    align = "c",
    table.attr = "class='table-model table-fullwidth'"
  ) %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover")
  )

```

<br>

$$F = \frac{(RSS_{S} - RSS_{C})/(df_S - df_C)}{RSS_{C}/df_C}$$

]






---

## Marginal Response Plot

```{r}

n <- 300

radio <- with(adverts, seq(min(radio), max(radio), length = n))
              
new_data <- expand.grid(
  tv = with(adverts, seq(min(tv), max(tv), length = n)),
  radio = radio
) %>% 
  as_tibble() %>% 
  mutate(
    newspaper = mean(adverts$newspaper)
  )

new_data$sales <- predict(adverts_lm_list$all, newdata = new_data)

margins <- with(adverts, quantile(tv, c(0.25, 0.5, 0.75)))
margins <- unname(margins)
margins <- round(margins, digits = 2)

responses <- lapply(
  margins, 
  function(z){
    
    nd <- tibble(
      radio = radio,
      tv = z,
      newspaper = mean(adverts$newspaper)
    )
    
    tibble(
      tv = z,
      radio = radio,
      sales = predict(adverts_lm_list$all, newdata = nd)
    )
    
  }) %>% 
  bind_rows() %>% 
  mutate(
    tv = factor(tv)
  )
  
```

```{r}

response_tile <- ggplot() +
  geom_contour_filled(
    data = new_data, 
    aes(radio, tv, z = sales),
    color = "transparent",
    bins = 5
  ) +
  scale_fill_viridis(
    name = "Sales\n(thousands of units)",
    labels = paste0("< ", seq(6, 30, by = 6)),
    discrete = TRUE
  ) +
  geom_hline(
    aes(
      yintercept = margins
    ),
    color = "white",
    size = 3
  ) +
  geom_hline(
    aes(
      yintercept = margins,
      color = factor(margins)
    ),
    size = 1.5
  ) +
  scale_color_manual(
    name = NULL,
    values = qaad_colors("dark_purple", "celadon_blue", "flame_orange")
  ) +
  labs(
    x = "Investment in Radio Advertising\n(thousands of dollars)",
    y = "Investment in TV Advertising\n(thousands of dollars)"
  ) +
  coord_cartesian(
    expand = FALSE
  ) +
  theme(
    legend.position = "top",
    legend.direction = "horizontal",
    legend.title.align = 0.5
  ) +
  guides(
    color = "none",
    fill = guide_legend(
      title.position = "top"
    )
  )

response_line <- ggplot(
  responses,
  aes(radio, sales, color = tv)
) +
  geom_line(
    size = 2
  ) +
  scale_color_manual(
    name = "Investment in TV Advertising\n(thousands of dollars)",
    values = qaad_colors("dark_purple", "celadon_blue", "flame_orange")
  ) +
  labs(
    x = "Investment in Radio Advertising\n(thousands of dollars)",
    y = "Sales\n(thousands of units)"
  ) +
  coord_cartesian(
    expand = FALSE
  ) +
  theme(
    legend.position = "top",
    legend.direction = "horizontal",
    legend.title.align = 0.5
  ) +
  guides(
    color = guide_legend(
      title.position = "top"
    )
  )

```

```{r, fig.width = 12.75, fig.asp = 0.56}

response_tile + 
  response_line & 
  theme(
    legend.margin = margin(0, 0, 0, 0.1, "in"),
    plot.margin = margin(0, 0.2, 0.2, 0.2, "in")
  )

```



---

## &#x1F52D; Looking Ahead

```{r}

schedule <- here("_misc", "course_outline.xlsx") %>% 
  readxl::read_excel(1) %>% 
  mutate(
    date = lubridate::ymd(date),
    description = glue::glue(
    "<b>{topic}</b> <span style = 'font-size: 80%; color: #9C9C9C;'>({date})</span><br>
    <div style = 'font-size: 80%; font-style: italic; color: #9C9C9C;'>{details}</div>"
  )) %>% 
  select(description)

```

.pull-left[

```{r}

schedule %>% 
  slice(7:10) %>% 
  kbl(escape = FALSE, col.names = NULL) %>% 
  kable_paper(c("striped", "hover"))

```

]