---
title: "Lecture 07: Linear Models 4"  
date: 'Last updated: `r Sys.Date()`'
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [xaringan-themer.css, custom_style.css]
    seal: true
    nature:
      highlightStyle: magula
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
---

```{r, include = FALSE, code=xfun::read_utf8(here::here("slides", "before_chunk.R"))}
```

```{r}

set.seed(42)

n <- 100

income <- tibble(
  sex = c(rep("female", n), rep("male", n)),
  education = rpois(n, 5) + ifelse(sex == "female", 10, 2),
  error = rnorm(2*n),
  income = ifelse(
    sex == "female",
    1 + 0.5 * education + error,
    14 + 0.8 * education + error
  )
)

kidiq <- read_csv(here("datasets", "kidiq.csv"))

mtcars <- mtcars %>% 
  select(mpg, hp, wt) %>% 
  as_tibble()

```

## &#x1F4CB; Lecture Outline

1. Qualitative variables
    - Dummy variables
    - Intercept offset
    - t-test
2. Centering 
2. Scaling
2. Interactions



---

## Qualitative Variables

.pull-left[

```{r}

limits <- 
  with(
  income,
  tibble(
    x = c(min(education)-1, max(education)+1),
    y = c(min(income)-1, max(income)+1)
  )
)

gg_income <- 
  ggplot(income) +
  scale_color_manual(
    name = NULL,
    values = qaad_colors("celadon_blue", "flame_orange")
  ) +
  labs(
    x = "Education (Years)", 
    y = "Income ($1000)"
  ) +
  theme(
    legend.position = "none"
  ) +
  coord_cartesian(
    x = limits[["x"]],
    y = limits[["y"]],
    expand = FALSE
  )

gg_income + 
  geom_point(
    aes(education, income, color = sex),
    size = 3,
    alpha = 0.6
  ) +
  geom_text(
    data = tibble(x = c(12, 10), y = c(21, 8), sex = c("male", "female")),
    aes(x, y, label = sex, color = sex),
    hjust = c(0, 1),
    size = 7
  )

```

]

---
count: false

## Qualitative Variables

.pull-left[

```{r}

lm_null <- lm(income ~ education, data = income)

new_data <- tibble(
  education = seq(0, 30, length = n*2)
)

estimate_null <- predict(lm_null, newdata = new_data, se.fit = TRUE)

estimate_null <- 
  with(
    estimate_null,
    tibble(
      y    = fit,
      ymin = fit - 2*se.fit,
      ymax = fit + 2*se.fit
    )
  ) %>% 
  bind_cols(new_data)

gg_income + 
  geom_ribbon(
    data = estimate_null,
    aes(education, ymin = ymin, ymax = ymax),
    fill = "gray95",
    color = "transparent"
  ) +
  geom_line(
    data = estimate_null,
    aes(education, y)
  ) + 
  geom_point(
    aes(education, income, color = sex),
    size = 3,
    alpha = 0.6
  ) +
  geom_text(
    data = tibble(x = c(12, 10), y = c(21, 8), sex = c("male", "female")),
    aes(x, y, label = sex, color = sex),
    hjust = c(0, 1),
    size = 7
  )

```

]

.pull-right[

 

```{r}

models <- broom::tidy(lm_null) %>% rename("t.statistic" = statistic)

models %>% 
  kbl(table.attr = "class='table-model table-fullwidth'") %>% 
  kable_paper(c("striped", "hover")) %>% 
  pack_rows("Sex Insensitive Model", 1, 2)

```
  
]


---
name: lmsex
count: false

## Qualitative Variables

.pull-left[

```{r}

lm_sex <- lm(income ~ education + sex, data = income)

new_data <- tibble(
  education = rep(seq(0, 30, length = n*2), 2),
  sex = c(rep("male", n*2), rep("female", n*2))
)

estimate_sex <- 
  with(
    predict(lm_sex, newdata = new_data, se.fit = TRUE),
    tibble(
      y    = fit,
      ymin = fit - 2*se.fit,
      ymax = fit + 2*se.fit
    )
  ) %>% 
  bind_cols(new_data)

gg_income + 
  geom_ribbon(
    data = estimate_sex,
    aes(education, ymin = ymin, ymax = ymax, group = sex),
    fill = "gray95",
    color = "transparent"
  ) +
  geom_line(
    data = estimate_sex,
    aes(education, y, group = sex)
  ) + 
  geom_point(
    aes(education, income, color = sex),
    size = 3,
    alpha = 0.6
  ) +
  geom_text(
    data = tibble(x = c(12, 10), y = c(21, 8), sex = c("male", "female")),
    aes(x, y, label = sex, color = sex),
    hjust = c(0, 1),
    size = 7
  )

```

]

.pull-right[

```{r}

models <- 
  broom::tidy(lm_sex) %>% 
  rename("t.statistic" = statistic) %>% 
  mutate(term = str_replace(term, "sexmale", "sex(male)")) %>% 
  bind_rows(models, .)

models %>% 
  kbl(table.attr = "class='table-model table-fullwidth'") %>% 
  kable_paper(c("striped", "hover")) %>% 
  pack_rows("Sex Insensitive Model", 1, 2) %>% 
  pack_rows("Sex Sensitive Model",   3, 5)

```

]


---

## Dummy Variables

.pull-left[

__Dummy variables__ code each observation:

$x_i=\left\{\begin{array}{c l} 1 & \text{if i-th observation in category} \\ 0 & \text{if i-th observation not in category} \end{array}\right.$

with zero being the __reference class__.

For _m_ categories, requires m-1 dummy variables.

]

.pull-right[

```{r}

tibble(
  Category = c("D<sub>1</sub>", "D<sub>2</sub>", "\\(\\vdots\\)", "D<sub>m-1</sub>", "D<sub>m</sub>"),
  "D<sub>1</sub>" = c(1, 0, "\\(\\vdots\\)", 0, 0),
  "D<sub>2</sub>" = c(0, 1, "\\(\\vdots\\)", 0, 0),
  "\\(\\cdots\\)" = c(0, 0, "\\(\\vdots\\)", 0, 0),
  "D<sub>m-1</sub>" = c(0, 0, "\\(\\vdots\\)", 1, 0)
) %>% 
  kbl(
    escape = FALSE,
    table.attr = "class='table-simple table-left'"
  ) %>% 
  kable_paper(
    bootstrap_options = "condensed",
    full_width = FALSE
  )

```

]


---

## Dummy Variables

.pull-left[

__Dummy variables__ code each observation:

$sex_i=\left\{\begin{array}{c l} 1 & \text{if i-th person is male} \\ 0 & \text{if i-th person is not male} \end{array}\right.$

Here the __reference class__ is _female_.

The sex variable has two categories, hence one dummy variable.

]

.pull-right[

```{r}

tibble(
  sex = c("male", "female", "\\(\\vdots\\)", "male", "female"),
  "male" = c(1, 0, "\\(\\vdots\\)", 1, 0)
) %>% 
  kbl(
    escape = FALSE,
    table.attr = "class='table-simple table-left'"
  ) %>% 
  kable_paper(
    bootstrap_options = "condensed",
    full_width = FALSE
  )

```

]


---

## Intercept Offset

.pull-left[

Simple linear model  

$y_i = \beta_0 + \beta_1x_1 + \epsilon_i$  


Adding a binary category $D$ as an offset  

$y_i = (\beta_0 + \gamma D) + \beta_1x_1 + \epsilon_i$  

with coefficient $\gamma$.  

]

.pull-right[

Model for females  

$y_i = (\beta_0 + \gamma\cdot 0) + \beta_1x_1$  
$y_i = \beta_0 + \beta_1x_1 + \epsilon_i$  

Model for males  

$y_i = (\beta_0 + \gamma\cdot 1) + \beta_1x_1$   
$y_i = (\beta_0 + \gamma) + \beta_1x_1 + \epsilon_i$  

]



---
template: lmsex


---

## t-test

.pull-left[

```{r}

sex_mean <- income %>% 
  group_by(sex) %>% 
  summarize(intercept = mean(income)) %>% 
  mutate(
    x = c(0.5, 0.5),
    xend = c(1, 2)
  )

ggplot(income) +
  scale_color_manual(
    name = NULL,
    values = qaad_colors("celadon_blue", "flame_orange")
  ) +
  labs(
    x = "Sex", 
    y = "Income ($1000)"
  ) +
  theme(
    legend.position = "none"
  ) +
  geom_point(
    aes(sex, income, color = sex),
    size = 3,
    alpha = 0.6
  ) +
  geom_segment(
    data = sex_mean,
    aes(x, intercept, xend = xend, yend = intercept)
  ) +
  coord_cartesian(
    xlim = c(0.5, 2.5),
    ylim = with(income, c(min(income)-1, max(income)+1)),
    expand = FALSE
  )

```

]

.pull-right[

```{r}

lm_ttest <- lm(income ~ sex, data = income)

broom::tidy(lm_ttest) %>% 
  rename("t.statistic" = statistic) %>% 
  kbl(table.attr = "class='table-model table-fullwidth'") %>% 
  kable_paper(c("striped", "hover")) %>% 
  pack_rows("Sex Sensitive Intercept Model", 1, 2)

```

mean(female) = `r sex_mean$intercept[[1]]`  
mean(male) = mean(female) + 11.16 = `r sex_mean$intercept[[2]]`  

]


---

## Centering

.pull-left[

```{r}

lm_kid <- lm(kid ~ mom, data = kidiq)

ggplot() + 
  geom_point(
    data = kidiq, 
    aes(mom, kid),
    size = 3
  ) +
  geom_abline(
    intercept = coefficients(lm_kid)[["(Intercept)"]],
    slope = coefficients(lm_kid)[["mom"]]
  ) +
  labs(
    x = "Mom's IQ",
    y = "Kid's IQ"
  ) +
  coord_cartesian(
    xlim = with(kidiq, c(min(mom)-5, max(mom)+5)),
    ylim = with(kidiq, c(min(kid)-5, max(kid)+5)),
    expand = FALSE
  )

```

]

.pull-right[

```{r}

models <- broom::tidy(lm_kid) %>% 
  rename("t.statistic" = statistic)

models %>% 
  kbl(table.attr = "class='table-model table-fullwidth'") %>% 
  kable_paper(c("striped", "hover")) %>% 
  pack_rows("Uncentered Model", 1, 2)

```

Model child IQ as a function of mother's IQ<sup>1</sup> 

__Question__ What is the intercept?

.footnote[<sup>1</sup>_Whether or not IQ is meaningful, this example is illustrative!_]

]


---
count: false

## Centering

.pull-left[

```{r}

ggplot() + 
  geom_point(
    data = kidiq, 
    aes(mom, kid),
    size = 3
  ) +
  geom_abline(
    intercept = coefficients(lm_kid)[["(Intercept)"]],
    slope = coefficients(lm_kid)[["mom"]]
  ) +
  labs(
    x = "Mom's IQ",
    y = "Kid's IQ"
  ) +
  coord_cartesian(
    xlim = with(kidiq, c(-5, max(mom)+5)),
    ylim = with(kidiq, c(min(kid)-5, max(kid)+5)),
    expand = FALSE
  ) +
  annotate(
    "segment",
    x = 0, xend = 0,
    y = 60, yend = coefficients(lm_kid)[["(Intercept)"]] + 2,
    color = qaad_colors("rufous_red"),
    arrow = arrow(length = unit(0.15, "inches"), type = "closed")
  ) +
  annotate(
    "text",
    x = 0,
    y = 62,
    label = "What is this?!!!",
    color = qaad_colors("rufous_red"),
    size = 7,
    hjust = 0,
    vjust = 0
  )

```

]

.pull-right[

```{r}

models %>% 
  kbl(table.attr = "class='table-model table-fullwidth'") %>% 
  kable_paper(c("striped", "hover")) %>% 
  pack_rows("Uncentered Model", 1, 2)

```

Model child IQ as a function of mother's IQ<sup>1</sup> 

.footnote[<sup>1</sup>_Whether or not IQ is meaningful, this example is illustrative!_]

]


---

## Centering

.pull-left[

```{r}

ggplot() + 
  geom_point(
    data = kidiq, 
    aes(mom, kid),
    size = 3
  ) +
  geom_abline(
    intercept = coefficients(lm_kid)[["(Intercept)"]],
    slope = coefficients(lm_kid)[["mom"]]
  ) +
  labs(
    x = "Mom's IQ",
    y = "Kid's IQ"
  ) +
  coord_cartesian(
    xlim = with(kidiq, c(-5, max(mom)+5)),
    ylim = with(kidiq, c(min(kid)-5, max(kid)+5)),
    expand = FALSE
  ) +
  annotate(
    "segment",
    x = 0, xend = 0,
    y = 60, yend = coefficients(lm_kid)[["(Intercept)"]] + 2,
    color = qaad_colors("rufous_red"),
    arrow = arrow(length = unit(0.15, "inches"), type = "closed")
  ) +
  annotate(
    "text",
    x = 0,
    y = 62,
    label = "What is this?!!!",
    color = qaad_colors("rufous_red"),
    size = 7,
    hjust = 0,
    vjust = 0
  )

```

]

.pull-right[

Centering will help make this intercept more interpretable.

To center, simply subtract a suitable measure of central tendency from the covariate:

$$\text{Mom's IQ} - \text{mean(Mom's IQ)}$$

That gets us this model...

]



---

## Centering

.pull-left[

```{r}

kidiq_scale <- kidiq %>% mutate(mom = scale(mom, scale = FALSE)[,1])

lm_kid_scale <- lm(kid ~ mom, data = kidiq_scale)

ggplot() + 
  geom_vline(
    xintercept = 0,
    color = qaad_colors("rufous_red"),
    linetype = "dashed"
  ) +
  geom_point(
    data = kidiq, 
    aes(mom - mean(mom), kid),
    size = 3
  ) +
  geom_abline(
    intercept = coefficients(lm_kid_scale)[["(Intercept)"]],
    slope = coefficients(lm_kid_scale)[["mom"]]
  ) +
  labs(
    x = "Mom's IQ",
    y = "Kid's IQ"
  ) +
  coord_cartesian(
    xlim = with(kidiq_scale, c(min(mom)-5, max(mom)+5)),
    ylim = with(kidiq_scale, c(min(kid)-5, max(kid)+5)),
    expand = FALSE
  )

```

]

.pull-right[

```{r}

broom::tidy(lm_kid_scale) %>% 
  rename("t.statistic" = statistic) %>% 
  bind_rows(models, .) %>% 
  kbl(table.attr = "class='table-model table-fullwidth'") %>% 
  kable_paper(c("striped", "hover")) %>% 
  pack_rows("Uncentered Model", 1, 2) %>% 
  pack_rows("Centered Model", 3, 4)

```

Now we interpret the intercept as expected IQ of a child for a mother with mean IQ. (Notice change in standard error!)

]



---

## Scaling

.pull-left[

```{r, fig.asp = 1}

lm_cars <- lm(mpg ~ hp + wt, data = mtcars)

mtcars_longer <- mtcars %>% 
  select(mpg, hp, wt) %>% 
  pivot_longer(
    -mpg,
    names_to = "variable",
    values_to = "x"
  )

new_data <- 
  tibble(
    hp = with(mtcars, seq(min(hp)-7, max(hp)+7, length = 300)),
    wt = with(mtcars, seq(min(wt)-0.13, max(wt)+0.13, length = 300))
  )

margins <-
  bind_rows(
    with(
      predict(
        lm_cars, 
        newdata = new_data %>% mutate(wt = mean(mtcars$wt)), 
        se.fit = TRUE
      ),
      tibble(
        variable = "hp",
        x = new_data$hp,
        y = fit,
        ymin = fit - 2*se.fit,
        ymax = fit + 2*se.fit
      )
    ),
    with(
      predict(
        lm_cars, 
        newdata = new_data %>% mutate(hp = mean(mtcars$hp)), 
        se.fit = TRUE
      ),
      tibble(
        variable = "wt",
        x = new_data$wt,
        y = fit,
        ymin = fit - 2*se.fit,
        ymax = fit + 2*se.fit
      )
    )
  )

gg_cars <- ggplot() + 
  geom_ribbon(
    data = margins,
    aes(x, ymin = ymin, ymax = ymax),
    fill = "gray95",
    color = "transparent"
  ) +
  geom_line(
    data = margins,
    aes(x,y)
  ) +
  geom_point(
    data = mtcars_longer, 
    aes(x, mpg),
    size = 3
  ) +
  labs(
    x = NULL,
    y = "Fuel Efficiency\n(Miles Per Gallon)"
  ) +
  coord_cartesian(
    ylim = with(mtcars, c(min(mpg)-2, max(mpg)+2)),
    expand = FALSE
  ) +
  facet_wrap(
    ~variable,
    nrow = 2,
    scale = "free_x",
    strip.position = "bottom",
    labeller = labeller(variable = c(wt = "Weight (1,000 pounds)", hp = "Horse Power"))
  ) +
  theme(
    strip.background = element_blank(),
    strip.placement = "outside"
  )

gg_cars

```

]

.pull-right[

```{r}

models <- broom::tidy(lm_cars) %>% 
  rename("t.statistic" = statistic)

models %>% 
  kbl(table.attr = "class='table-model table-fullwidth'") %>% 
  kable_paper(c("striped", "hover")) %>% 
  pack_rows("Unscaled Model", 1, 3)

```

Model fuel efficiency as a function of weight and horse power.  

__Question__ Which covariate has a bigger effect?  

]


---

## Scaling

.pull-left[

```{r, fig.asp = 1}

gg_cars

```

]

.pull-right[

__Problem__ We cannot directly compare these coefficients, as they use different units? What, for example, is the equivalent in horsepower of an increase of 500 pounds?  

__Solution__ We standardize the variables by centering them on their means and dividing by their standard deviations, i.e., we convert them to z-scores. 

$$z_i = \frac{x_i-\bar{x}}{\sigma_{x}}$$

All coefficients will now give the change in y for a 1 $\sigma$ change in x.  

]


---

## Scaling

.pull-left[

```{r, fig.asp = 1}

rescale <- function(x) scale(x)[,1]

mtcars_scaled <- mtcars %>% mutate(across(c(hp, wt), rescale))

lm_cars <- lm(mpg ~ hp + wt, data = mtcars_scaled)

mtcars_longer <- mtcars_scaled %>% 
  pivot_longer(
    -mpg,
    names_to = "variable",
    values_to = "x"
  )

new_data <- 
  tibble(
    hp = with(mtcars_longer, seq(min(x), max(x)+0.1, length = 300)),
    wt = hp
  )

margins <-
  bind_rows(
    with(
      predict(
        lm_cars, 
        newdata = new_data %>% mutate(wt = 0), 
        se.fit = TRUE
      ),
      tibble(
        variable = "hp",
        x = new_data$hp,
        y = fit,
        ymin = fit - 2*se.fit,
        ymax = fit + 2*se.fit
      )
    ),
    with(
      predict(
        lm_cars, 
        newdata = new_data %>% mutate(hp = 0), 
        se.fit = TRUE
      ),
      tibble(
        variable = "wt",
        x = new_data$wt,
        y = fit,
        ymin = fit - 2*se.fit,
        ymax = fit + 2*se.fit
      )
    )
  )

ggplot() + 
  geom_ribbon(
    data = margins,
    aes(x, ymin = ymin, ymax = ymax),
    fill = "gray95",
    color = "transparent"
  ) +
  geom_line(
    data = margins,
    aes(x,y)
  ) +
  geom_point(
    data = mtcars_longer, 
    aes(x, mpg),
    size = 3
  ) +
  labs(
    x = NULL,
    y = "Fuel Efficiency\n(Miles Per Gallon)"
  ) +
  coord_cartesian(
    ylim = with(margins, c(min(y)-2, max(y)+2)),
    expand = FALSE
  ) +
  facet_wrap(
    ~variable, 
    nrow = 2,
    scale = "free_x",
    strip.position = "bottom",
    labeller = labeller(variable = c(wt = "Weight (z)", hp = "Horsepower (z)"))
  ) +
  theme(
    strip.background = element_blank(),
    strip.placement = "outside"
  )

```

]

.pull-right[

```{r}

models <- broom::tidy(lm_cars) %>% 
  rename("t.statistic" = statistic) %>% 
  bind_rows(models, .)

models %>% 
  kbl(table.attr = "class='table-model table-fullwidth'") %>% 
  kable_paper(c("striped", "hover")) %>% 
  pack_rows("Unscaled Model", 1, 3) %>% 
  pack_rows("Scaled Model", 4, 6)

```


]


---

## Interactions

.pull-left[

```{r}

kidiq_scale <- kidiq_scale %>% mutate(mom_hs = factor(mom_hs))

gg_kid_hs <- ggplot() + 
  geom_point(
    data = kidiq_scale, 
    aes(mom, kid, color = mom_hs),
    size = 3
  ) +
  scale_color_manual(
    name = NULL,
    values = qaad_colors("maize_crayola", "dark_purple"),
    labels = c("No High School", "High School")
  ) +
  labs(
    x = "Mom's IQ",
    y = "Kid's IQ"
  ) +
  coord_cartesian(
    xlim = with(kidiq_scale, c(min(mom)-5, max(mom)+5)),
    ylim = with(kidiq_scale, c(min(kid)-5, max(kid)+5)),
    expand = FALSE
  ) +
  theme(
    legend.position = c(0.8, 0.14),
    legend.box.just = c("right", "bottom")
  )

gg_kid_hs

```

]

.pull-right[

```{r}

models <- broom::tidy(lm_kid_scale) %>% 
  rename("t.statistic" = statistic)

models %>% 
  kbl(table.attr = "class='table-model table-fullwidth'") %>% 
  kable_paper(c("striped", "hover")) %>% 
  pack_rows("Additive Model", 1, 2)

```

__Question__ What if the relationship between child's IQ and mother's IQ depends on the mother's educational attainment?

]


---

## Interactions

.pull-left[

```{r}

gg_kid_hs

```

]

.pull-right[

Simple formula becomes:

$$y_i = (\beta_0 + \gamma D) + (\beta_1x_1 + \omega(x_1 \cdot D)) + \epsilon$$

With  
- $\gamma$ giving the change in $\beta_0$ and  
- $\omega$ giving the change in $\beta_1$.  

That's a __change in intercept__ and a __change in slope__!

]

---

## Interactions

.pull-left[

```{r}

lm_kidx <- lm(kid ~ mom * mom_hs, data = kidiq_scale)

new_data <- 
  tibble(
    mom = with(kidiq_scale, rep(seq(min(mom)-5, max(mom)+5, length = 300), 2)),
    mom_hs = factor(rep(c(0, 1), each = 300))
  )

margins <-
  with(
      predict(
        lm_kidx, 
        newdata = new_data, 
        se.fit = TRUE
      ),
      tibble(
        mom_hs = new_data$mom_hs,
        x = new_data$mom,
        y = fit,
        ymin = fit - 2*se.fit,
        ymax = fit + 2*se.fit
      )
    )

ggplot() + 
  geom_ribbon(
    data = margins,
    aes(x, ymin = ymin, ymax = ymax, group = mom_hs),
    fill = "gray95",
    color = "transparent"
  ) +
  geom_line(
    data = margins,
    aes(x, y, color = mom_hs),
    size = 1
  ) + 
  geom_point(
    data = kidiq_scale, 
    aes(mom, kid, color = mom_hs),
    size = 3
  ) +
  scale_color_manual(
    name = NULL,
    values = qaad_colors("maize_crayola", "dark_purple"),
    labels = c("No High School", "High School")
  ) +
  labs(
    x = "Mom's IQ",
    y = "Kid's IQ"
  ) +
  coord_cartesian(
    xlim = with(kidiq_scale, c(min(mom)-5, max(mom)+5)),
    ylim = with(kidiq_scale, c(min(kid)-5, max(kid)+5)),
    expand = FALSE
  ) +
  theme(
    legend.position = c(0.8, 0.14),
    legend.box.just = c("right", "bottom")
  )

```

]

.pull-right[

```{r}

models <- broom::tidy(lm_kidx) %>% 
  rename("t.statistic" = statistic) %>% 
  bind_rows(models, .) %>% 
  mutate(term = str_replace(term, "mom_hs1", "high_school"))

models %>% 
  kbl(table.attr = "class='table-model table-fullwidth'") %>% 
  kable_paper(c("striped", "hover")) %>% 
  pack_rows("Additive Model", 1, 2) %>% 
  pack_rows("Interaction Model", 3, 6)

```


]



---

## &#x1F52D; Looking Ahead

```{r}

schedule <- here("_misc", "course_outline.xlsx") %>% 
  readxl::read_excel(1) %>% 
  mutate(
    date = lubridate::ymd(date),
    description = glue::glue(
    "<b>{topic}</b> <span style = 'font-size: 80%; color: #9C9C9C;'>({date})</span><br>
    <div style = 'font-size: 80%; font-style: italic; color: #9C9C9C;'>{details}</div>"
  )) %>% 
  select(description)

```

.pull-left[

```{r}

schedule %>% 
  slice(8:11) %>% 
  kbl(escape = FALSE, col.names = NULL) %>% 
  kable_paper(c("striped", "hover"))

```

]