---
title: "Lecture 08: Multiple Linear Models"  
date: 'Last updated: `r Sys.Date()`'
---

```{r setup}
#| include: false

here::here("slides", "_setup.R") |> source()

library(broom)
library(car)

adverts <- read_csv("https://www.statlearning.com/s/Advertising.csv") |> 
  rename_with(tolower) |> 
  select(sales, tv, radio, newspaper)

adverts_lm <- lm(sales ~ tv, data = adverts)

estimate <- augment(adverts_lm)

predictors <- tidy(adverts_lm)

beta0 <- predictors |> filter(term == "(Intercept)") |> pull(estimate)
beta1 <- predictors |> filter(term == "tv") |> pull(estimate)

```

## 📋 Lecture Outline

- Simple Linear Regression
- Multiple Linear Regression
- Collinearity
- Multicollinearity
- ANOVA
- Partial Dependence

## Simple Linear Regression

:::::: {.r-stack}
::::: {.fragment .fade-out fragment-index=0 .w100}
:::: {.columns}
::: {.column}

```{r}
#| fig-asp: 0.9

ggplot(estimate) +
  geom_point(
    aes(tv, sales),
    size = 2.5
  ) +
  labs(
    x = "Investment in TV Advertising\n(thousands of dollars)",
    y = "Sales\n(thousands of units)"
  ) +
  theme(axis.title = element_text(lineheight = 1.4))

```

::: {style="text-align:center"}
_From [James et al (2021). ISLR. 2nd Ed.](https://www.statlearning.com/)_
:::

:::
::: {.column}

```{r}
#| results: asis

adverts |> 
  gt() |> 
  tab_options(
    column_labels.font.weight = "bold",
    container.width = pct(100),
    container.height = px(350),
    container.overflow.y = TRUE,
    row.striping.background_color = "#ededed",
    row.striping.include_table_body = TRUE,
    table.border.top.style = "hidden",
    table.font.size = px(20),
    table.width = pct(100)
  ) |> 
  as_raw_html()

```

:::
::::
:::::
::::: {.fragment .fade-in fragment-index=0 .w100}
:::: {.columns}
::: {.column}

```{r}
#| fig-asp: 0.9

ggplot(estimate) +
  geom_segment(
    aes(tv, sales, xend = tv, yend = .fitted),
    color = "gray60"
  ) +
  geom_abline(
    intercept = beta0,
    slope = beta1,
    color = qcolors("kobe"),
    size = 2
  ) +
  geom_point(
    aes(tv, sales),
    size = 3
  ) +
  labs(
    x = "Investment in TV Advertising\n(thousands of dollars)",
    y = "Sales\n(thousands of units)"
  ) +
  theme(axis.title = element_text(lineheight = 1.4))

```

::: {style="text-align:center"}
_From [James et al (2021). ISLR. 2nd Ed.](https://www.statlearning.com/)_
:::

:::
::: {.column}

```{r}
#| results: asis

predictors |>
  rename("t.statistic" = statistic) |> 
  mutate(across(where(is.numeric), ~round(.x, 3))) |> 
  gt() |> 
  tab_options(
    column_labels.font.weight = "bold",
    container.width = pct(100),
    row.striping.background_color = "#ededed",
    row.striping.include_table_body = TRUE,
    table.border.top.style = "hidden",
    table.font.size = px(20),
    table.width = pct(100)
  ) |> 
  as_raw_html()

```

__Note the units!__ We interpret the slope coefficient here as saying, "For every $1000 spent on television advertising, sales increased by approximately 50 units."

__Question__ What does the intercept mean?

:::
::::
:::::
::::::

## But, muh data...

__Question__ What about the money invested in radio and newspaper advertising? 

. . .

<br>

To answer this, we could try building three separate models:  

1. sales $\sim$ tv
2. sales $\sim$ newspaper
3. sales $\sim$ radio

But, this assumes that the coefficient estimates are independent of each other... 

. . .

<br>

__Solution__ Make one big model!  

1. sales $\sim$ tv + newspaper + radio

## Multiple Linear Regression

```{r}

adverts_lm_list <- list(
  tv = adverts_lm,
  radio = lm(sales ~ radio, data = adverts),
  newspaper = lm(sales ~ newspaper, data = adverts),
  all = lm(sales ~ tv + radio + newspaper, data = adverts)
)

adverts_lm_table <- adverts_lm_list |> 
  lapply(tidy) |> 
  bind_rows() |> 
  mutate(model = c(rep("simple", 6), rep("multiple", 4))) |> 
  rename("t.statistic" = statistic) |> 
  select(model, everything())

```

:::::: {.r-stack}
::::: {.fragment .fade-out fragment-index=0 .w100 .mt-0}
:::: {.columns}
::: {.column}

```{r}
#| fig-asp: 0.75

adverts_lm_table |> 
  filter(term != "(Intercept)") |> 
  ggplot(aes(estimate, term, color = model)) +
  geom_vline(
    xintercept = 0,
    color = qcolors("kobe"),
    size = 0.4
  ) +
  geom_errorbarh(
    aes(xmin = estimate - std.error, xmax = estimate + std.error),
    height = 0.25,
    size = 1
  ) +
  geom_point(
    size = 4
  ) +
  scale_color_manual(
    values = c("transparent", qcolors("gold"))
  ) +
  labs(
    x = "Coefficient Estimates"
  ) +
  theme(
    axis.title.y = element_blank(),
    legend.position = c(0.85, 0.87),
    legend.box.just = c("right", "top"),
    legend.title = element_blank()
  )

```

::: {style="text-align:center"}
_From [James et al (2021). ISLR. 2nd Ed.](https://www.statlearning.com/)_
:::

:::
::: {.column}

Coefficients are estimated in the same way, but the __interpretation__ changes! Consider 

$$\beta_{radio}=0.189$$ 

Read: "For some amount of `tv` and `newspaper` advertising, spending an additional $1000 on `radio` advertising increases sales by approximately 189 units."

:::
::::
:::::
::::: {.fragment .fade-in fragment-index=0 .w100 .mt-0}
:::: {.columns}
::: {.column}

```{r}
#| fig-asp: 0.75

adverts_lm_table |> 
  filter(term != "(Intercept)") |> 
  ggplot(aes(estimate, term, color = model)) +
  geom_vline(
    xintercept = 0,
    color = qcolors("kobe"),
    size = 0.4
  ) +
  geom_errorbarh(
    aes(xmin = estimate - std.error, xmax = estimate + std.error),
    height = 0.25,
    size = 1
  ) +
  geom_point(
    size = 4
  ) +
  scale_color_manual(
    values = c(qcolors("eerie", "gold"))
  ) +
  labs(
    x = "Coefficient Estimates"
  ) +
  theme(
    axis.title.y = element_blank(),
    legend.position = c(0.85, 0.87),
    legend.box.just = c("right", "top"),
    legend.title = element_blank()
  )

```

::: {style="text-align:center"}
_From [James et al (2021). ISLR. 2nd Ed.](https://www.statlearning.com/)_
:::

:::
::: {.column}

The estimate for `newspaper` is no longer significantly different than zero!

__Question__ Why might this be?

:::
::::
:::::
::::::

## Collinearity

:::::: {.columns}
::::: {.column width="45%"}

```{r}
#| fig-align: left 
#| fig-asp: 1

pnt_matrix <- lapply(
  seq_along(adverts),
  function(x){
    
    n.var <- ncol(adverts)
    n.obs <- nrow(adverts)
    
    s.var <- names(adverts)[[x]]
    
    var1 <- adverts[x]
    
    names(var1) <- "x"
    
    var1 <- var1 |> 
      slice(rep(1:n.obs, n.var)) |> 
      mutate(var1 = s.var)
    
    var2 <- adverts |> 
      pivot_longer(
        everything(),
        names_to = "var2",
        values_to = "y"
      ) |> 
      arrange(var2)
    
    bind_cols(var1, var2) |> 
      select(var1, var2, x, y)
    
  }
) |> 
  bind_rows() |> 
  mutate(
    across(c(var1, var2), factor),
    across(c(var1, var2), ~forcats::fct_relevel(.x, "sales", "tv", "radio")),
  )

lbl_matrix <- pnt_matrix |> 
  group_by(var1, var2) |> 
  summarize(
    x = 0.5 * (min(x) + max(x)),
    y = 0.5 * (min(y) + max(y))
  ) |> 
  filter(
    var2 == var1
  )

cor_matrix <- cor(adverts) |> 
  round(digits = 2) |> 
  as_tibble(rownames = "var1") |> 
  tidyr::pivot_longer(
    -var1, 
    names_to = "var2", 
    values_to = "value"
  ) |> 
  mutate(
    across(c(var1, var2), factor),
    across(c(var1, var2), ~forcats::fct_relevel(.x, "sales", "tv", "radio")),
    color = ifelse(value > 0.5 | is.na(value), "black", "white")
  )

bob_matrix <- pnt_matrix |> 
  group_by(var1, var2) |> 
  summarize(
    xmid = 0.5 * (min(x) + max(x)),
    ymid = 0.5 * (min(y) + max(y)),
    xmin = min(x),
    xmax = max(x),
    ymin = min(y),
    ymax = max(y)
  ) |> 
  filter(
    !(var2 == var1),
    !(var2 == "sales"),
    !(var2 == "tv" & var1 %in% c("radio", "newspaper")),
    !(var2 == "radio" & var1 == "newspaper")
  )

cor_matrix <- left_join(bob_matrix, cor_matrix, by = c("var1", "var2"))

pnt_matrix <- pnt_matrix |> 
  filter(
    !(var2 == var1),
    !(var2 == "newspaper"),
    !(var2 == "radio" & var1 %in% c("sales", "tv")),
    !(var2 == "tv" & var1 == "sales")
  )

ggplot() +
  geom_point(
    data = pnt_matrix, 
    aes(x, y),
    size = 1.8,
    shape = 21,
    color = qcolors("kobe"),
    fill = alpha(qcolors("bronze"), 0.75)
  ) +
  geom_smooth(
    data = pnt_matrix, 
    aes(x, y),
    method = "lm",
    se = FALSE,
    color = qcolors("sapphire")
  ) +
  geom_text(
    data = lbl_matrix,
    aes(x, y, label = var1),
    size = 11/.pt
  ) +
  geom_rect(
    data = cor_matrix,
    aes(
      fill = value,
      xmin = xmin,
      xmax = xmax,
      ymin = ymin,
      ymax = ymax
    )
  ) +
  scale_fill_viridis(
    option = "magma",
    na.value = "white"
  ) +
  geom_text(
    data = cor_matrix,
    aes(
      xmid,
      ymid,
      label = value, 
      color = color
    ),
    hjust = 0.5,
    vjust = 0.5,
    size = 11/.pt
  ) +
  scale_color_manual(
    values = c("black", "white")
  ) +
  facet_grid(var2 ~ var1, scales = "free") +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    legend.position = "none",
    strip.background = element_blank(),
    strip.text = element_blank()
  ) +
  coord_cartesian(expand = FALSE)

```

:::::
::::: {.column width="55%"}

__Answer:__ `newspaper` and `radio` advertising are correlated ($\rho$ = 0.35).  

So, `newspaper` advertising was never really increasing `sales`. It was just taking "credit" for `radio`'s contribution.

:::::
::::::

## Multicollinearity

:::::: {.r-stack}
::::: {.fragment .fade-out fragment-index=0 .w100 .mt-0 .ml-0}

__Problem__ It is possible for multiple variables to correlate even if no single pair does. 

<br>

__Solution__ Measure the Variance Inflation Factor for each covariate.

$$VIF(\hat{B_j}) = \frac{1}{1-R^{2}_{j}}$$

where $R^{2}_{j}$ is the coefficient of determination for the model $x_j \sim x_1 + x_2 + \ldots + x_i$.

<br>

_Variance Inflation_ inflates the standard errors, too!

$$se(\hat{B_j}) = \sqrt{Var(\hat{B_j}) \cdot VIF(\hat{B_j})}$$

:::::
::::: {.fragment .fade-in-then-out fragment-index=0 .w100 .mt-0 .ml-0}

```{r}
#| results: asis

vif_vct <- vif(adverts_lm_list$all)

tibble(
  covariate = names(vif_vct),
  VIF = unname(vif_vct)
) |> 
  gt() |> 
  tab_options(
    column_labels.font.weight = "bold",
    container.width = pct(100),
    container.overflow.y = TRUE,
    row.striping.background_color = "#ededed",
    row.striping.include_table_body = TRUE,
    table.align = "left",
    table.border.top.style = "hidden",
    table.font.size = px(20),
    table.width = pct(40)
  ) |> 
  as_raw_html()

```

VIF close to 1 means no multicollinearity.  

__Rule of Thumb__ VIF that exceeds 5 is too much! Or is it 10? 🤔

:::::
::::: {.fragment .fade-in fragment-index=1 .w100 .mt-0 .ml-0}

Strategies for handling multicollinearity include:

- Removing one or more correlated variables (simplest strategy)
- Aggregating correlated variables (e.g., product or mean)
- Using an interaction term (more later)
- Using an ordination technique (more later)

:::::
::::::

## ANOVA for model

:::::: {.r-stack}
::::: {.fragment .fade-out fragment-index=0 .w100 .mt-0 .ml-0}

We have two __nested__ models of our data:  

- $S:$ sales $\sim$ tv
- $C:$ sales $\sim$ tv + newspaper + radio

__Question:__ Does the complex model (C) explain more variance than the simple model (S)? Does the difference arise by chance?  

Here are our  hypotheses:  

- $H_0:$ no difference in variance explained.
- $H_1:$ difference in variance explained.

:::::
::::: {.fragment .fade-in fragment-index=0 .w100 .mt-0 .ml-0}

```{r}
#| results: asis

adverts_anova <- with(adverts_lm_list, anova(tv, all))

adverts_anova <- adverts_anova |>
  tidy() |>
  rename(
    "RSS" = rss,
    "F" = statistic
  ) |>
  mutate(
    m = c("S", "C"),
    p.value = ifelse(
      p.value < 2.2e-16, 
      "< 2.2e-16", 
      as.character(signif(p.value, 3))
    )
  ) |>
  rename(
    "Pr(>F)" = p.value
  ) |>
  select(m, everything())

adverts_anova |>
  gt() |> 
  tab_options(
    column_labels.font.weight = "bold",
    container.width = pct(100),
    row.striping.background_color = "#ededed",
    row.striping.include_table_body = TRUE,
    table.align = "left",
    table.border.top.style = "hidden",
    table.font.size = px(20),
    table.width = pct(70)
  ) |> 
  sub_missing(missing_text = "") |> 
  as_raw_html()

```

::: {style="width:70%; margin-right:auto; margin-top:1em;"}
$$F = \frac{(RSS_{S} - RSS_{C})/(df_S - df_C)}{RSS_{C}/df_C}$$
:::

:::::
::::::

## Partial Dependence Plot

```{r}
#| layout-ncol: 2
#| out-width: 100%
#| fig-asp: 1

n <- 300

radio <- with(adverts, seq(min(radio), max(radio), length = n))
              
new_data <- expand.grid(
  tv = with(adverts, seq(min(tv), max(tv), length = n)),
  radio = radio
) |> 
  as_tibble() |> 
  mutate(
    newspaper = mean(adverts$newspaper)
  )

new_data$sales <- predict(adverts_lm_list$all, newdata = new_data)

margins <- with(adverts, quantile(tv, c(0.25, 0.5, 0.75)))
margins <- unname(margins)
margins <- round(margins, digits = 2)

responses <- lapply(
  margins, 
  function(z){
    
    nd <- tibble(
      radio = radio,
      tv = z,
      newspaper = mean(adverts$newspaper)
    )
    
    tibble(
      tv = z,
      radio = radio,
      sales = predict(adverts_lm_list$all, newdata = nd)
    )
    
  }) |> 
  bind_rows() |> 
  mutate(
    tv = factor(tv)
  )
  
ggplot() +
  geom_contour_filled(
    data = new_data, 
    aes(radio, tv, z = sales),
    color = "transparent",
    bins = 5
  ) +
  scale_fill_viridis(
    name = "Sales\n(thousands of units)",
    labels = paste0("< ", seq(6, 30, by = 6)),
    discrete = TRUE
  ) +
  geom_hline(
    aes(yintercept = margins),
    color = "white",
    linewidth = 3
  ) +
  geom_hline(
    aes(
      yintercept = margins,
      color = factor(margins)
    ),
    linewidth = 1.5
  ) +
  scale_color_manual(
    name = NULL,
    values = qcolors("eerie", "blue", "bronze")
  ) +
  labs(
    x = "Investment in Radio Advertising\n(thousands of dollars)",
    y = "Investment in TV Advertising\n(thousands of dollars)"
  ) +
  coord_cartesian(expand = FALSE) +
  theme(
    axis.title = element_text(lineheight = 1.4),
    legend.position = "top",
    legend.direction = "horizontal",
    legend.title = element_text(lineheight = 1.4),
    legend.title.align = 0.5,
    legend.margin = margin(0, 0, 0, 0.1, "in"),
    plot.margin = margin(0, 0.2, 0.2, 0.2, "in")
  ) +
  guides(
    color = "none",
    fill = guide_legend(
      title.position = "top"
    )
  )

ggplot(responses, aes(radio, sales, color = tv)) +
  geom_line(size = 2) +
  scale_color_manual(
    name = "Investment in TV Advertising\n(thousands of dollars)",
    values = qcolors("eerie", "blue", "bronze")
  ) +
  labs(
    x = "Investment in Radio Advertising\n(thousands of dollars)",
    y = "Sales\n(thousands of units)"
  ) +
  coord_cartesian(expand = FALSE) +
  theme(
    axis.title = element_text(lineheight = 1.4),
    legend.position = "top",
    legend.direction = "horizontal",
    legend.title = element_text(lineheight = 1.4),
    legend.title.align = 0.5,
    legend.margin = margin(0, 0, 0, 0.1, "in"),
    plot.margin = margin(0, 0.2, 0.2, 0.2, "in")
  ) +
  guides(color = guide_legend(title.position = "top"))

```
