---
title: "Quantitative Analysis of Archaeological Data"
subtitle: "Lecture 10: Generalized Linear Models 2"  
date: 'Last updated: `r Sys.Date()`'
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [xaringan-themer.css, custom_style.css]
    seal: TRUE
    nature:
      highlightStyle: magula
      highlightLines: TRUE
      countIncrementalSlides: TRUE
      ratio: '16:9'
---

```{r} 
#| include = FALSE, 
#| code = xfun::read_utf8(here::here("slides", "before_chunk.R"))
```

```{r}

snodgrass <- here("datasets", "snodgrass.csv") %>% 
  read_csv() %>% 
  select(inside, area)  

sites <- here("datasets", "site_counts.csv") %>% 
  read_csv()

library(archdata)

data("DartPoints")

darts <- DartPoints %>% 
  as_tibble() %>% 
  rename_with(tolower) %>% 
  select(name, length, width, thickness, weight) %>% 
  filter(weight < 20)

remove(DartPoints)

```

## Outline

- General Linear Model
- Generalized Linear Model
    - Distribution function
    - Linear predictor
    - Link function
- Gaussian for continuous data
- Binomial for binary data (and proportions)
- Poisson for count data
- Deviance
- Information Criteria
- ANOVA _aka_ Likelihood Ratio Test

---

## General Linear Models

$$y = E(Y) + \epsilon$$
$$E(Y) = \mu = \beta X$$

- $E(Y)$ is the expected value (equal to the conditional mean, $\mu$)  
- $\beta X$ is the linear predictor  
- The distibution of $\epsilon$ is  

$$\epsilon \sim N(0, \sigma^2)$$


---

## Generalized Linear Models

$$y = E(Y) + \epsilon$$
$$E(Y) = g(\mu) = \beta X$$

- $g$ is the link function, makes the relationship linear 
- $\beta X$ is the linear predictor
- The distibution of $\epsilon$ is

$$\epsilon \sim Exponential$$


---

## Generalized Linear Models

.panelset.sideways[

.panel[.panel-name[Exponential distribution]

- Function describing the expected distribution of the response
    - When centered on the mean, gives the distribution of the errors.
- A family of distributions: 
    - Normal - continuous and unbounded
    - Gamma - continuous and non-negative
    - Binomial - binary (yes/no) or proportional (0-1) 
    - Poisson - count data
- Two parameters: mean and variance
- Variance is a function of the mean:

$$Var(\epsilon) = \phi \mu$$

.ml4[
where $\phi$ is a scaling parameter, assumed to be equal to 1, meaning the variance is assumed to be equal to the mean.
]

]

.panel[.panel-name[Linear predictor]

- Incorporates information about the independent variables into the model
- Combination of $x$ variables and associated coefficients

$$\beta X = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \ldots + \beta_n x_n$$

]

.panel[.panel-name[Link function]

- $g()$ modifies the relationship between the predictors and the expected value.

$$E(Y) = g(u) = \beta X$$

- Makes this relationship linear

```{r}

tibble(
  Distribution = c(
    "Normal", "Gamma", "Poisson", "Binomial"
    ),
  Name = c(
    "Identity", "Inverse", "Log", "Logit"
    ),
  Link = c(
    "\\(\\beta X = \\mu\\)", 
    "\\(\\beta X = \\mu^{-1}\\)", 
    "\\(\\beta X = ln(\\mu)\\)", 
    "\\(\\beta X = ln\\left(\\frac{\\mu}{1-\\mu}\\right)\\)"
    ),
  Mean = c(
    "\\(\\mu = \\beta X\\)", 
    "\\(\\mu = (\\beta X)^{-1}\\)", 
    "\\(\\mu = exp(\\beta X)\\)", 
    "\\(\\mu = \\frac{1}{1 + exp(-\\beta X)}\\)"
  )
) %>% 
  kbl(table.attr = "class='table-model table-fullwidth'") %>% 
  kable_styling(c("striped", "hover"))

```

]

]




---

## Gaussian GLM

.pull-left[

```{r}

darts_gg <- ggplot(darts, aes(weight, length)) +
  labs(
    x = "Weight (g)",
    y = "Length (mm)"
  )

darts_gg + 
  geom_point(size = 3, alpha = 0.6)

```

.right.gray[_Archaic dart points. Sample from Fort Hood, Texas_]

]

.pull-right[

Assume that dart length is normally distributed with

$Var(\epsilon) = \sigma^2$  
$E(Y) = \mu = \beta X$  

__Question__ What is the probability that we observe these data (the lengths) given a model with parameters $\beta$ and $\sigma^2$? 

Likelihood function for the normal is 

$$\mathcal{L} = \prod_{i=1}^{n} N(X_i,\, y_i,\, \beta,\, \sigma^2)$$

]

---

## Gaussian GLM

```{r}

log_likelihood <- function(e){
  
  n <- length(e)
  
  sigma <- sqrt(sum(e^2)/n)
  
  sum(dnorm(e, mean = 0, sd = sigma, log = TRUE))

}

```

```{r}

darts_glm <- glm(length ~ weight, data = darts)

b0 <- coefficients(darts_glm)[["(Intercept)"]]
b1 <- coefficients(darts_glm)[["weight"]]

b0_hat <- round(b0 - 2, digits = 2)
b1_hat <- round(b1 - 1, digits = 2)

yhat <- b0_hat + b1_hat * darts$weight

errors <- with(darts, length - yhat)

sigma <- sqrt(sum(errors^2)/length(errors))

ll <- round(log_likelihood(errors), digits = 2)

```

.pull-left[

```{r}

darts_gg +
  geom_segment(
    data = tibble(
      x = darts$weight,
      y = darts$length,
      xend = x,
      yend = yhat
    ),
    aes(x, y, xend = xend, yend = yend),
    color = "gray60",
    linetype = "dashed"
  ) +
  geom_abline(
    intercept = b0_hat,
    slope = b1_hat
  ) +
  geom_point(size = 3, alpha = 0.6) +
  geom_text(
    data = tibble(
      x = min(darts$weight), 
      y = with(darts, c(max(length), max(length) - 5)), 
      label = c(
        paste0("\U03B2", "0 = ", b0_hat),
        paste0("\U03B2", "1 = ", b1_hat)
      )
    ),
    aes(x, y, label = label),
    size = 7,
    hjust = 0,
    vjust = 1
  )

```

.right.gray[_Archaic dart points. Sample from Fort Hood, Texas_]

]

.pull-right[

```{r}

ggplot() + 
  geom_histogram(
    data = tibble(x = errors),
    aes(x, y = ..density..),
    bins = 15,
    color = "black",
    fill = "gray65"
  ) +
  geom_polygon(
    data = bind_rows(
      c(x = -10, y = 0),
      tibble(
        x = seq(-10, 45, length = 300), 
        y = dnorm(x, sd = sigma)
      )
    ),
    aes(x, y),
    color = qaad_colors("celadon_blue"),
    fill = alpha(qaad_colors("celadon_blue"), 0.5),
    size = 1
  ) +
  annotate(
    "segment",
    x = 0,
    y = 0,
    xend = 0,
    yend = dnorm(1, 0, sigma),
    color = qaad_colors("rufous_red"),
    size = 1.5
  ) +
  labs(
    x = "Residuals",
    y = "Density"
  ) +
  coord_cartesian(
    xlim = c(-10, 45)
  ) +
  annotate(
    "text",
    label = paste0("logLik", " = ", ll),
    x = Inf,
    y = Inf,
    hjust = 1.1,
    vjust = 2,
    size = 7
  ) +
  scale_y_continuous(
    labels = scales::label_number(accuracy = 0.01)
  )

```

]

---

## Gaussian GLM

```{r}

log_likelihood <- function(e){
  
  n <- length(e)
  
  sigma_ML <- sqrt(sum(e^2)/n)
  
  sum(dnorm(e, mean = 0, sd = sigma_ML, log = TRUE))

}

```

```{r}

darts_glm <- glm(length ~ weight, data = darts)

b0 <- coefficients(darts_glm)[["(Intercept)"]]
b1 <- coefficients(darts_glm)[["weight"]]

b0_hat <- round(b0 - 1, digits = 2)
b1_hat <- round(b1 - 0.5, digits = 2)

yhat <- b0_hat + b1_hat * darts$weight

errors <- with(darts, length - yhat)

sigma <- sqrt(sum(errors^2)/length(errors))

ll <- round(log_likelihood(errors), digits = 2)

```

.pull-left[

```{r}

darts_gg +
  geom_segment(
    data = tibble(
      x = darts$weight,
      y = darts$length,
      xend = x,
      yend = yhat
    ),
    aes(x, y, xend = xend, yend = yend),
    color = "gray60",
    linetype = "dashed"
  ) +
  geom_abline(
    intercept = b0_hat,
    slope = b1_hat
  ) +
  geom_point(size = 3, alpha = 0.6) +
  geom_text(
    data = tibble(
      x = min(darts$weight), 
      y = with(darts, c(max(length), max(length) - 5)), 
      label = c(
        paste0("\U03B2", "0 = ", b0_hat),
        paste0("\U03B2", "1 = ", b1_hat)
      )
    ),
    aes(x, y, label = label),
    size = 7,
    hjust = 0,
    vjust = 1
  )

```

.right.gray[_Archaic dart points. Sample from Fort Hood, Texas_]

]

.pull-right[

```{r}

ggplot() + 
  geom_histogram(
    data = tibble(x = errors),
    aes(x, y = ..density..),
    bins = 15,
    color = "black",
    fill = "gray65"
  ) +
  geom_polygon(
    data = bind_rows(
      c(x = -10, y = 0),
      tibble(
        x = seq(-10, 45, length = 300), 
        y = dnorm(x, sd = sigma)
      )
    ),
    aes(x, y),
    color = qaad_colors("celadon_blue"),
    fill = alpha(qaad_colors("celadon_blue"), 0.5),
    size = 1
  ) +
  annotate(
    "segment",
    x = 0,
    y = 0,
    xend = 0,
    yend = dnorm(1, 0, sigma),
    color = qaad_colors("rufous_red"),
    size = 1.5
  ) +
  labs(
    x = "Residuals",
    y = "Density"
  ) +
  coord_cartesian(
    xlim = c(-10, 45)
  ) +
  annotate(
    "text",
    label = paste0("logLik", " = ", ll),
    x = Inf,
    y = Inf,
    hjust = 1.1,
    vjust = 2,
    size = 7
  ) +
  scale_y_continuous(
    labels = scales::label_number(accuracy = 0.01)
  )

```

]

---

## Gaussian GLM

```{r}

log_likelihood <- function(e){
  
  n <- length(e)
  
  sigma_ML <- sqrt(sum(e^2)/n)
  
  sum(dnorm(e, mean = 0, sd = sigma_ML, log = TRUE))

}

```

```{r}

darts_glm <- glm(length ~ weight, data = darts)

b0 <- coefficients(darts_glm)[["(Intercept)"]]
b1 <- coefficients(darts_glm)[["weight"]]

b0_hat <- round(b0 - 0, digits = 2)
b1_hat <- round(b1 - 0, digits = 2)

yhat <- b0_hat + b1_hat * darts$weight

errors <- with(darts, length - yhat)

sigma <- sqrt(sum(errors^2)/length(errors))

ll <- round(log_likelihood(errors), digits = 2)

```

.pull-left[

```{r}

darts_gg +
  geom_segment(
    data = tibble(
      x = darts$weight,
      y = darts$length,
      xend = x,
      yend = yhat
    ),
    aes(x, y, xend = xend, yend = yend),
    color = "gray60",
    linetype = "dashed"
  ) +
  geom_abline(
    intercept = b0_hat,
    slope = b1_hat
  ) +
  geom_point(size = 3, alpha = 0.6) +
  geom_text(
    data = tibble(
      x = min(darts$weight), 
      y = with(darts, c(max(length), max(length) - 5)), 
      label = c(
        paste0("\U03B2", "0 = ", b0_hat),
        paste0("\U03B2", "1 = ", b1_hat)
      )
    ),
    aes(x, y, label = label),
    size = 7,
    hjust = 0,
    vjust = 1
  )

```

.right.gray[_Archaic dart points. Sample from Fort Hood, Texas_]

]

.pull-right[

```{r}

ggplot() + 
  geom_histogram(
    data = tibble(x = errors),
    aes(x, y = ..density..),
    bins = 15,
    color = "black",
    fill = "gray65"
  ) +
  geom_polygon(
    data = bind_rows(
      c(x = -10, y = 0),
      tibble(
        x = seq(-10, 45, length = 300), 
        y = dnorm(x, sd = sigma)
      )
    ),
    aes(x, y),
    color = qaad_colors("celadon_blue"),
    fill = alpha(qaad_colors("celadon_blue"), 0.5),
    size = 1
  ) +
  annotate(
    "segment",
    x = 0,
    y = 0,
    xend = 0,
    yend = dnorm(1, 0, sigma),
    color = qaad_colors("rufous_red"),
    size = 1.5
  ) +
  labs(
    x = "Residuals",
    y = "Density"
  ) +
  coord_cartesian(
    xlim = c(-10, 45)
  ) +
  annotate(
    "text",
    label = paste0("logLik", " = ", ll),
    x = Inf,
    y = Inf,
    hjust = 1.1,
    vjust = 2,
    size = 7
  ) +
  scale_y_continuous(
    labels = scales::label_number(accuracy = 0.01)
  )

```

]

---

## Binomial GLM

.pull-left[

```{r}

snodgrass_gg <- ggplot(snodgrass, aes(area, inside)) +
  labs(
    x = "Area of House Structure (sq ft)",
    y = "Inside Inner Wall"
  ) +
  scale_y_continuous(
    breaks = c(0, 0.5, 1),
    labels = c(0, 0.5, 1)
  )

snodgrass_gg + 
  geom_point(size = 3, alpha = 0.4)

```

.gray.right[Data from the Snodgrass site in Butler County, MS]

]

.pull-right[

Location inside or outside of the inner wall at the Snodgrass site is a binomially distributed variable and has expectation $E(Y) = p$ where

$$p = \frac{1}{1 + exp(-\beta X)}$$
This defines a [logistic curve](https://en.wikipedia.org/wiki/Logistic_function) or sigmoid, with $p$ being the probability of success. This constrains the estimate $E(Y)$ to be in the range 0 to 1.

]

---

## Binomial GLM

.pull-left[

```{r}

snodgrass_gg + 
  geom_point(size = 3, alpha = 0.4)

```

.gray.right[Data from the Snodgrass site in Butler County, MS]

]

.pull-right[

Taking the log of $p$ gives us

$$log(p) = log\left(\frac{p}{1 - p}\right) = \beta X$$

This is known as the "logit" or [log odds](https://en.wikipedia.org/wiki/Logit). 

__Question__ What is the probability that we observe these data (these admissions) given a model with parameters $\beta$? 

]

---

## Binomial GLM

```{r}

snodgrass_glm <- glm(
  inside ~ area,
  family = binomial(link = "logit"),
  data = snodgrass
)

b0 <- coefficients(snodgrass_glm)[["(Intercept)"]]
b1 <- coefficients(snodgrass_glm)[["area"]]

ll <- logLik(snodgrass_glm)

```

.pull-left[

```{r}

snodgrass_gg + 
  geom_point(size = 3, alpha = 0.4) +
  geom_line(
    data = tibble(
      x = snodgrass$area,
      y = predict(snodgrass_glm, type = "response")
    ),
    aes(x, y),
    color = qaad_colors("rufous_red"),
    size = 1
  )

```

.gray.right[Data from the Snodgrass site in Butler County, MS]

]

.pull-right[

Estimated coefficients: 

$\beta_0 = `r b0`$  
$\beta_1 = `r b1`$   

For these, the log Likelihood is

$\mathcal{l} = `r ll`$  

]

---

## Binomial GLM

.pull-left[

```{r}

snodgrass_gg + 
  geom_point(size = 3, alpha = 0.4) +
  geom_line(
    data = tibble(
      x = snodgrass$area,
      y = predict(snodgrass_glm, type = "response")
    ),
    aes(x, y),
    color = qaad_colors("rufous_red"),
    size = 1
  )

```

.gray.right[Data from the Snodgrass site in Butler County, MS]

]

.pull-right[

Estimated coefficients: 

$\beta_0 = `r b0`$  
$\beta_1 = `r b1`$   

Note that these coefficient estimates are log-odds! To get the odds, we take the exponent.   

$\beta_0 = exp(`r b0`) = `r exp(b0)`$  
$\beta_1 = exp(`r b1`) = `r exp(b1)`$  

For a one unit increase in area, the odds of being in the inside wall increase by `r exp(b1)`.

]

---

## Binomial GLM

```{r}

est <- predict(
  snodgrass_glm, 
  newdata = tibble(area = 300), 
  type = "response"
)

```

.pull-left[

```{r}

snodgrass_gg + 
  geom_point(size = 3, alpha = 0.4) +
  geom_line(
    data = tibble(
      x = snodgrass$area,
      y = predict(snodgrass_glm, type = "response")
    ),
    aes(x, y),
    color = qaad_colors("rufous_red"),
    size = 1
  ) +
  annotate(
    "segment",
    x = 300,
    xend = 300,
    y = 0,
    yend = est,
    linetype = "dashed",
    color = "gray50"
  ) +
  annotate(
    "segment",
    x = min(snodgrass$area),
    xend = 300,
    y = est,
    yend = est,
    linetype = "dashed",
    color = "gray50"
  ) +
  annotate(
    "point", 
    x = 300,
    y = est,
    size = 5
  )

```

.gray.right[Data from the Snodgrass site in Butler County, MS]

]

.pull-right[

Estimated coefficients: 

$\beta_0 = `r b0`$  
$\beta_1 = `r b1`$   

To get the probability, we can use the mean function (also known as the inverse link):

$$p = \frac{1}{1+exp(-\beta X)}$$

For a house structure with an area of 300 square feet, the estimated probability that it occurs inside the inner wall is `r est`.

]

---

## Poisson GLM

.pull-left[

```{r}

sites_gg <- ggplot(sites, aes(elevation, sites)) +
  labs(
    x = "Elevation (km)",
    y = "Site count"
  )

sites_gg + 
  geom_point(
    size = 3,
    alpha = 0.7
  )

```

.gray.right[_Hypothetical dataset for demonstration._]

]

.pull-right[

Site counts is a Poisson distributed variable and has expectation $E(Y) = \lambda$ where

$$\lambda = exp(\beta X)$$

By taking the exponent, this constrains the expected count to be greater than zero.

]

---

## Poisson GLM

.pull-left[

```{r}

sites_gg + 
  geom_point(
    size = 3,
    alpha = 0.7
  )

```

.gray.right[_Hypothetical dataset for demonstration._]

]

.pull-right[

Taking the log gives us

$$log(\lambda) = \beta X$$

Now there is a linear relation to expected value.

]

---

## Poisson GLM

```{r}

sites_glm <- glm(
  sites ~ elevation,
  family = poisson,
  data = sites
)

b0 <- coefficients(sites_glm)[["(Intercept)"]]
b1 <- coefficients(sites_glm)[["elevation"]]

ll <- logLik(sites_glm)

```


.pull-left[

```{r}

sites_gg + 
  geom_line(
    data = tibble(
      x = sites$elevation,
      y = predict(sites_glm, type = "response")
    ),
    aes(x, y),
    color = qaad_colors("rufous_red"),
    size = 1
  ) +
  geom_point(
    size = 3,
    alpha = 0.7
  )

```

.gray.right[_Hypothetical dataset for demonstration._]

]

.pull-right[

Estimated coefficients: 

$\beta_0 = `r b0`$  
$\beta_1 = `r b1`$   

For these, the log Likelihood is

$\mathcal{l} = `r ll`$  

]

---

## Poisson GLM

.pull-left[

```{r}

sites_gg + 
  geom_line(
    data = tibble(
      x = sites$elevation,
      y = predict(sites_glm, type = "response")
    ),
    aes(x, y),
    color = qaad_colors("rufous_red"),
    size = 1
  ) +
  geom_point(
    size = 3,
    alpha = 0.7
  )

```

.gray.right[_Hypothetical dataset for demonstration._]

]

.pull-right[

Estimated coefficients: 

$\beta_0 = `r b0`$  
$\beta_1 = `r b1`$   

Note that these coefficient estimates are on the log scale! To get back to counts, need to take the exponent.   

$\beta_0 = exp(`r b0`) = `r exp(b0)`$  
$\beta_1 = exp(`r b1`) = `r exp(b1)`$  

For a one unit increase in elevation, the count of sites increases by `r exp(b1)`.

]

---

## Deviance

- Measure of goodness of fit, so __smaller is better__
- Gives the difference in log-Likelihood between a model $M$ and a __saturated__ model $M_S$
    - $M_S$ has a parameter for each observation (i.e., zero degrees of freedom)

$$D = -2(log\,\mathcal{L}(M_1) - log\,\mathcal{L}(M_S))$$

- __Residual deviance__ = deviance of proposed model
- __Null deviance__ = deviance of null model (ie, intercept-only model)

---

## Information Criteria

- A function of the deviance, so __smaller is better__
- Two penalties: 
    - $n$ = number of observations 
    - $p$ = number of parameters
- Akaike Information Criteria
    - $AIC = D + 2p$
- Bayesian Information Criteria 
    - $BIC = D + (p \cdot log(n))$

---

## ANOVA

- Analysis of Deviance _aka_ Likelihood Ratio Test
- For Binomial and Poisson models
- Test statistic is the logged ratio of likelihoods between proposed, $M_1$, and null, $M_0$, models

$$
\begin{aligned}
\chi^2 &= -2\,log\frac{\mathcal{L}(M_0)}{\mathcal{L}(M_1)}\\\\
&= -2[log\,\mathcal{L}(M_0) - log\,\mathcal{L}(M_1)]
\end{aligned}
$$

- Compare to a $\chi^2$ distribution with $k$ degrees of freedom - asks what the probability is of getting a value greater than the observed $\chi^2$ value.
- Null hypothesis is no difference in log-Likelihood.

---

## &#x1F52D; Looking Ahead

```{r}

schedule <- here("_misc", "course_outline.xlsx") %>% 
  readxl::read_excel(1) %>% 
  mutate(
    date = lubridate::ymd(date),
    description = glue::glue(
    "<b>{topic}</b> <span style = 'font-size: 80%; color: #9C9C9C;'>({date})</span><br>
    <div style = 'font-size: 80%; font-style: italic; color: #9C9C9C;'>{details}</div>"
  )) %>% 
  select(description)

```

.pull-left[

```{r}

schedule %>% 
  slice(11:14) %>% 
  kbl(escape = FALSE, col.names = NULL) %>% 
  kable_paper(c("striped", "hover"))

```

]