---
title: "Lecture 10: Transforming Variables"  
date: 'Last updated: `r Sys.Date()`'
---

```{r setup}
#| include: false

here::here("slides", "_setup.R") |> source()

kidiq <- here("_misc", "kidiq.csv") |> read_csv()

model_table_defaults <- function(x){

  tab_options(
    x,
    column_labels.font.weight = "bold",
    container.width = pct(100),
    table.border.top.style = "hidden",
    table.font.size = px(20),
    table.width = pct(100)
  )
  
}

```

## ðŸ“‹ Lecture Outline

- Qualitative variables
    - Dummy variables
    - Intercept offset
    - t-test
- Centering 
- Scaling
- Interactions
- Polynomial transformations for non-linearity
- Log transformations for non-linearity

## Qualitative Variables

```{r}

set.seed(42)

n <- 100

income <- tibble(
  sex = c(rep("female", n), rep("male", n)),
  education = rpois(n, 5) + ifelse(sex == "female", 10, 2),
  error = rnorm(2*n),
  income = ifelse(
    sex == "female",
    1 + 0.5 * education + error,
    14 + 0.8 * education + error
  )
)

```

::::::::: {.r-stack}
:::::::: {.fragment .fade-out fragment-index=0 .mt-0 .w100}
:::::: {.columns}
::::: {.column}

```{r}

limits <- 
  with(
  income,
  tibble(
    x = c(min(education)-1, max(education)+1),
    y = c(min(income)-1, max(income)+1)
  )
)

gg_income <- 
  ggplot(income) +
  scale_color_manual(
    name = NULL,
    values = qcolors("sapphire", "bronze")
  ) +
  labs(
    x = "Education (Years)", 
    y = "Income ($1000)"
  ) +
  theme(legend.position = "none") +
  coord_cartesian(
    x = limits[["x"]],
    y = limits[["y"]],
    expand = FALSE
  )

gg_income + 
  geom_point(
    aes(education, income, color = sex),
    size = 3,
    alpha = 0.6
  ) +
  geom_text(
    data = tibble(x = c(12, 10), y = c(21, 8), sex = c("male", "female")),
    aes(x, y, label = sex, color = sex),
    hjust = c(0, 1),
    size = 13/.pt
  )

```

:::::
::::: {.column}

:::::
::::::
::::::::
:::::::: {.fragment .fade-in-then-out fragment-index=0 .mt-0 .w100}
:::::: {.columns}
::::: {.column}
```{r}

lm_null <- lm(income ~ education, data = income)

new_data <- tibble(
  education = seq(0, 30, length = n*2)
)

estimate_null <- predict(lm_null, newdata = new_data, se.fit = TRUE)

estimate_null <- 
  with(
    estimate_null,
    tibble(
      y    = fit,
      ymin = fit - 2*se.fit,
      ymax = fit + 2*se.fit
    )
  ) |> 
  bind_cols(new_data)

gg_income + 
  geom_ribbon(
    data = estimate_null,
    aes(education, ymin = ymin, ymax = ymax),
    fill = "gray95",
    color = "transparent"
  ) +
  geom_line(
    data = estimate_null,
    aes(education, y)
  ) + 
  geom_point(
    aes(education, income, color = sex),
    size = 3,
    alpha = 0.6
  ) +
  geom_text(
    data = tibble(x = c(12, 10), y = c(21, 8), sex = c("male", "female")),
    aes(x, y, label = sex, color = sex),
    hjust = c(0, 1),
    size = 13/.pt
  )

```
:::::
::::: {.column}
```{r}

broom::tidy(lm_null) |> 
  rename("t.statistic" = statistic) |> 
  mutate(across(where(is.numeric), ~round(.x, 3))) |> 
  gt() |> 
  tab_row_group("Sex Insensitive Model", 1:2, id = "bob") |> 
  tab_style(
    style = list(cell_text(weight = "bold", color = qcolors("kobe"))),
    locations = cells_row_groups("bob")
  ) |> 
  model_table_defaults() |> 
  as_raw_html()

```
:::::
::::::
::::::::
:::::::: {.fragment .fade-in fragment-index=1 .mt-0 .w100}
:::::: {.columns}
::::: {.column}
```{r}

lm_sex <- lm(income ~ education + sex, data = income)

new_data <- tibble(
  education = rep(seq(0, 30, length = n*2), 2),
  sex = c(rep("male", n*2), rep("female", n*2))
)

estimate_sex <- 
  with(
    predict(lm_sex, newdata = new_data, se.fit = TRUE),
    tibble(
      y    = fit,
      ymin = fit - 2*se.fit,
      ymax = fit + 2*se.fit
    )
  ) |> 
  bind_cols(new_data)

gg_income + 
  geom_ribbon(
    data = estimate_sex,
    aes(education, ymin = ymin, ymax = ymax, group = sex),
    fill = "gray95",
    color = "transparent"
  ) +
  geom_line(
    data = estimate_sex,
    aes(education, y, group = sex)
  ) + 
  geom_point(
    aes(education, income, color = sex),
    size = 3,
    alpha = 0.6
  ) +
  geom_text(
    data = tibble(x = c(12, 10), y = c(21, 8), sex = c("male", "female")),
    aes(x, y, label = sex, color = sex),
    hjust = c(0, 1),
    size = 13/.pt
  )

```
:::::
::::: {.column}
```{r}

bind_rows(
  broom::tidy(lm_null),
  broom::tidy(lm_sex)
) |> 
  rename("t.statistic" = statistic) |> 
  mutate(
    term = str_replace(term, "sexmale", "sex(male)"),
    across(where(is.numeric), ~round(.x, 3))
  ) |> 
  gt() |> 
  tab_row_group("Sex Sensitive Model", 3:5, id = "tom") |> 
  tab_row_group("Sex Insensitive Model", 1:2, id = "bob") |>
  tab_style(
    style = list(cell_text(weight = "bold", color = qcolors("kobe"))),
    locations = cells_row_groups(c("bob", "tom"))
  ) |> 
  model_table_defaults() |> 
  as_raw_html()

```
:::::
::::::
::::::::
:::::::::

## Dummy Variables

::::::::: {.r-stack}
:::::::: {.fragment .fade-out fragment-index=0 .mt-0 .w100}
:::::: {.columns}
::::: {.column}
__Dummy variables__ code each observation:

$x_i=\left\{\begin{array}{c l} 1 & \text{if i in category} \\ 0 & \text{if i not in category} \end{array}\right.$

with zero being the __reference class__.

For _m_ categories, requires _m-1_ dummy variables.
:::::
::::: {.column}
```{r}

tibble(
  Category = c("D<sub>1</sub>", "D<sub>2</sub>", "&#8942;", "D<sub>m-1</sub>", "D<sub>m</sub>"),
  "D1" = c(1, 0, "&#8942;", 0, 0),
  "D2" = c(0, 1, "&#8942;", 0, 0),
  "D3" = c(0, 0, "&#8942;", 0, 0),
  "D4" = c(0, 0, "&#8942;", 1, 0)
) |> 
  gt() |> 
  fmt_markdown(columns = everything()) |> 
  cols_label(
    D1 = gt::html("D<sub>1</sub>"),
    D2 = gt::html("D<sub>2</sub>"),
    D3 = gt::html("&#8230;"),
    D4 = gt::html("D<sub>m-1</sub>")
  ) |> 
  tab_options(
    column_labels.font.weight = "bold",
    table.border.top.style = "hidden",
    table.font.size = px(20)
  ) |> 
  as_raw_html()

```
:::::
::::::
::::::::
:::::::: {.fragment .fade-in fragment-index=0 .mt-0 .w100}
:::::: {.columns}
::::: {.column}
__Dummy variables__ code each observation:

$sex_i=\left\{\begin{array}{c l} 1 & \text{if i-th person is male} \\ 0 & \text{if i-th person is not male} \end{array}\right.$

Here the __reference class__ is _female_.

The sex variable has two categories, hence one dummy variable.
:::::
::::: {.column}
```{r}

tibble(
  sex = c("male", "female", "&#8942;", "male", "female"),
  "male" = c(1, 0, "&#8942;", 1, 0)
) |> 
  gt() |> 
  fmt_markdown(columns = everything()) |>
  tab_options(
    column_labels.font.weight = "bold",
    table.border.top.style = "hidden",
    table.font.size = px(20)
  ) |> 
  as_raw_html()

```
:::::
::::::
::::::::
:::::::::

## Intercept Offset

```{css}

.blank-table table {
  margin: 0 auto 0 0;
}

.blank-table table td {
  padding: 0.6em 1em 0.6em 0;
  border-bottom: 0 !important;
}

.blank-table table tr[first-child] {
  display: none;
}

```

::: {.blank-table}

|                                    |                                                             |
|------------------------------------|-------------------------------------------------------------|
| Simple linear model                | $y_i = \beta_0 + \beta_1x_1 + \epsilon_i$                   |
| Add dummy $D$ with $\gamma$ offset | $y_i = (\beta_0 + \gamma\cdot D) + \beta_1x_1 + \epsilon_i$      |
| _for a binary class:_              |                                                             |
| Model for reference class          | $y_i = (\beta_0 + \gamma\cdot 0) + \beta_1x_1 + \epsilon_i$ |
| Model for other class              | $y_i = (\beta_0 + \gamma\cdot 1) + \beta_1x_1 + \epsilon_i$ |

:::

## t-test

:::::: {.columns}
::::: {.column}
```{r}

sex_mean <- income |> 
  group_by(sex) |> 
  summarize(intercept = mean(income)) |> 
  mutate(
    x = c(0.5, 0.5),
    xend = c(1, 2)
  )

ggplot(income) +
  scale_color_manual(
    name = NULL,
    values = qcolors("sapphire", "bronze")
  ) +
  labs(
    x = "Sex", 
    y = "Income ($1000)"
  ) +
  theme(
    legend.position = "none"
  ) +
  geom_point(
    aes(sex, income, color = sex),
    size = 3.5,
    alpha = 0.6
  ) +
  geom_segment(
    data = sex_mean,
    aes(x, intercept, xend = xend, yend = intercept)
  ) +
  coord_cartesian(
    xlim = c(0.5, 2.5),
    ylim = with(income, c(min(income)-1, max(income)+1)),
    expand = FALSE
  )

```
:::::
::::: {.column}
```{r}

lm_ttest <- lm(income ~ sex, data = income)

broom::tidy(lm_ttest) |> 
  rename("t.statistic" = statistic) |> 
  mutate(
    term = str_replace(term, "sexmale", "sex(male)"),
    across(where(is.numeric), ~round(.x, 3))
  ) |> 
  gt() |> 
  tab_row_group("Sex Sensitive Model", 1:2, id = "tom") |> 
  tab_style(
    style = list(cell_text(weight = "bold", color = qcolors("kobe"))),
    locations = cells_row_groups("tom")
  ) |> 
  model_table_defaults() |> 
  as_raw_html()

```

$H_{0}$: no difference in mean

model: $income \sim sex$

$\bar{y}_{F} = `r round(sex_mean$intercept[[1]], 3)`$  
$\bar{y}_{M} = \bar{y}_{F} + 11.165 = `r round(sex_mean$intercept[[2]], 3)`$   
:::::
::::::

## Centering

::::::::: {.r-stack}
:::::::: {.fragment .fade-out fragment-index=0 .mt-0 .w100}
:::::: {.columns}
::::: {.column}
```{r}

lm_kid <- lm(kid ~ mom, data = kidiq)

ggplot() + 
  geom_point(
    data = kidiq, 
    aes(mom, kid),
    size = 3
  ) +
  geom_abline(
    intercept = coefficients(lm_kid)[["(Intercept)"]],
    slope = coefficients(lm_kid)[["mom"]],
    color = qcolors("blue"),
    linewidth = 1
  ) +
  labs(
    x = "Mom's IQ",
    y = "Kid's IQ"
  ) +
  coord_cartesian(
    xlim = with(kidiq, c(min(mom)-5, max(mom)+5)),
    ylim = with(kidiq, c(min(kid)-5, max(kid)+5)),
    expand = FALSE
  )

```
:::::
::::: {.column}
```{r}

broom::tidy(lm_kid) |> 
  rename("t.statistic" = statistic) |> 
  mutate(
    term = c("(Intercept)", "Mom's IQ"),
    across(where(is.numeric), ~round(.x, 3))
  ) |>
  gt() |> 
  tab_row_group("Uncentered Model", 1:2, id = "bob") |> 
  tab_style(
    style = list(cell_text(weight = "bold", color = qcolors("kobe"))),
    locations = cells_row_groups("bob")
  ) |> 
  model_table_defaults() |> 
  as_raw_html()

```

Model child IQ as a function of mother's IQ.

:::::
::::::
::::::::
:::::::: {.fragment .fade-in-then-out fragment-index=0 .mt-0 .w100}
:::::: {.columns}
::::: {.column}
```{r}

ggplot() + 
  geom_point(
    data = kidiq, 
    aes(mom, kid),
    size = 3
  ) +
  geom_abline(
    intercept = coefficients(lm_kid)[["(Intercept)"]],
    slope = coefficients(lm_kid)[["mom"]],
    color = qcolors("blue"),
    linewidth = 1
  ) +
  labs(
    x = "Mom's IQ",
    y = "Kid's IQ"
  ) +
  coord_cartesian(
    xlim = with(kidiq, c(-5, max(mom)+5)),
    ylim = with(kidiq, c(min(kid)-5, max(kid)+5)),
    expand = FALSE
  ) +
  annotate(
    "segment",
    x = 0, xend = 0,
    y = 65, yend = coefficients(lm_kid)[["(Intercept)"]] + 2,
    color = qcolors("kobe"),
    arrow = arrow(length = unit(0.15, "inches"), type = "closed")
  ) +
  annotate(
    "text",
    x = 0,
    y = 67,
    label = "What is this?!!!",
    color = qcolors("kobe"),
    size = 14/.pt,
    hjust = 0,
    vjust = 0
  )

```
:::::
::::: {.column}
```{r}

broom::tidy(lm_kid) |> 
  rename("t.statistic" = statistic) |> 
  mutate(
    term = c("(Intercept)", "Mom's IQ"),
    across(where(is.numeric), ~round(.x, 3))
  ) |>
  gt() |> 
  tab_row_group("Uncentered Model", 1:2, id = "bob") |> 
  tab_style(
    style = list(cell_text(weight = "bold", color = qcolors("kobe"))),
    locations = cells_row_groups("bob")
  ) |> 
  model_table_defaults() |> 
  as_raw_html()

```

Model child IQ as a function of mother's IQ.
:::::
::::::
::::::::
:::::::: {.fragment .fade-in-then-out fragment-index=1 .mt-0 .w100}
:::::: {.columns}
::::: {.column}
```{r}

ggplot() + 
  geom_point(
    data = kidiq, 
    aes(mom, kid),
    size = 3
  ) +
  geom_abline(
    intercept = coefficients(lm_kid)[["(Intercept)"]],
    slope = coefficients(lm_kid)[["mom"]],
    color = qcolors("blue"),
    linewidth = 1
  ) +
  labs(
    x = "Mom's IQ",
    y = "Kid's IQ"
  ) +
  coord_cartesian(
    xlim = with(kidiq, c(-5, max(mom)+5)),
    ylim = with(kidiq, c(min(kid)-5, max(kid)+5)),
    expand = FALSE
  ) +
  annotate(
    "segment",
    x = 0, xend = 0,
    y = 65, yend = coefficients(lm_kid)[["(Intercept)"]] + 2,
    color = qcolors("kobe"),
    arrow = arrow(length = unit(0.15, "inches"), type = "closed")
  ) +
  annotate(
    "text",
    x = 0,
    y = 67,
    label = "What is this?!!!",
    color = qcolors("kobe"),
    size = 14/.pt,
    hjust = 0,
    vjust = 0
  )

```
:::::
::::: {.column}
Centering will make intercept more interpretable.

To center, subtract the mean:

$$\text{Mom's IQ} - \text{mean(Mom's IQ)}$$

That gets us this model...
:::::
::::::
::::::::
:::::::: {.fragment .fade-in fragment-index=2 .mt-0 .w100}
:::::: {.columns}
::::: {.column}
```{r}

kidiq_scale <- kidiq |> mutate(mom = scale(mom, scale = FALSE)[,1])

lm_kid_scale <- lm(kid ~ mom, data = kidiq_scale)

ggplot() + 
  geom_vline(
    xintercept = 0,
    color = qcolors("kobe"),
    linetype = "dashed",
    linewidth = 1
  ) +
  geom_point(
    data = kidiq, 
    aes(mom - mean(mom), kid),
    size = 3
  ) +
  geom_abline(
    intercept = coefficients(lm_kid_scale)[["(Intercept)"]],
    slope = coefficients(lm_kid_scale)[["mom"]],
    color = qcolors("blue"),
    linewidth = 1
  ) +
  labs(
    x = "Mom's IQ",
    y = "Kid's IQ"
  ) +
  coord_cartesian(
    xlim = with(kidiq_scale, c(min(mom)-5, max(mom)+5)),
    ylim = with(kidiq_scale, c(min(kid)-5, max(kid)+5)),
    expand = FALSE
  )

```
:::::
::::: {.column}
```{r}

bind_rows(
  broom::tidy(lm_kid),
  broom::tidy(lm_kid_scale)
) |> 
  rename("t.statistic" = statistic) |> 
  mutate(
    term = c("(Intercept)", "Mom's IQ", "(Intercept)", "Mom's IQ"),
    across(where(is.numeric), ~round(.x, 3))
  ) |>
  gt() |> 
  tab_row_group("Centered Model", 3:4, id = "tom") |> 
  tab_row_group("Uncentered Model", 1:2, id = "bob") |>
  tab_style(
    style = list(cell_text(weight = "bold", color = qcolors("kobe"))),
    locations = cells_row_groups(c("bob", "tom"))
  ) |> 
  model_table_defaults() |> 
  as_raw_html()

```

Now we interpret the intercept as expected IQ of a child for a mother with mean IQ. (Notice change in standard error!)
:::::
::::::
::::::::
:::::::::

::: aside
No commitment here to IQ being meaningful. Just an example I'm using till I have time to find a better one.
:::

## Scaling

::::::::: {.r-stack}
:::::::: {.fragment .fade-out fragment-index=0 .mt-0 .w100}
:::::: {.columns}
::::: {.column}
```{r}
#| fig-asp: 1

lm_cars <- lm(mpg ~ hp + wt, data = mtcars)

mtcars_longer <- mtcars |> 
  select(mpg, hp, wt) |> 
  pivot_longer(
    -mpg,
    names_to = "variable",
    values_to = "x"
  )

new_data <- 
  tibble(
    hp = with(mtcars, seq(min(hp)-7, max(hp)+7, length = 300)),
    wt = with(mtcars, seq(min(wt)-0.13, max(wt)+0.13, length = 300))
  )

margins <-
  bind_rows(
    with(
      predict(
        lm_cars, 
        newdata = new_data |> mutate(wt = mean(mtcars$wt)), 
        se.fit = TRUE
      ),
      tibble(
        variable = "hp",
        x = new_data$hp,
        y = fit,
        ymin = fit - 2*se.fit,
        ymax = fit + 2*se.fit
      )
    ),
    with(
      predict(
        lm_cars, 
        newdata = new_data |> mutate(hp = mean(mtcars$hp)), 
        se.fit = TRUE
      ),
      tibble(
        variable = "wt",
        x = new_data$wt,
        y = fit,
        ymin = fit - 2*se.fit,
        ymax = fit + 2*se.fit
      )
    )
  )

gg_cars <- ggplot() + 
  geom_ribbon(
    data = margins,
    aes(x, ymin = ymin, ymax = ymax),
    fill = "gray95",
    color = "transparent"
  ) +
  geom_line(
    data = margins,
    aes(x,y)
  ) +
  geom_point(
    data = mtcars_longer, 
    aes(x, mpg),
    size = 3
  ) +
  labs(
    x = NULL,
    y = "Fuel Efficiency\n(Miles Per Gallon)"
  ) +
  coord_cartesian(
    ylim = with(mtcars, c(min(mpg)-2, max(mpg)+2)),
    expand = FALSE
  ) +
  facet_wrap(
    ~variable,
    nrow = 2,
    scale = "free_x",
    strip.position = "bottom",
    labeller = labeller(variable = c(wt = "Weight (1,000 pounds)", hp = "Horse Power"))
  ) +
  theme(
    axis.title.y = element_text(lineheight = 1.3),
    strip.background = element_blank(),
    strip.placement = "outside"
  )

gg_cars

```
:::::
::::: {.column}
```{r}

broom::tidy(lm_cars) |> 
  rename("t.statistic" = statistic) |> 
  mutate(
    term = c("(Intercept)", "Horsepower", "Weight"),
    across(where(is.numeric), ~round(.x, 3))
  ) |>
  gt() |> 
  tab_row_group("Unscaled Model", 1:3, id = "bob") |>
  tab_style(
    style = list(cell_text(weight = "bold", color = qcolors("kobe"))),
    locations = cells_row_groups("bob")
  ) |> 
  model_table_defaults() |> 
  as_raw_html()

```

Model fuel efficiency as a function of weight and horse power.  

__Question__ Which covariate has a bigger effect?  
:::::
::::::
::::::::
:::::::: {.fragment .fade-in-then-out fragment-index=0 .mt-0 .w100}
:::::: {.columns}
::::: {.column}
```{r}
#| fig-asp: 1

gg_cars

```
:::::
::::: {.column}
__Problem__ Cannot directly compare coefficients on different scales.  

__Solution__ Convert variable values to their z-scores: 

$$z_i = \frac{x_i-\bar{x}}{\sigma_{x}}$$

All coefficients now give change in $y$ for 1$\sigma$ change in $x$.  
:::::
::::::
::::::::
:::::::: {.fragment .fade-in fragment-index=1 .mt-0 .w100}
:::::: {.columns}
::::: {.column}
```{r}
#| fig-asp: 1

rescale <- function(x) scale(x)[,1]

mtcars_scaled <- mtcars |> mutate(across(c(hp, wt), rescale))

lm_cars_scaled <- lm(mpg ~ hp + wt, data = mtcars_scaled)

mtcars_longer <- mtcars_scaled |> 
  pivot_longer(
    -mpg,
    names_to = "variable",
    values_to = "x"
  ) |> 
  filter(variable %in% c("wt", "hp"))

new_data <- 
  tibble(
    hp = with(mtcars_longer, seq(min(x), max(x)+0.1, length = 300)),
    wt = hp
  )

margins <-
  bind_rows(
    with(
      predict(
        lm_cars_scaled, 
        newdata = new_data |> mutate(wt = 0), 
        se.fit = TRUE
      ),
      tibble(
        variable = "hp",
        x = new_data$hp,
        y = fit,
        ymin = fit - 2*se.fit,
        ymax = fit + 2*se.fit
      )
    ),
    with(
      predict(
        lm_cars_scaled, 
        newdata = new_data |> mutate(hp = 0), 
        se.fit = TRUE
      ),
      tibble(
        variable = "wt",
        x = new_data$wt,
        y = fit,
        ymin = fit - 2*se.fit,
        ymax = fit + 2*se.fit
      )
    )
  )

ggplot() + 
  geom_ribbon(
    data = margins,
    aes(x, ymin = ymin, ymax = ymax),
    fill = "gray95",
    color = "transparent"
  ) +
  geom_line(
    data = margins,
    aes(x, y)
  ) +
  geom_point(
    data = mtcars_longer, 
    aes(x, mpg),
    size = 3
  ) +
  labs(
    x = NULL,
    y = "Fuel Efficiency\n(Miles Per Gallon)"
  ) +
  coord_cartesian(
    ylim = with(margins, c(min(y)-2, max(y)+2)),
    expand = FALSE
  ) +
  facet_wrap(
    vars(variable), 
    nrow = 2,
    scale = "free_x",
    strip.position = "bottom",
    labeller = labeller(variable = c(wt = "Weight (z)", hp = "Horsepower (z)"))
  ) +
  theme(
    axis.title.y = element_text(lineheight = 1.3),
    strip.background = element_blank(),
    strip.placement = "outside"
  )

```

:::::
::::: {.column}
```{r}

bind_rows(
  broom::tidy(lm_cars),
  broom::tidy(lm_cars_scaled)
) |> 
  rename("t.statistic" = statistic) |> 
  mutate(
    term = c("(Intercept)", "Horsepower", "Weight", "(Intercept)", "Horsepower", "Weight"),
    across(where(is.numeric), ~round(.x, 3))
  ) |>
  gt() |> 
  tab_row_group("Scaled Model", 4:6, id = "tom") |> 
  tab_row_group("Unscaled Model", 1:3, id = "bob") |>
  tab_style(
    style = list(cell_text(weight = "bold", color = qcolors("kobe"))),
    locations = cells_row_groups(c("bob", "tom"))
  ) |> 
  model_table_defaults() |> 
  as_raw_html()

```
:::::
::::::
::::::::
:::::::::

## Interactions

::::::::: {.r-stack}
:::::::: {.fragment .fade-out fragment-index=0 .mt-0 .w100}
:::::: {.columns}
::::: {.column}
```{r}

kidiq_scale <- kidiq_scale |> mutate(mom_hs = factor(mom_hs))

gg_kid_hs <- ggplot() + 
  geom_point(
    data = kidiq_scale, 
    aes(mom, kid, color = mom_hs),
    size = 3
  ) +
  scale_color_manual(
    name = NULL,
    values = qcolors("gold", "eerie"),
    labels = c("No High School", "High School")
  ) +
  labs(
    x = "Mom's IQ",
    y = "Kid's IQ"
  ) +
  coord_cartesian(
    xlim = with(kidiq_scale, c(min(mom)-5, max(mom)+5)),
    ylim = with(kidiq_scale, c(min(kid)-5, max(kid)+5)),
    expand = FALSE
  ) +
  theme(
    legend.position = c(0.8, 0.14),
    legend.box.just = c("right", "bottom")
  )

gg_kid_hs

```
:::::
::::: {.column}
```{r}

broom::tidy(lm_kid_scale) |> 
  rename("t.statistic" = statistic) |> 
  mutate(
    term = c("(Intercept)", "Mom's IQ"),
    across(where(is.numeric), ~round(.x, 3))
  ) |>
  gt() |> 
  tab_row_group("Additive Model", 1:2, id = "bob") |> 
  tab_style(
    style = list(cell_text(weight = "bold", color = qcolors("kobe"))),
    locations = cells_row_groups("bob")
  ) |> 
  model_table_defaults() |> 
  as_raw_html()

```

__Question__ What if the relationship between child's IQ and mother's IQ depends on the mother's educational attainment?
:::::
::::::
::::::::
:::::::: {.fragment .fade-in-then-out fragment-index=0 .mt-0 .w100}
:::::: {.columns}
::::: {.column}
```{r}

gg_kid_hs

```
:::::
::::: {.column}
Simple formula becomes:

$$y_i = (\beta_0 + \gamma D) + (\beta_1x_1 + \omega D x_1) + \epsilon_i$$

With  
- $\gamma$ giving the change in $\beta_0$ and  
- $\omega$ giving the change in $\beta_1$.  

That's a __change in intercept__ and a __change in slope__!
:::::
::::::
::::::::
:::::::: {.fragment .fade-in fragment-index=1 .mt-0 .w100}
:::::: {.columns}
::::: {.column}
```{r}

lm_kidx <- lm(kid ~ mom * mom_hs, data = kidiq_scale)

new_data <- 
  tibble(
    mom = with(kidiq_scale, rep(seq(min(mom)-5, max(mom)+5, length = 300), 2)),
    mom_hs = factor(rep(c(0, 1), each = 300))
  )

margins <-
  with(
      predict(
        lm_kidx, 
        newdata = new_data, 
        se.fit = TRUE
      ),
      tibble(
        mom_hs = new_data$mom_hs,
        x = new_data$mom,
        y = fit,
        ymin = fit - 2*se.fit,
        ymax = fit + 2*se.fit
      )
    )

ggplot() + 
  geom_ribbon(
    data = margins,
    aes(x, ymin = ymin, ymax = ymax, group = mom_hs),
    fill = "gray95",
    color = "transparent"
  ) +
  geom_line(
    data = margins,
    aes(x, y, color = mom_hs),
    linewidth = 1
  ) + 
  geom_point(
    data = kidiq_scale, 
    aes(mom, kid, color = mom_hs),
    size = 3
  ) +
  scale_color_manual(
    name = NULL,
    values = qcolors("gold", "eerie"),
    labels = c("No High School", "High School")
  ) +
  labs(
    x = "Mom's IQ",
    y = "Kid's IQ"
  ) +
  coord_cartesian(
    xlim = with(kidiq_scale, c(min(mom)-5, max(mom)+5)),
    ylim = with(kidiq_scale, c(min(kid)-5, max(kid)+5)),
    expand = FALSE
  ) +
  theme(
    legend.position = c(0.8, 0.14),
    legend.box.just = c("right", "bottom")
  )

```
:::::
::::: {.column}
```{r}

bind_rows(
  broom::tidy(lm_kid_scale),
  broom::tidy(lm_kidx)
) |> 
  rename("t.statistic" = statistic) |> 
  mutate(
    term = c("(Intercept)", "Mom's IQ", "(Intercept)", "Mom's IQ", "High School", "Mom's IQ x HS"),
    across(where(is.numeric), ~round(.x, 3))
  ) |>
  gt() |> 
  tab_row_group("Interactive Model", 3:6, id = "tom") |> 
  tab_row_group("Additive Model", 1:2, id = "bob") |>
  tab_style(
    style = list(cell_text(weight = "bold", color = qcolors("kobe"))),
    locations = cells_row_groups(c("bob", "tom"))
  ) |> 
  model_table_defaults() |> 
  as_raw_html()

```
:::::
::::::
::::::::
:::::::::

## Polynomial transformations for non-linearity

:::::: {.r-stack}
::: {.fragment .fade-out fragment-index=0}

```{r}

set.seed(25)

n <- 100

observations <- tibble(
  x = seq(0, 1, length = n),
  y = 1 - 3.7 * (x - 0.5)^2 + (0.75 * x) + rnorm(n, sd = 0.2)
) |> 
  mutate(
    y = scales::rescale(y, c(0.1, 0.9))
  )

bad_mod <- lm(y ~ x, data = observations)

bad_b0 <- coefficients(bad_mod)[[1]]
bad_b1 <- coefficients(bad_mod)[[2]]

bad_fit <- tibble(
  observation = 1:n,
  estimate = predict(bad_mod),
  residuals = residuals(bad_mod)
)

bad_r2 <- summary(bad_mod)$r.squared

```

__Simple Linear Model:__ $y_{i} = \beta_{0} + \beta_{1}X + \epsilon_{i}$  
$R^2 = `r bad_r2`$

```{r}
#| out-width: 100%
#| layout-ncol: 2

ggplot(observations, aes(x,y)) +
  geom_abline(
    intercept = bad_b0,
    slope = bad_b1,
    color = qcolors("kobe"),
    size = 1
  ) + 
  geom_point(
    size = 3
  ) +
  labs(
    x = "X",
    y = "Y"
  ) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )

ggplot(bad_fit, aes(estimate, residuals)) +
  geom_segment(
    aes(xend = estimate, yend = 0),
    linetype = "dashed"
  ) +
  geom_hline(
    aes(yintercept = 0),
    color = qcolors("kobe"),
    size = 1
  ) +
  geom_point(
    size = 3
  ) +
  labs(
    x = "Estimates",
    y = "Residuals"
  ) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )

```

:::
::: {.fragment .fade-in fragment-index=0}

```{r}

meh_mod <- lm(y ~ poly(x, 2), data = observations)

meh_fit <- tibble(
  observation = 1:n,
  estimate = predict(meh_mod),
  residuals = residuals(meh_mod)
)

meh_r2 <- summary(meh_mod)$r.squared

```

__Quadratic Model:__ $y_{i} = \beta_{0} + \beta_{1}X + \beta_{2}X^{2} + \epsilon_{i}$  
$R^2=`r meh_r2`$

```{r}
#| out-width: 100%
#| layout-ncol: 2

ggplot(observations, aes(x,y)) +
  geom_line(
    aes(y = meh_fit$estimate),
    color = qcolors("kobe"),
    size = 1
  ) + 
  geom_point(
    size = 3
  ) +
  labs(
    x = "X",
    y = "Y"
  ) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )

ggplot(meh_fit, aes(estimate, residuals)) +
  geom_segment(
    aes(xend = estimate, yend = 0),
    linetype = "dashed"
  ) +
  geom_hline(
    aes(yintercept = 0),
    color = qcolors("kobe"),
    size = 1
  ) +
  geom_point(
    size = 3
  ) +
  labs(
    x = "Estimates",
    y = "Residuals"
  ) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )

```


:::
::::::

## Log transformations

```{r}

n <- 100

x <- seq(0.1,5,length.out = n)

set.seed(1)

e <- rnorm(n, mean = 0, sd = 0.2)

y <- exp(0.3 + 0.5 * x + e)

```

:::::: {.columns}
::::: {.column}

```{r}
#| fig-asp: 0.9

dat <- tibble(
  v = rep(c("Y", "log(Y)"), each = n),
  y = c(y, log(y))
)

txt <- tibble(
  x = c(3.2, 14),
  y = c(35, 6),
  v = c("log(Y)", "Y")
)

ggplot() +
  geom_histogram(
    data = dat,
    position = "identity",
    aes(y, color = v, fill = v),
    bins = 20
  ) +
  scale_color_manual(values = qcolors("blue", "bronze")) +
  scale_fill_manual(values = alpha(qcolors("blue", "bronze"), 0.75)
  ) +
  geom_text(
    data = txt,
    aes(x, y, color = v, label = v),
    size = 8,
    hjust = 0
  ) +
  labs(
    x = "Value",
    y = "Count"
  ) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  )

```

:::::
::::: {.column}

__Logging__ a skewed variable normalizes it.  

This is the inverse of __exponentiation__: 

$$Y = log(exp(Y))$$  

May be applied to $X$ and $Y$.  

__Question__: but, whyyyyyy???

:::::
::::::

## Log transformations for non-linearity

:::::: {.r-stack}
::: {.fragment .fade-out fragment-index=0}

```{r}

linear_mod <- lm(y ~ x)

linear_b0 <- coefficients(linear_mod)[[1]]
linear_b1 <- coefficients(linear_mod)[[2]]

linear_fit <- tibble(
  observation = 1:n,
  estimate = predict(linear_mod),
  residuals = residuals(linear_mod)
)

linear_r2 <- summary(linear_mod)$r.squared

```

__Simple Linear Model:__ $y_{i} = \beta_{0} + \beta_{1}X + \epsilon_{i}$  
$R^2 = `r linear_r2`$

```{r}
#| out-width: 100%
#| layout-ncol: 2

ggplot() +
  geom_abline(
    intercept = linear_b0,
    slope = linear_b1,
    color = qcolors("kobe")
  ) +
  geom_point(
    aes(x, y),
    size = 3
  ) +
  labs(
    x = "X",
    y = "Y"
  ) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )

ggplot(linear_fit, aes(estimate, residuals)) +
  geom_segment(
    aes(xend = estimate, yend = 0),
    linetype = "dashed"
  ) +
  geom_hline(
    aes(yintercept = 0),
    color = qcolors("kobe"),
    size = 1
  ) +
  geom_point(
    size = 3
  ) +
  labs(
    x = "Estimates",
    y = "Residuals"
  ) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )

```

:::
::: {.fragment .fade-in fragment-index=0}

```{r}

log_mod <- lm(log(y) ~ x)

log_b0 <- coefficients(log_mod)[[1]]
log_b1 <- coefficients(log_mod)[[2]]

log_fit <- tibble(
  observation = 1:n,
  estimate = predict(log_mod),
  residuals = residuals(log_mod)
)

log_r2 <- summary(log_mod)$r.squared

```

__Log Linear Model:__ $log(y_{i}) = \beta_{0} + \beta_{1}X + \epsilon_{i}$  
$R^2 = `r log_r2`$

```{r}
#| out-width: 100%
#| layout-ncol: 2

ggplot() +
  geom_abline(
    intercept = log_b0,
    slope = log_b1,
    color = qcolors("kobe")
  ) +
  geom_point(
    aes(x, log(y)),
    size = 3
  ) +
  labs(
    x = "X",
    y = "log(Y)"
  ) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )

ggplot(log_fit, aes(estimate, residuals)) +
  geom_segment(
    aes(xend = estimate, yend = 0),
    linetype = "dashed"
  ) +
  geom_hline(
    aes(yintercept = 0),
    color = qcolors("kobe"),
    size = 1
  ) +
  geom_point(
    size = 3
  ) +
  labs(
    x = "log(Estimates)",
    y = "Residuals"
  ) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )

```

:::
::::::

## âš ï¸ [Careful!]{.kobe}

Have to take care how we interpret log-models!  

$\beta_1$ means...

<br>

:::::: {.columns}
::::: {.column}
__Linear Model__  
$y_{i} = \beta_{0} + \beta_{1}X + \epsilon_{i}$  
_Increase in Y for one unit increase in X._  

<br>

__Log-Linear Model__  
$log(y_{i}) = \beta_{0} + \beta_{1}X + \epsilon_{i}$  
_Percent increase in Y for one unit increase in X._  
:::::
::::: {.column}
__Linear-Log Model__  
$y_{i} = \beta_{0} + \beta_{1}log(X) + \epsilon_{i}$   
_Increase in Y for percent increase in X._  

<br>

__Log-Log Model__  
$log(y_{i}) = \beta_{0} + \beta_{1}log(X) + \epsilon_{i}$  
_Percent increase in Y for percent increase in X._  
:::::
::::::
