---
title: "Quantitative Analysis of Archaeological Data"
subtitle: "Lecture 12: Generalized Linear Mixed Models"  
date: 'Last updated: `r Sys.Date()`'
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [xaringan-themer.css, custom_style.css]
    seal: TRUE
    nature:
      highlightStyle: magula
      highlightLines: TRUE
      countIncrementalSlides: TRUE
      ratio: '16:9'
---

```{r} 
#| include = FALSE, 
#| code = xfun::read_utf8(here::here("slides", "before_chunk.R"))
```

```{r}

elevation <- here("datasets", "elevation.tif") %>% rast()

gpkg <- here("datasets", "qaad.gpkg")

surveys <- read_sf(gpkg, "surveys")
sites <- read_sf(gpkg, "sites")

sites_tbl <- here("datasets", "survey-polygons_sites.csv") %>% read_csv()

```


## Outline

1. A hypothetical survey of sites
2. The problem of natural groupings
2. What to do about them?
2. Generalized Linear Mixed-Effects Models (GLMMs)
2. GLM with dummy variables
2. GLMM method
2. GLMM - Random Intercept
2. GLMM - Random Slope

---
class: center middle

## __Question__ How many residential features at each site?

```{r}
#| out.width = "70%"

ggplot() +
  geom_tile(
    data = as.data.frame(elevation/1000, xy = TRUE),
    aes(x, y, fill = elevation)
  ) +
  scale_fill_viridis(name = "Elevation (km)") +
  geom_sf(
    data = surveys,
    fill = "transparent"
  ) +
  geom_sf(
    data = sites,
    shape = 21,
    color = "black",
    fill = "white", 
    size = 2
  ) +
  coord_sf(expand = FALSE) +
  theme_void(15) +
  theme(
    panel.spacing = margin(),
    plot.margin = margin(),
    legend.position = c(0.98, 0.98),
    legend.justification = c("right", "top")
  )

```

---

## Natural groupings

.pull-left[

Lead to:

1. autocorrelation (non-independence)
2. heteroscedasticity (non-constant variance)

]

.pull-right[

```{r}
#| out.width = "100%"

ggplot(sites_tbl) +
  geom_point(
    aes(elevation, residential, fill = survey),
    shape = 21,
    color = "gray80",
    size = 4
  ) +
  scale_fill_viridis(name = "Survey", discrete = TRUE, alpha = 0.8) +
  labs(
    x = "Elevation (km)",
    y = "Number of Residential Features"
  )

```

]

---

## What to do about natural groupings?

1. Nothing? 
    - Violation of independence assumption
--

2. Aggregate data
    - Loss of information, particularly within-group differences
    - Risk of committing the ecological fallacy
--

3. Dummy variables
    - A coefficient for every level gobbles up degrees of freedom
    - Difficult to generalize
--

4. Mixed effects model
    - Models variance at multiple levels
    - Addresses independence issues
    - Straightforward conceptualization of multilevel data

---

## Generalized linear mixed effect models (GLMMs)

Extend
- GLMs by incorporating random effects
- LMMs by incorporating non-Normal responses

Your pick really...

--

Conceptually similar to a GLM with dummy variables, but it
- treats coefficients themselves as random variables with mean and variance
- doesn't require a reference class (uses grand-mean instead)
- uses similarities between groups to refine estimates
- allows for inference testing on cross-level effects

---

## GLM with a dummy variable

```{r}

features_glm <- glm(
  residential ~ elevation * survey,
  family = poisson,
  data = sites_tbl
)

inverse_link <- family(features_glm)$linkinv

responses <- lapply(
  sites_tbl$survey,
  function(x){
    
    new_data <- tibble(
      elevation = with(sites_tbl, seq(min(elevation), max(elevation), length = 300)),
      survey = x
    )
    
    estimates <- predict(features_glm, newdata = new_data, se.fit = TRUE)
    
    new_data <- new_data %>% 
      mutate(
        residential = inverse_link(estimates$fit),
        ymin = with(estimates, inverse_link(fit - 2*se.fit)),
        ymax = with(estimates, inverse_link(fit + 2*se.fit))
      )
    
    new_data
    
  }
) %>% 
  bind_rows()

```

.pull-left[

```{r}
#| out.height = "100%",
#| fig.asp = 1

facet_labels <- tibble(
  x = 0, y = 45,
  label = LETTERS[1:8],
  group = LETTERS[1:8],
  survey = LETTERS[1:8]
)

gg_responses <- ggplot(responses, aes(elevation, residential, color = survey)) +
  geom_point(
    data = sites_tbl,
    aes(elevation, residential, fill = survey),
    shape = 21,
    color = "gray80",
    size = 3.5
  ) +
  scale_fill_viridis(name = "Survey", discrete = TRUE, alpha = 0.8) +
  geom_line(
    size = 3,
    alpha = 0.35
  ) +
  geom_line(
    size = 1.3
  ) +
  geom_text(
    data = facet_labels,
    aes(x, y, label = label, color = survey),
    size = 8,
    hjust = 0,
    vjust = 1
  ) +
  scale_color_viridis(name = "Survey", discrete = TRUE) +
  labs(
    x = "Elevation (km)",
    y = "Number of Residential Features"
  ) +
  facet_wrap(~survey, nrow = 3, ncol = 3) +
  theme(
    strip.background = element_blank(),
    strip.text = element_blank(),
    plot.margin = margin(6,3,3,3),
    legend.position = "none"
  ) +
  coord_cartesian(ylim = c(0, 45))

gg_responses

```

]

.pull-right[

```{r}

coefficients_table <- tidy(features_glm)

coefficients_table %>% 
  mutate("q" = if_else(p.value > 0.05, "&#10060;", "&#10004;")) %>% 
  kbl(
    format = "html", 
    escape = FALSE,
    col.names = c("", names(.)[2:5], ""),
    align = c("lllllc"),
    table.attr = "class='table-model table-fullwidth'"
  ) %>% 
  kable_paper(
    "striped", 
    position = "left"
  ) %>% 
  scroll_box(height = "5in")

```

]

---

## GLM with a dummy variable

.pull-left[

```{r}
#| out.height = "100%",
#| fig.asp = 1

gg_responses

```

]

.pull-right[

Can, in principle, calculate mean and variance of coefficients across all groups in a post-hoc fashion, using a model's estimates, but it's better to calculate these parameters directly from the data as part of the model fitting process.

]

---

## Two-Level GLMM

__Level One__ Model the dependent variable as a function of your level-one predictors.

$$y_{j} = \beta_{j}X_{j} + \epsilon_{j}\\ \epsilon \sim N(0, \sigma^2)$$

__Level Two__ Model the coefficients as a function of your level-two predictors.

$$\beta_{j} = \gamma_0 + \gamma_{1}W_{j} + u_{j}\\ u \sim N(0, \tau^2)$$

Where

- $W$ is some property of the group (not required)
- $\gamma_0$ is the overall intercept and slope across all groups
- $\gamma_1$ is the variation of the intercept and slope between groups
- $u$ is the error around the intercept and slope for each group

---

## GLMM - Random Intercept

```{r}

features_glmm <- lme4::glmer(
  residential ~ elevation + (1|survey),
  data = sites_tbl,
  family = poisson
)

responses <- ggpredict(
  features_glmm, 
  terms = c("elevation", "survey"), 
  type = 'random'
)

```

.pull-left[

```{r}
#| out.height = "100%",
#| fig.asp = 1

ggplot(responses, aes(x = x, y = predicted, color = group)) +
  geom_point(
    data = sites_tbl %>% rename("group" = survey),
    aes(elevation, residential, fill = group),
    shape = 21,
    color = "gray80",
    size = 3.5
  ) +
  scale_fill_viridis(name = "Survey", discrete = TRUE, alpha = 0.8) +
  geom_line(
    size = 3,
    alpha = 0.35
  ) +
  geom_line(
    aes(color = group),
    size = 1.3
  ) +
  geom_text(
    data = facet_labels,
    aes(x, y, label = label, color = group),
    size = 8,
    hjust = 0,
    vjust = 1
  ) +
  scale_color_viridis(name = "Survey", discrete = TRUE) +
  labs(
    x = "Elevation (km)",
    y = "Number of Residential Features"
  ) +
  facet_wrap(~group, nrow = 3, ncol = 3) +
  theme(
    strip.background = element_blank(),
    strip.text = element_blank(),
    plot.margin = margin(6,3,3,3),
    legend.position = "none"
  ) +
  coord_cartesian(ylim = c(0, 45))

```

]

.pull-right[

```{r}

coefficients_table <- tidy(features_glmm) %>% 
  select(-effect) %>% 
  mutate(
    term = c("(Intercept)", "elevation", "(Intercept) &#x3C3;<sup>2</sup>")
  )

coefficients_table %>% 
  mutate("q" = if_else(p.value > 0.05, "&#10060;", "&#10004;")) %>% 
  kbl(
    format = "html", 
    escape = FALSE,
    col.names = c("", "", names(.)[3:6], ""),
    align = c("lllllc"),
    table.attr = "class='table-model table-fullwidth'"
  ) %>% 
  kable_paper(
    "striped", 
    position = "left"
  ) %>% 
  pack_rows("Fixed Effects",  1, 2, indent = FALSE) %>% 
  pack_rows("Random Effects", 3, 3, indent = FALSE) %>% 
  collapse_rows(1)

```

]

---

## GLMM - Random Slopes

```{r}

features_glmm_slope <- lme4::glmer(
  residential ~ elevation + (elevation|survey),
  data = sites_tbl,
  family = poisson
)

responses <- ggpredict(
  features_glmm_slope, 
  terms = c("elevation", "survey"), 
  type = 'random'
)

```

.pull-left[

```{r}
#| out.height = "100%",
#| fig.asp = 1

ggplot(responses, aes(x = x, y = predicted, color = group)) +
  geom_point(
    data = sites_tbl %>% rename("group" = survey),
    aes(elevation, residential, fill = group),
    shape = 21,
    color = "gray80",
    size = 3.5
  ) +
  scale_fill_viridis(name = "Survey", discrete = TRUE, alpha = 0.8) +
  geom_line(
    size = 3,
    alpha = 0.35
  ) +
  geom_line(
    aes(color = group),
    size = 1.3
  ) +
  geom_text(
    data = facet_labels,
    aes(x, y, label = label, color = group),
    size = 8,
    hjust = 0,
    vjust = 1
  ) +
  scale_color_viridis(name = "Survey", discrete = TRUE) +
  labs(
    x = "Elevation (km)",
    y = "Number of Residential Features"
  ) +
  facet_wrap(~group, nrow = 3, ncol = 3) +
  theme(
    strip.background = element_blank(),
    strip.text = element_blank(),
    plot.margin = margin(6,3,3,3),
    legend.position = "none"
  ) +
  coord_cartesian(ylim = c(0, 45))

```

]

.pull-right[

```{r}

coefficients_table <- tidy(features_glmm_slope) %>% 
  select(-effect) %>% 
  mutate(
    term = c("(Intercept)", "elevation", "(Intercept) &#x3C3;<sup>2</sup>", 
             "ICC*", "elevation &#x3C4;<sup>2</sup>")
  )

coefficients_table %>% 
  mutate("q" = if_else(p.value > 0.05, "&#10060;", "&#10004;")) %>% 
  kbl(
    format = "html", 
    escape = FALSE,
    col.names = c("", "", names(.)[3:6], ""),
    align = c("lllllc"),
    table.attr = "class='table-model table-fullwidth'"
  ) %>% 
  kable_paper(
    "striped", 
    position = "left"
  ) %>% 
  pack_rows("Fixed Effects",  1, 2, indent = FALSE) %>% 
  pack_rows("Random Effects", 3, 5, indent = FALSE) %>% 
  collapse_rows(1)

```

.grey[_ICC = `\\(\tau^2/(\tau^2 + \sigma^2)\\)`. If ICC = 0, then group effect is negligible._]

]

---

## &#x1F52D; Looking Ahead

```{r}

schedule <- here("_misc", "course_outline.xlsx") %>% 
  readxl::read_excel(1) %>% 
  mutate(
    date = lubridate::ymd(date),
    description = glue::glue(
    "<b>{topic}</b> <span style = 'font-size: 80%; color: #9C9C9C;'>({date})</span><br>
    <div style = 'font-size: 80%; font-style: italic; color: #9C9C9C;'>{details}</div>"
  )) %>% 
  select(description)

```

.pull-left[

```{r}

schedule %>% 
  slice(13:16) %>% 
  kbl(escape = FALSE, col.names = NULL) %>% 
  kable_paper(c("striped", "hover"))

```

]
