---
title: "Lecture 12: Logistic Regression"  
date: 'Last updated: `r Sys.Date()`'
---

```{r setup}
#| include: false

here::here("slides", "_setup.R") |> source()

model_table_defaults <- function(x){

  tab_options(
    x,
    column_labels.font.weight = "bold",
    container.width = pct(100),
    table.border.top.style = "hidden",
    table.font.size = px(20),
    table.width = pct(100)
  )
  
}

library(archdata)

data("Snodgrass")

snodgrass <- Snodgrass |> 
  as_tibble() |> 
  rename_with(tolower) |> 
  select(inside, area) |> 
  mutate(inside = ifelse(inside == "Inside", 1, 0))

data("DartPoints")

darts <- DartPoints |> 
  as_tibble() |> 
  rename_with(tolower) |> 
  select(name, length, width, thickness, weight) |> 
  filter(weight < 20)

data("OxfordPots")

oxford <- OxfordPots |> 
  as_tibble() |> 
  rename(
    "p" = OxfordPct,
    "distance" = OxfordDst
  ) |> 
  select(p, distance) |> 
  mutate(p = p/100)

remove(Snodgrass, DartPoints, OxfordPots)

```

## 📋 Lecture Outline

- General Linear Model
- Generalized Linear Model
    - Distribution function
    - Linear predictor
    - Link function
- Gaussian outcomes
- Bernoulli outcomes
- Proportional outcomes
- Deviance
- Information Criteria
- ANOVE _aka_ Likelihood Ratio Test

## General Linear Models

$$y = E(Y) + \epsilon$$
$$E(Y) = \mu = \beta X$$

- $E(Y)$ is the expected value (equal to the conditional mean, $\mu$)  
- $\beta X$ is the linear predictor  
- Error is normal, $\epsilon \sim N(0, \sigma^2)$

## Generalized Linear Models

$$y = E(Y) + \epsilon$$
$$E(Y) = g(\mu) = \beta X$$

- $g$ is the link function, makes the relationship linear 
- $\beta X$ is the linear predictor
- Error is exponential, $\epsilon \sim Exponential$

## Generalized Linear Models

```{css}

@media (min-width: 30em) {

  .panel-tabset {
  	display: grid;
  	grid-gap: 1em;
  	grid-template-columns: 25% 75%;
  }
  
  .panel-tabset-tabby {
    border-bottom: none !important;
  	border-right: 1px solid #bbb; 
  }
  
  .panel-tabset [role=tab][aria-selected=true] {
    background-color: transparent;
    border: 1px solid #bbb !important;
  }
  
  h3 {
  	margin-top: 0;
  }
  
}

```

:::::: {.panel-tabset}

### Exponential

- A family of distributions: 
    - **Normal** - continuous and unbounded
    - **Gamma** - continuous and non-negative
    - **Binomial** - binary (yes/no) or proportional (0-1) 
    - **Poisson** - count data
- Describes the distribution of the response
- Two parameters: mean and variance
- Variance is a function of the mean:

$$Var(\epsilon) = \phi \mu$$

where $\phi$ is a scaling parameter, assumed to be equal to 1, meaning the variance is assumed to be equal to the mean.

### Linear predictor

- Incorporates information about independent variables  
- Combination of $x$ variables and associated coefficients  

$$\beta X = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \ldots + \beta_n x_n$$

- Commonly denoted with Greek letter "eta", as in $\eta = \beta X$

### Link function

- $g()$ modifies relationship between predictors and expected value.  
- Makes this relationship linear

```{css}

.bob table td {
  height: 70px;
  vertical-align: middle;
}

```

::: {.bob}

| Distribution | Name     | Link                                            | Mean                                 |
|--------------|----------|-------------------------------------------------|--------------------------------------|
| Normal       | Identity | $\beta X = \mu$                                 | $\mu = \beta X$                      |
| Gamma        | Inverse  | $\beta X = \mu^{-1}$                            | $\mu = (\beta X)^{-1}$               |
| Poisson      | Log      | $\beta X = ln\,(\mu)$                           | $\mu = exp\, (\beta X)$              |
| Binomial     | Logit    | $\beta X = ln\, \left(\frac{\mu}{1-\mu}\right)$ | $\mu = \frac{1}{1+exp\, (-\beta X)}$ |

:::

::::::

## Gaussian response

```{r}

log_likelihood <- function(e){
  
  n <- length(e)
  
  sigma <- sqrt(sum(e^2)/n)
  
  sum(dnorm(e, mean = 0, sd = sigma, log = TRUE))

}

darts_glm <- glm(length ~ weight, data = darts)

b0 <- coefficients(darts_glm)[["(Intercept)"]]
b1 <- coefficients(darts_glm)[["weight"]]

```

::::::::: {.r-stack .flt}
:::::: {.columns .fragment .fade-out fragment-index=0 .w100}
::::: {.column}

```{r}

darts_gg <- ggplot(darts, aes(weight, length)) +
  labs(
    x = "Weight (g)",
    y = "Length (mm)"
  )

darts_gg + geom_point(size = 3, alpha = 0.6)

```

:::::
::::: {.column}

Assume length is Gaussian with

$Var(\epsilon) = \sigma^2$  
$E(Y) = \mu = \beta X$  

__Question__ What is the probability that we observe these data given a model with parameters $\beta$ and $\sigma^2$? 

:::::
::::::

:::::: {.columns .fragment .fade-in-then-out fragment-index=0 .w100}
::::: {.column}

```{r}

b0_hat <- round(b0 - 2, digits = 2)
b1_hat <- round(b1 - 1, digits = 2)

yhat <- b0_hat + b1_hat * darts$weight

errors <- with(darts, length - yhat)

sigma <- sqrt(sum(errors^2)/length(errors))

ll <- round(log_likelihood(errors), digits = 2)

darts_gg +
  geom_segment(
    data = tibble(
      x = darts$weight,
      y = darts$length,
      xend = x,
      yend = yhat
    ),
    aes(x, y, xend = xend, yend = yend),
    color = "gray60",
    linetype = "dashed"
  ) +
  geom_abline(
    intercept = b0_hat,
    slope = b1_hat
  ) +
  geom_point(size = 3, alpha = 0.6) +
  geom_text(
    data = tibble(
      x = min(darts$weight), 
      y = with(darts, c(max(length), max(length) - 5)), 
      label = c(
        paste0("\U03B2", "0 = ", b0_hat),
        paste0("\U03B2", "1 = ", b1_hat)
      )
    ),
    aes(x, y, label = label),
    size = 12/.pt,
    hjust = 0,
    vjust = 1
  )

```

:::::
::::: {.column}

```{r}

ggplot() + 
  geom_histogram(
    data = tibble(x = errors),
    aes(x, y = after_stat(density)),
    bins = 15,
    color = "black",
    fill = "gray65"
  ) +
  geom_polygon(
    data = bind_rows(
      c(x = -10, y = 0),
      tibble(
        x = seq(-10, 45, length = 300), 
        y = dnorm(x, sd = sigma)
      )
    ),
    aes(x, y),
    color = qcolors("sapphire"),
    fill = alpha(qcolors("sapphire"), 0.5),
    linewidth = 1
  ) +
  annotate(
    "segment",
    x = 0,
    y = 0,
    xend = 0,
    yend = dnorm(1, 0, sigma),
    color = qcolors("kobe"),
    linewidth = 1.5
  ) +
  labs(
    x = "Residuals",
    y = "Density"
  ) +
  coord_cartesian(
    xlim = c(-10, 45)
  ) +
  annotate(
    "text",
    label = paste0("logLik", " = ", ll),
    x = Inf,
    y = Inf,
    hjust = 1.1,
    vjust = 2,
    size = 12/.pt
  ) +
  scale_y_continuous(
    labels = scales::label_number(accuracy = 0.01)
  )

```

:::::
::::::

:::::: {.columns .fragment .fade-in-then-out fragment-index=1 .w100}
::::: {.column}

```{r}

b0_hat <- round(b0 - 1, digits = 2)
b1_hat <- round(b1 - 0.5, digits = 2)

yhat <- b0_hat + b1_hat * darts$weight

errors <- with(darts, length - yhat)

sigma <- sqrt(sum(errors^2)/length(errors))

ll <- round(log_likelihood(errors), digits = 2)

darts_gg +
  geom_segment(
    data = tibble(
      x = darts$weight,
      y = darts$length,
      xend = x,
      yend = yhat
    ),
    aes(x, y, xend = xend, yend = yend),
    color = "gray60",
    linetype = "dashed"
  ) +
  geom_abline(
    intercept = b0_hat,
    slope = b1_hat
  ) +
  geom_point(size = 3, alpha = 0.6) +
  geom_text(
    data = tibble(
      x = min(darts$weight), 
      y = with(darts, c(max(length), max(length) - 5)), 
      label = c(
        paste0("\U03B2", "0 = ", b0_hat),
        paste0("\U03B2", "1 = ", b1_hat)
      )
    ),
    aes(x, y, label = label),
    size = 12/.pt,
    hjust = 0,
    vjust = 1
  )

```

:::::
::::: {.column}

```{r}

ggplot() + 
  geom_histogram(
    data = tibble(x = errors),
    aes(x, y = after_stat(density)),
    bins = 15,
    color = "black",
    fill = "gray65"
  ) +
  geom_polygon(
    data = bind_rows(
      c(x = -10, y = 0),
      tibble(
        x = seq(-10, 45, length = 300), 
        y = dnorm(x, sd = sigma)
      )
    ),
    aes(x, y),
    color = qcolors("sapphire"),
    fill = alpha(qcolors("sapphire"), 0.5),
    linewidth = 1
  ) +
  annotate(
    "segment",
    x = 0,
    y = 0,
    xend = 0,
    yend = dnorm(1, 0, sigma),
    color = qcolors("kobe"),
    linewidth = 1.5
  ) +
  labs(
    x = "Residuals",
    y = "Density"
  ) +
  coord_cartesian(
    xlim = c(-10, 45)
  ) +
  annotate(
    "text",
    label = paste0("logLik", " = ", ll),
    x = Inf,
    y = Inf,
    hjust = 1.1,
    vjust = 2,
    size = 12/.pt
  ) +
  scale_y_continuous(
    labels = scales::label_number(accuracy = 0.01)
  )

```

:::::
::::::

:::::: {.columns .fragment .fade-in fragment-index=2 .w100}
::::: {.column}

```{r}

b0_hat <- round(b0 - 0, digits = 2)
b1_hat <- round(b1 - 0, digits = 2)

yhat <- b0_hat + b1_hat * darts$weight

errors <- with(darts, length - yhat)

sigma <- sqrt(sum(errors^2)/length(errors))

ll <- round(log_likelihood(errors), digits = 2)

darts_gg +
  geom_segment(
    data = tibble(
      x = darts$weight,
      y = darts$length,
      xend = x,
      yend = yhat
    ),
    aes(x, y, xend = xend, yend = yend),
    color = "gray60",
    linetype = "dashed"
  ) +
  geom_abline(
    intercept = b0_hat,
    slope = b1_hat
  ) +
  geom_point(size = 3, alpha = 0.6) +
  geom_text(
    data = tibble(
      x = min(darts$weight), 
      y = with(darts, c(max(length), max(length) - 5)), 
      label = c(
        paste0("\U03B2", "0 = ", b0_hat),
        paste0("\U03B2", "1 = ", b1_hat)
      )
    ),
    aes(x, y, label = label),
    size = 12/.pt,
    hjust = 0,
    vjust = 1
  )

```

:::::
::::: {.column}

```{r}

ggplot() + 
  geom_histogram(
    data = tibble(x = errors),
    aes(x, y = after_stat(density)),
    bins = 15,
    color = "black",
    fill = "gray65"
  ) +
  geom_polygon(
    data = bind_rows(
      c(x = -10, y = 0),
      tibble(
        x = seq(-10, 45, length = 300), 
        y = dnorm(x, sd = sigma)
      )
    ),
    aes(x, y),
    color = qcolors("sapphire"),
    fill = alpha(qcolors("sapphire"), 0.5),
    linewidth = 1
  ) +
  annotate(
    "segment",
    x = 0,
    y = 0,
    xend = 0,
    yend = dnorm(1, 0, sigma),
    color = qcolors("kobe"),
    linewidth = 1.5
  ) +
  labs(
    x = "Residuals",
    y = "Density"
  ) +
  coord_cartesian(
    xlim = c(-10, 45)
  ) +
  annotate(
    "text",
    label = paste0("logLik", " = ", ll),
    x = Inf,
    y = Inf,
    hjust = 1.1,
    vjust = 2,
    size = 12/.pt
  ) +
  scale_y_continuous(
    labels = scales::label_number(accuracy = 0.01)
  )

```

:::::
::::::

:::::::::

::: aside
_Archaic dart points. Sample from Fort Hood, Texas_
:::

## Log Odds

::: {.bob style="width:65%;"}

Location of residential features at the Snodgrass sites

|             | Inside wall                       | Outside wall                      | Total |
|-------------|-----------------------------------|-----------------------------------|-------|
| Count       | 38                                | 53                                | 91    |
| Probability | &nbsp; 0.42 = $\frac{38}{91}$     | &nbsp; 0.58 = $\frac{53}{91}$     | 1     |
| Odds        | &nbsp; 0.72 = $\frac{0.42}{0.58}$ | &nbsp; 1.40 = $\frac{0.58}{0.42}$ |       |
| Log Odds    | -0.33 = log(0.72)                 | &nbsp; 0.33 = log(1.40)           |       |

:::

**Why Log Odds?**  
Because the distribution of odds can be highly skewed, and taking the log normalizes it (makes it more symmetric).    

## Bernoulli response

::::::::: {.r-stack .flt}
:::::: {.columns .fragment .fade-out fragment-index=0 .w100}
::::: {.column}

```{r}

snodgrass_gg <- ggplot(snodgrass, aes(area, inside)) +
  labs(
    x = "Area of House Structure (sq ft)",
    y = "Inside Inner Wall"
  ) +
  scale_y_continuous(
    breaks = c(0, 0.5, 1),
    labels = c(0, 0.5, 1)
  )

snodgrass_gg + geom_point(size = 3, alpha = 0.4)

```

:::::
::::: {.column}

Location inside or outside of the inner wall at the Snodgrass site is a Bernoulli variable and has expectation $E(Y) = p$ where

$$p = \frac{1}{1 + exp(-\beta X)}$$
This defines a [logistic curve](https://en.wikipedia.org/wiki/Logistic_function) or sigmoid, with $p$ being the probability of success. This constrains the estimate $E(Y)$ to be in the range 0 to 1.

:::::
::::::

:::::: {.columns .fragment .fade-in-then-out fragment-index=0 .w100}
::::: {.column}

```{r}

snodgrass_gg + geom_point(size = 3, alpha = 0.4)

```

:::::
::::: {.column}

Taking the log of $p$ gives us

$$log(p) = log\left(\frac{p}{1 - p}\right) = \beta X$$

This is known as the "logit" or [log odds](https://en.wikipedia.org/wiki/Logit). 

__Question__ What is the probability that we observe these data (these inside features) given a model with parameters $\beta$? 

:::::
::::::

:::::: {.columns .fragment .fade-in fragment-index=1 .w100}
::::: {.column}

```{r}

snodgrass_glm <- glm(
  inside ~ area,
  family = binomial(link = "logit"),
  data = snodgrass
)

b0 <- coefficients(snodgrass_glm)[["(Intercept)"]]
b1 <- coefficients(snodgrass_glm)[["area"]]

ll <- logLik(snodgrass_glm)

snodgrass_gg + 
  geom_point(size = 3, alpha = 0.4) +
  geom_line(
    data = tibble(
      x = snodgrass$area,
      y = predict(snodgrass_glm, type = "response")
    ),
    aes(x, y),
    color = qcolors("kobe"),
    linewidth = 1
  )

```

:::::
::::: {.column}

Estimated coefficients: 

$\beta_0 = `r b0`$  
$\beta_1 = `r b1`$   

For these, the log Likelihood is

$\mathcal{l} = `r ll`$  

:::::
::::::
:::::::::

::: aside
*Data from the Snodgrass site in Butler County, MS*
:::

## Interpretation

::::::::: {.r-stack .flt}
:::::: {.columns .fragment .fade-out fragment-index=0 .w100}
::::: {.column}

```{r}

snodgrass_gg + 
  geom_point(size = 3, alpha = 0.4) +
  geom_line(
    data = tibble(
      x = snodgrass$area,
      y = predict(snodgrass_glm, type = "response")
    ),
    aes(x, y),
    color = qcolors("kobe"),
    linewidth = 1
  )

```

:::::
::::: {.column}

Estimated coefficients: 

$\beta_0 = `r b0`$  
$\beta_1 = `r b1`$   

Note that these coefficient estimates are log-odds! To get the odds, we take the exponent.   

$\beta_0 = exp(`r b0`) = `r exp(b0)`$  
$\beta_1 = exp(`r b1`) = `r exp(b1)`$  

For a one unit increase in area, the odds of being in the inside wall increase by `r exp(b1)`.

:::::
::::::

:::::: {.columns .fragment .fade-in fragment-index=0 .w100}
::::: {.column}

```{r}

est <- predict(
  snodgrass_glm, 
  newdata = tibble(area = 300), 
  type = "response"
)

snodgrass_gg + 
  geom_point(size = 3, alpha = 0.4) +
  geom_line(
    data = tibble(
      x = snodgrass$area,
      y = predict(snodgrass_glm, type = "response")
    ),
    aes(x, y),
    color = qcolors("kobe"),
    size = 1
  ) +
  annotate(
    "segment",
    x = 300,
    xend = 300,
    y = 0,
    yend = est,
    linetype = "dashed",
    color = "gray50"
  ) +
  annotate(
    "segment",
    x = min(snodgrass$area),
    xend = 300,
    y = est,
    yend = est,
    linetype = "dashed",
    color = "gray50"
  ) +
  annotate(
    "point", 
    x = 300,
    y = est,
    size = 5
  )

```

:::::
::::: {.column}

Estimated coefficients: 

$\beta_0 = `r b0`$  
$\beta_1 = `r b1`$   

To get the probability, we can use the mean function (also known as the inverse link):

$$p = \frac{1}{1+exp(-\beta X)}$$

For a house structure with an area of 300 square feet, the estimated probability that it occurs inside the inner wall is `r est`.

:::::
::::::
:::::::::

::: aside
*Data from the Snodgrass site in Butler County, MS*
:::

## Proportional response

::::::::: {.r-stack .flt}
:::::: {.columns .fragment .fade-out fragment-index=0 .w100}
::::: {.column}

```{r}

oxford_gg <- ggplot(oxford, aes(distance, p)) +
  labs(
    x = "Distance from Oxford (miles)",
    y = "Proportion of Roman Pots"
  ) +
  scale_y_continuous(
    limits = c(0, 0.23),
    breaks = c(0, 0.1, 0.2),
    labels = c(0, 0.1, 0.2)
  )

oxford_gg + geom_point(size = 3, alpha = 0.4)

```

:::::
::::: {.column}

Proportion of Roman pottery is a binomial variable and has expectation $E(Y) = p$ where

$$p = \frac{1}{1 + exp(-\beta X)}$$
This defines a [logistic curve](https://en.wikipedia.org/wiki/Logistic_function) or sigmoid, with $p$ being the proportion of successful Bernoulli trials. This constrains the estimate $E(Y)$ to be in the range 0 to 1.

:::::
::::::

:::::: {.columns .fragment .fade-in-then-out fragment-index=0 .w100}
::::: {.column}

```{r}

oxford_gg + geom_point(size = 3, alpha = 0.4)

```

:::::
::::: {.column}

Taking the log of $p$ gives us

$$log(p) = log\left(\frac{p}{1 - p}\right) = \beta X$$

This is known as the "logit" or [log odds](https://en.wikipedia.org/wiki/Logit). 

__Question__ What is the probability that we observe these data (these proportions) given a model with parameters $\beta$? 

:::::
::::::

:::::: {.columns .fragment .fade-in fragment-index=1 .w100}
::::: {.column}

```{r}

oxford_glm <- glm(
  p ~ distance,
  family = binomial(link = "logit"),
  data = oxford
)

b0 <- coefficients(oxford_glm)[["(Intercept)"]]
b1 <- coefficients(oxford_glm)[["distance"]]

ll <- logLik(oxford_glm)

oxford_gg + 
  geom_point(size = 3, alpha = 0.4) +
  geom_line(
    data = tibble(
      x = oxford$distance,
      y = predict(oxford_glm, type = "response")
    ),
    aes(x, y),
    color = qcolors("kobe"),
    linewidth = 1
  )

```

:::::
::::: {.column}

Estimated coefficients: 

$\beta_0 = `r b0`$  
$\beta_1 = `r b1`$   

For these, the log Likelihood is

$\mathcal{l} = `r ll`$  

:::::
::::::
:::::::::

::: aside
*Data from Fulford and Hodder (1974). A Regression Analysis of Some Late Romano-British Pottery: A
Case Study*
:::

## Deviance

- Measure of goodness of fit, so __smaller is better__
- Gives the difference in log-Likelihood between a model $M$ and a __saturated__ model $M_S$
    - $M_S$ has a parameter for each observation (i.e., zero degrees of freedom)

$$D = -2\,\Big[\,log\,\mathcal{L}(M_1) - log\,\mathcal{L}(M_S)\,\Big]$$

- __Residual deviance__ = deviance of proposed model
- __Null deviance__ = deviance of null model (ie, intercept-only model)

## Information Criteria

- A function of the deviance, so __smaller is better__
- Two penalties: 
    - $n$ = number of observations 
    - $p$ = number of parameters
- Akaike Information Criteria
    - $AIC = D + 2p$
- Bayesian Information Criteria 
    - $BIC = D + (p \cdot log(n))$

## ANOVA

- Analysis of Deviance _aka_ Likelihood Ratio Test
- For Binomial and Poisson models
- Test statistic is the logged ratio of likelihoods between proposed, $M_1$, and null, $M_0$, models

$$
\begin{aligned}
\chi^2 &= -2\,log\,\frac{\mathcal{L}(M_0)}{\mathcal{L}(M_1)}\\\\
&= -2\,\Big[\,log\,\mathcal{L}(M_0) - log\,\mathcal{L}(M_1)\,\Big]
\end{aligned}
$$

- Compare to a $\chi^2$ distribution with $k$ degrees of freedom - asks what the probability is of getting a value greater than the observed $\chi^2$ value.
- Null hypothesis is no difference in log-Likelihood.