---
title: "Quantitative Analysis of Archaeological Data"
subtitle: "Lecture 13: Generalized Additive Models"  
date: 'Last updated: `r Sys.Date()`'
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [xaringan-themer.css, custom_style.css]
    seal: TRUE
    nature:
      highlightStyle: magula
      highlightLines: TRUE
      countIncrementalSlides: TRUE
      ratio: '16:9'
---

```{r} 
#| include = FALSE, 
#| code = xfun::read_utf8(here::here("slides", "before_chunk.R"))
```

```{r}

patch <- here("datasets", "patch-residence.csv") %>% 
  read_csv() %>% 
  mutate(
    sex = factor(sex)
  )

```


## Outline

1. A Goldilocks distance for central-place foragers
    - A simple linear model?
    - Polynomial regression?
2. A general strategy for non-linear data
    - Model responses of Y across _regions_ of X
    - Three challenges: (1) How to define regions? (2) How to make models continuous? and<br>(3) How to avoid over-fitting?
2. Basis functions
    - Piecewise constant (step function) and piecewise polynomial
    - Natural spline (with truncated power)
    - Smoothing spline (with roughness penalty)  
2. Generalized Additive Models
    - GAM = GLM + Splines
    - Effective Degrees of Freedom

---

## A Goldilocks Distance?

.pull-left[

```{r}

sigmoid <- function(x, x0) 1 / (1 + exp(-(x-x0)))

assumptions <- tibble(
  distance = seq(0, 10, length = 100)
) %>% 
  mutate(
    "Within-Group Competition" = 1 - sigmoid(distance, 4),
    "Predation Threat" = sigmoid(distance, 6),
    "Patch Quality" = 0.75
  ) %>% 
  pivot_longer(
    cols = -distance, 
    names_to = "variable",
    values_to = "value"
  ) %>% 
  select(variable, distance, value) %>% 
  arrange(variable, distance) %>% 
  mutate(
    variable = factor(variable),
    variable = fct_relevel(variable, "Within-Group Competition", "Predation Threat")
  )

mylabels <- tibble(
  x = c(0, 10, 10),
  y = c(0.05, 0.72, 0.05),
  label = c("Predation\nThreat", "Patch Quality", "Within-Group\nCompetition"),
  color = c("Predation Threat", "Patch Quality", "Within-Group Competition")
)

ggplot() +
  geom_line(
    data = assumptions,
    aes(distance, value, color = variable),
    size = 2
  ) +
  geom_text(
    data = mylabels,
    aes(x, y, label = label, color = color),
    hjust = c(0, 1, 1),
    vjust = c(0, 1, 0),
    size = 7,
    lineheight = 0.85
  ) +
  scale_color_viridis(
    end = 0.97,
    discrete = TRUE
  ) +
  labs(
    x = "Distance from Central-Place",
    y = NULL
  ) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  )

remove(sigmoid, assumptions, mylabels)

```

]

.pull-right[

Assumptions

- Probability of predation increases with distance from a central place.
- Within-group competition decreases with distance from a central place.
- Patch quality is constant.

Prediction

- Foragers will spend relatively more time in patches at some intermediate distance from a central place (and relatively less time in patches closer and farther away).

]

---

## A Goldilocks Distance?

.pull-left[

```{r}

ggplot(patch) +
  geom_point(
    aes(distance, time),
    size = 4,
    color = "gray25", 
    alpha = 0.75
  ) +
  labs(
    x = "Distance from central place (km)",
    y = "Time in Patch (hours)"
  )

```

]

.pull-right[

__Question__ How do we model this?

]


---
count: false

## A Goldilocks Distance?

.pull-left[

```{r}

fit <- lm(time ~ distance, data = patch)
b0 <- coefficients(fit)[["(Intercept)"]]
b1 <- coefficients(fit)[["distance"]]

ggplot(patch) +
  geom_point(
    aes(distance, time),
    size = 4,
    color = "gray25", 
    alpha = 0.75
  ) +
  geom_abline(
    intercept = b0,
    slope = b1,
    color = qaad_colors("rufous_red"),
    size = 1.5
  ) +
  labs(
    x = "Distance from central place (km)",
    y = "Time in Patch (hours)"
  ) 

remove(fit, b0, b1)

```

]

.pull-right[

__Question__ How do we model this?

Can try a linear-model, but we get... 

- auto-correlation,
- heteroscedasticity, and
- the slope is not-significant!

]





---

## A Goldilocks Distance?

.pull-left[

```{r}

fit <- lm(time ~ poly(distance, 2), data = patch)

estimate <- tibble(
  distance = patch %>% pull(distance),
  time = predict(fit)
)

ggplot(patch) +
  geom_point(
    aes(distance, time),
    size = 4,
    color = "gray25", 
    alpha = 0.75
  ) +
  geom_line(
    data = estimate,
    aes(distance, time),
    color = qaad_colors("rufous_red"),
    size = 1.5
  ) +
  labs(
    x = "Distance from central place (km)",
    y = "Time in Patch (hours)"
  ) 

```

]

.pull-right[

__Question__ How do we model this?

Could try a linear model with polynomials...

$$Y = \beta X^d$$

...but this lacks generality.

]

---

## A general strategy for non-linear data

.pull-left[

```{r}

ggplot(patch) +
  geom_point(
    aes(distance, time),
    size = 4,
    color = "gray25", 
    alpha = 0.75
  ) +
  geom_vline(
    xintercept = c(2.5, 5, 7.5),
    linetype = "dashed"
  ) +
  labs(
    x = "Distance from central place (km)",
    y = "Time in Patch (hours)"
  ) 

```

]

.pull-right[

Divide X into regions using __knots__ (dashed lines).

.pull-left[

.center[**Regions**]

$$
\begin{aligned}
&\; X < 2.5 \\  
2.5 \leq &\; X < 5.0 \\ 
5.0 \leq &\; X < 7.5 \\
7.5 \leq &\; X
\end{aligned}
$$  
]

]

---

## A general strategy for non-linear data

.pull-left[

```{r}

patch_cuts <- patch %>% 
  mutate(
    cuts = cut(distance, breaks = c(-Inf, 2.5, 5, 7.5, Inf))
  ) %>% 
  nest_by(cuts) %>% 
  mutate(
    model = list(lm(time ~ distance, data = data)),
    estimate = list(unname(predict(model)))
  ) %>% 
  select(-model) %>% 
  unnest(everything())

ggplot(patch) +
  geom_point(
    aes(distance, time),
    size = 4,
    color = "gray25", 
    alpha = 0.75
  ) +
  geom_vline(
    xintercept = c(2.5, 5, 7.5),
    linetype = "dashed"
  ) +
  geom_line(
    data = patch_cuts,
    aes(distance, estimate, group = cuts),
    size = 1.5,
    color = qaad_colors("rufous_red")
  ) +
  labs(
    x = "Distance from central place (km)",
    y = "Time in Patch (hours)"
  ) 

```

]

.pull-right[

Divide X into regions using __knots__ (dashed lines).

Model each region separately.

]

---

## A general strategy for non-linear data

.pull-left[

```{r}

ggplot(patch) +
  geom_point(
    aes(distance, time),
    size = 4,
    color = "gray25", 
    alpha = 0.75
  ) +
  geom_vline(
    xintercept = c(2.5, 5, 7.5),
    linetype = "dashed"
  ) +
  geom_line(
    data = patch_cuts,
    aes(distance, estimate, group = cuts),
    size = 1.5,
    color = qaad_colors("rufous_red")
  ) +
  labs(
    x = "Distance from central place (km)",
    y = "Time in Patch (hours)"
  ) 

```

]

.pull-right[

Divide X into regions using __knots__ (dashed lines).

Model each region separately.


Challenges: 

1. How to define regions (knots)?
2. How to make models continuous?
2. How to avoid over-fitting?

]

---

## Basis Functions

- Simple formula for a linear model.

$$y_{i} = E(y_{i}) + \epsilon_{i}$$
$$E(y_{i}) = \beta X_{i}$$

- A basis function is any function applied to the set of predictor variables $X$.

$$E(y_{i}) = \beta\, f(X_{i})$$

- Examples include polynomials, log-transforms, and exponentiation
- As these are fixed and known, they can still be combined in an ordinary linear model.  

---

## Piecewise Functions

.pull-left[

```{r}

mylabels <- tibble(
  x = seq(0, 7.5, by = 2.5) + 0.3,
  y = 5.9,
  label = LETTERS[1:4]
)

ggplot(patch) +
  geom_point(
    aes(distance, time),
    size = 4,
    color = "gray25", 
    alpha = 0.75
  ) +
  geom_vline(
    xintercept = c(2.5, 5, 7.5),
    linetype = "dashed"
  ) +
  geom_text(
    data = mylabels,
    aes(x, y, label = label),
    size = 10,
    hjust = 0,
    vjust = 1,
    color = qaad_colors("rufous_red")
  ) +
  labs(
    x = "Distance from central place (km)",
    y = "Time in Patch (hours)"
  ) 

```

]

.pull-right[

**Piecewise functions** are basis functions that use dummy variables to group data into regions of $X$ as defined by knots, $k$ (dashed lines), with $X < k_1$ being the reference class.  

Crucially, this reduces the degrees of freedom by one less than the number of $k$!  

]

---

## Piecewise Constant Functions

.pull-left[

```{r}

fit <- lm(
  time ~ cut(distance, breaks = c(-Inf, 2.5, 5, 7.5, Inf)),
  data = patch
)

patch_cuts <- tibble(
  distance = seq(0, 10, length = 200)
) %>% 
  mutate(
    estimate = predict(fit, newdata = tibble(distance))
  )

ggplot(patch) +
  geom_point(
    aes(distance, time),
    size = 4,
    color = "gray25", 
    alpha = 0.75
  ) +
  geom_vline(
    xintercept = c(2.5, 5, 7.5),
    linetype = "dashed"
  ) +
  geom_line(
    data = patch_cuts,
    aes(distance, estimate),
    size = 1.5,
    color = qaad_colors("rufous_red")
  ) +
  geom_text(
    data = mylabels,
    aes(x, y, label = label),
    size = 10,
    hjust = 0,
    vjust = 1,
    color = qaad_colors("rufous_red")
  ) +
  labs(
    x = "Distance from central place (km)",
    y = "Time in Patch (hours)"
  ) 

```

]

.pull-right[

A **piecewise constant** (or step) function models differences in intercept.  

Unfortunately, these do not capture variation within each region.  

]

---

## Piecewise Polynomial Functions

.pull-left[

```{r}

patch_cuts <- patch %>% 
  mutate(
    cuts = cut(distance, breaks = c(-Inf, 2.5, 5, 7.5, Inf))
  ) %>% 
  nest_by(cuts) %>% 
  mutate(
    model = list(lm(time ~ poly(distance, 3), data = data)),
    estimate = list(unname(predict(model)))
  ) %>% 
  select(-model) %>% 
  unnest(everything())

ggplot(patch) +
  geom_point(
    aes(distance, time),
    size = 4,
    color = "gray25", 
    alpha = 0.75
  ) +
  geom_vline(
    xintercept = c(2.5, 5, 7.5),
    linetype = "dashed"
  ) +
  geom_line(
    data = patch_cuts,
    aes(distance, estimate, group = cuts),
    size = 1.5,
    color = qaad_colors("rufous_red")
  ) +
  geom_text(
    data = mylabels,
    aes(x, y, label = label),
    size = 10,
    hjust = 0,
    vjust = 1,
    color = qaad_colors("rufous_red")
  ) +
  labs(
    x = "Distance from central place (km)",
    y = "Time in Patch (hours)"
  ) 

```

.right[_Piecewise cubic polynomial_]

]

.pull-right[

A **piecewise polynomial** function models differences in slope and intercept.

Unfortunately, these are not continuous.

]

---

## Natural Spline

.pull-left[

```{r}

fit <- lm(
  time ~ ns(distance, knots = c(2.5, 5, 7.5)),
  data = patch
)

patch_cuts <- tibble(
  distance = seq(0, 10, length = 200)
) %>% 
  mutate(
    estimate = predict(fit, newdata = tibble(distance))
  )

ggplot(patch) +
  geom_point(
    aes(distance, time),
    size = 4,
    color = "gray25", 
    alpha = 0.75
  ) +
  geom_vline(
    xintercept = c(2.5, 5, 7.5),
    linetype = "dashed"
  ) +
  geom_line(
    data = patch_cuts,
    aes(distance, estimate),
    size = 1.5,
    color = qaad_colors("rufous_red")
  ) +
  geom_text(
    data = mylabels,
    aes(x, y, label = label),
    size = 10,
    hjust = 0,
    vjust = 1,
    color = qaad_colors("rufous_red")
  ) +
  labs(
    x = "Distance from central place (km)",
    y = "Time in Patch (hours)"
  ) 

```

.right[_Natural cubic spline_]

]

.pull-right[

A **spline** is a piecewise polynomial function subject to the constraint of continuity.

To ensure continuity, splines use a **truncated power** transform on each successive knot, $k$:

$$f(X, k)_{+}^{d} = \left\{\begin{array}{ c l } (X-k)^d & if\; X > k; \\ 0 & otherwise \end{array} \right.$$

A **natural spline** includes an additional constraint to reign in variance at the boundaries (in regions A and D in the example). It must be linear in these areas.  

Note that each constraint returns a degree of freedom!

]


---

## Smooth Spline

.pull-left[

```{r}

fit <- with(patch, smooth.spline(distance, time))

patch_cuts <- tibble(
  distance = seq(0, 10, length = 200)
) %>% 
  mutate(
    estimate = predict(fit, distance)$y
  )

ggplot(patch) +
  geom_line(
    aes(distance, time),
    color = "gray80",
    size = 0.5
  ) +
  geom_point(
    aes(distance, time),
    size = 4,
    color = "gray25", 
    alpha = 0.75
  ) +
  geom_line(
    data = patch_cuts,
    aes(distance, estimate),
    size = 1.5,
    color = qaad_colors("rufous_red")
  ) +
  labs(
    x = "Distance from central place (km)",
    y = "Time in Patch (hours)"
  ) 

```

.right[_Cubic spline_]

]

.pull-right[

A **smoothing spline** is actually a "LOSS+PENALTY" model fitting procedure  

$$-\mathcal{l} + (\lambda\cdot roughness)$$

where  
- the first term is the negative log-Likelihood of a model fit using a spline with an arbitrarily large number of knots and

- the second terms is a __roughness penalty__, with $\lambda$ being a _tuning parameter_ that scales the penalty.

Note that increasing $\lambda$ is like adding constraints in that it reduces model complexity, in this case the **effective degrees of freedom**.

]

---

## Generalized Additive Models

.pull-left[

```{r}

fit <- gam(time ~ s(distance), data = patch)

patch_cuts <- tibble(
  distance = seq(0, 10, length = 200)
) %>% 
  mutate(
    estimate = predict(fit, newdata = tibble(distance))
  )

ggplot(patch) +
  geom_point(
    aes(distance, time),
    size = 4,
    color = "gray25", 
    alpha = 0.75
  ) +
  geom_line(
    data = patch_cuts,
    aes(distance, estimate),
    size = 1.5,
    color = qaad_colors("rufous_red")
  ) +
  labs(
    x = "Distance from central place (km)",
    y = "Time in Patch (hours)"
  ) 

```

]

.pull-right[

A **Generalized Additive Model** or GAM is a method for fitting multiple splines.

$$g(E(y_i)) = \beta_0 + \sum_{j=1}^{n} f_j(x_{ij})$$

- Extension to a GLM, so it can model non-normal responses.
- Called an _additive_ model because it calculates a separate spline for each term and adds them together.

]

---

## Generalized Additive Models

.pull-left[

```{r}

fit <- gam(time ~ s(distance, by = sex), data = patch)

patch_cuts <- tibble(
  distance = rep(seq(0, 10, length = 200), 2)
) %>% 
  mutate(
    sex = rep(c("female", "male"), each = n()/2),
    estimate = predict(fit, newdata = tibble(distance, sex))
  )

ggplot(patch) +
  geom_point(
    aes(distance, time, color = sex),
    size = 4, 
    alpha = 0.75
  ) +
  geom_line(
    data = patch_cuts,
    aes(distance, estimate, color = sex),
    size = 1.5
  ) +
  scale_color_viridis(
    name = NULL,
    discrete = TRUE
  ) +
  labs(
    x = "Distance from central place (km)",
    y = "Time in Patch (hours)"
  ) + 
  theme(
    legend.position = c(0.03, 0.95),
    legend.justification = c("left", "top"),
    legend.text = element_text(size = 18)
  )

```

]

.pull-right[

Can incorporate all the nice features of a GLM, including interaction terms and mixed-effects.

```{r}

broom::tidy(fit) %>% 
  rename("t.statistic" = statistic) %>% 
  kbl(table.attr = "class='table-model table-fullwidth'") %>% 
  kable_paper(c("striped", "hover"))

```

]

---

## &#x1F52D; Looking Ahead

```{r}

schedule <- here("_misc", "course_outline.xlsx") %>% 
  readxl::read_excel(1) %>% 
  mutate(
    date = lubridate::ymd(date),
    description = glue::glue(
    "<b>{topic}</b> <span style = 'font-size: 80%; color: #9C9C9C;'>({date})</span><br>
    <div style = 'font-size: 80%; font-style: italic; color: #9C9C9C;'>{details}</div>"
  )) %>% 
  select(description)

```

.pull-left[

```{r}

schedule %>% 
  slice(14:16) %>% 
  kbl(escape = FALSE, col.names = NULL) %>% 
  kable_paper(c("striped", "hover"))

```

]