---
title: "Lab 03"
subtitle: "Statistical Inference"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float:
      smooth_scroll: false
    self_contained: false
    css: custom_style.css
    includes:
      in_header: toggle.html
    theme: 
      bootswatch: sandstone
---

```{r, include = FALSE, code=xfun::read_utf8(here::here("exercises", "before_chunk.R"))}
```


## Outline {#Outline}

__tl;dr__ wrangling data and evaluating models. 

### Objectives {.objectives}

This lab will guide you through the process of  

1. loading (or "attaching") R packages with `library()`
2. generating summary statistics for your data
2. visualizing data with the grammar of graphics (gg)
    - aesthetic mappings
    - geometric objects
    - facets
    - scales
    - themes


### R Packages

We will be using the following packages:

- [archdata](https://cran.r-project.org/web/packages/archdata/index.html)
- [palmerpenguins](https://allisonhorst.github.io/palmerpenguins/)
- [skimr](https://docs.ropensci.org/skimr/)
- [ggplot2](https://ggplot2.tidyverse.org/index.html)
- [viridis](https://sjmgarnier.github.io/viridis/)

To install these packages, run the following code in your console:

```{r, eval = FALSE}

install.packages(
  c("archdata", "palmerpenguins", "skimr", "ggplot2", "viridis")
)

```

```{r, echo = FALSE}

library(khroma)
library(palmerpenguins)
library(skimr)

```


### Data

- `penguins` 
    - Includes measurements for penguin species, island in Palmer Archipelago, size (flipper length, body mass, bill dimensions), and sex. 
    - package: `palmerpenguins`
    - reference: <https://allisonhorst.github.io/palmerpenguins/reference/penguins.html>
- `DartPoints`
    - Includes measurements of 91 Archaic dart points recovered during surface surveys at Fort Hood, Texas.
    - package: `archdata`
    - reference: <https://cran.r-project.org/web/packages/archdata/archdata.pdf>
    
    
## Homework {#Homework}