---
title: "Lab 08"
subtitle: "Linear Model Extensions"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float:
      smooth_scroll: false
    self_contained: false
    css: custom_style.css
    includes:
      in_header: toggle.html
    theme: 
      bootswatch: sandstone
---

```{r, include = FALSE, code=xfun::read_utf8(here::here("exercises", "before_chunk.R"))}
```


## Outline {#Outline}

__TL;DR__  

__Caution!__ Please note that all labs assume that you are working in an RStudio Project directory!


### Objectives

This lab will guide you through the process of  

1. 


### R Packages

We will be using the following packages:

- [dplyr](https://dplyr.tidyverse.org/)
- [palmerpenguins](https://allisonhorst.github.io/palmerpenguins/)

To install these packages, __run the following code in your console__:

```{r, eval = FALSE}

install.packages(
  c("dplyr", "palmerpenguins")
)

```

__Note:__ You should not `install.packages()` in an Rmd document. Use that function in your R console instead. Then use `library()` as part of the preamble in your Rmd document to check packages out of the library and use them in that R session. This should always go at the start of your document!

```{r}

library(dplyr)
library(palmerpenguins)

```


### Data

- `penguins` 
    - Includes measurements for penguin species, island in Palmer Archipelago, size (flipper length, body mass, bill dimensions), and sex. 
    - package: `palmerpenguins`
    - reference: <https://allisonhorst.github.io/palmerpenguins/reference/penguins.html>


## Gaussian GLM {#Gaussian_GLM}

## Binomial GLM {#Binomial_GLM}

## Poisson GLM {#Poisson_GLM}
    
## Homework {#Homework}

No homework this week! Enjoy your spring break!