---
title: "Lab 13"
subtitle: "Generalized Additive Models"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float:
      smooth_scroll: false
    self_contained: false
    css: custom_style.css
    includes:
      in_header: toggle.html
    theme: 
      bootswatch: sandstone
---

```{r}
#| include = FALSE,
#| code = xfun::read_utf8(here::here("exercises", "before_chunk.R"))
```

## Outline {#Outline}

**TL;DR** Non-linear responses!

**Caution!** Please note that all labs assume that you are working in an RStudio Project directory!

### Objectives

This lab will guide you through the process of

1. Fitting a Generalized Additive Model
2. Prediction with a GAM
2. Fitting a GAM with an interaction term

### R Packages

We will be using the following packages:

- [dplyr](https://dplyr.tidyverse.org/)
- [ggeffects](https://strengejacke.github.io/ggeffects/index.html)
- [ggplot2](https://ggplot2.tidyverse.org/index.html)
- [here](https://here.r-lib.org/)
- [mgcv](https://cran.r-project.org/web/packages/mgcv/mgcv.pdf)
- [performance](https://easystats.github.io/performance/)
- [skimr](https://docs.ropensci.org/skimr/)
- [viridis](https://sjmgarnier.github.io/viridis/)

To install these packages, **run the following code in your console**:

```{r, eval = FALSE}

install.packages(
  c("dplyr", "ggeffects", "ggplot2", "here", "mgcv", "performance", "skimr", "viridis")
)

```

**Note:** You should not `install.packages()` in an Rmd document. Use that function in your R console instead. Then use `library()` as part of the preamble in your Rmd document to check packages out of the library and use them in that R session. This should always go at the start of your document!

```{r}

library(dplyr)
library(ggeffects)
library(ggplot2)
library(here)
library(mgcv)
library(performance)
library(skimr)
library(viridis)

```

### Data

- `patch-residence`
    - A hypothetical dataset including distance from a central-place (in km) and time in patch (in hours).
    - package: NA
    - reference: <https://github.com/kbvernon/qaad/tree/master/datasets>

## GAM {#GAM}

In this section, we'll learn how to fit a Generalized Additive Model (or GAM) using the `gam()` function from the `mgcv` package. Here, we'll be using the `patch-residence` dataset to answer the following question:

**Question** Do central-place foragers spend more time in patches some intermediate distance from a central-place when within-group competition and predation threat are both high?

For the sake of this exercise, we will assume that within-group competition is high closer to a central-place (because of population density) and decreases with distance. Conversely, predation threat is low near a central-place because of shared vigilance and increases with distance. Assuming that all patches are relatively equal in quality, this should lead individual foragers to invest most of their foraging effort into patches some intermediate distnace from a central place.

```{r}
#| echo = FALSE

sigmoid <- function(x, x0) 1 / (1 + exp(-(x-x0)))

assumptions <- tibble(
  distance = seq(0, 10, length = 100)
) %>% 
  mutate(
    "Within-Group Competition" = 1 - sigmoid(distance, 4),
    "Predation Threat" = sigmoid(distance, 6),
    "Patch Quality" = 0.75
  ) %>% 
  pivot_longer(
    cols = -distance, 
    names_to = "variable",
    values_to = "value"
  ) %>% 
  select(variable, distance, value) %>% 
  arrange(variable, distance) %>% 
  mutate(
    variable = factor(variable),
    variable = fct_relevel(variable, "Within-Group Competition", "Predation Threat")
  )
mylabels <- tibble(
  x = c(0, 10, 10),
  y = c(0.05, 0.72, 0.05),
  label = c("Predation\nThreat", "Patch Quality", "Within-Group\nCompetition"),
  color = c("Predation Threat", "Patch Quality", "Within-Group Competition")
)
ggplot() +
  geom_line(
    data = assumptions,
    aes(distance, value, color = variable),
    size = 2
  ) +
  geom_text(
    data = mylabels,
    aes(x, y, label = label, color = color),
    hjust = c(0, 1, 1),
    vjust = c(0, 1, 0),
    size = 7,
    lineheight = 0.85
  ) +
  scale_color_viridis(
    end = 0.97,
    discrete = TRUE
  ) +
  labs(
    x = "Distance from Central-Place",
    y = NULL
  ) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  )

remove(sigmoid, assumptions, mylabels)

```

To test this model, let's assume we go out and do focal follows with a hypothetical group of hunter-gatherers in an area known to have relatively high predation threat and relatively high population density at a central-place. You can download this dataset with the following code.

```{r}
#| echo = FALSE

patch <- here("datasets", "patch-residence.csv") %>% read_csv()

```

```{r}
#| eval = FALSE

download.file(
  "https://raw.githubusercontent.com/kbvernon/qaad/master/datasets/patch-residence.csv",
  destfile = here("data", "patch-residence.csv")
)

patch <- here("data", "patch-residence.csv") %>% 
  read.csv() %>% 
  as_tibble()

skim(patch)

```

```{r}
#| echo = FALSE

skim_without_charts(patch)

```

<br>

Notice that there are three variables in this dataset, including `sex` (a factor level term), `distance` (for distance from a central place, measured in kilometers), and `time` (for time in patch, measured in hours). We'll return to the `sex` variable in the next section. For now, let's visualize the relationship between time in patch and distance from a central place using a scatterplot.

```{r}
#| echo = FALSE

ggplot2::theme_set(ggplot2::theme_gray(14))

```


```{r}

ggplot(patch) +
  geom_point(
    aes(distance, time),
    size = 4,
    color = "gray25", 
    alpha = 0.75
  ) +
  labs(
    x = "Distance from central place (km)",
    y = "Time in Patch (hours)"
  ) +
  theme_bw(14) +
  theme(
    panel.grid = element_blank()
  )

```

As you can see, it looks like there is a non-linear trend here. Just to be sure, though, let's have a look at a simple linear model and plot the diagnostics for it.

```{r}

patch_lm <- lm(time ~ distance, data = patch)

summary(patch_lm)

```

Looks like `distance` is almost, but not-quite significant in this model. Let's check some diagnostic plots with `check_model()` from the `performance` package.

```{r}
#| fig.asp = 1.3

check_model(patch_lm)

```

Here, we can see that the residuals are normally distributed (a good sign). However, the linearity check shows an extremely non-linear relationship, and there's a lot of heteroscedasticity. Let's see if we can address the non-linearity issue with a Generalized Additive Model (GAM).

The function call for a GAM is very similar to that for a GLM. We still want to specify the family distribution for the conditional response and link function, and we still supply it with a dataset. The only difference is that we need to specify a basis function (in particular, a spline) in the model formula. This is done using the `s()` function (though there are alternatives to this that we won't cover here). It looks like this:

```{r}
#| eval = FALSE

time ~ s(distance, bs = "tp")

```

Note the parameter `bs` here. This specifies the type of spline, in particular the type of constraint or penalty used. The default is "tp," short for "thin plate regression spline." The nature of this spline is beyond the scope of this lab. For now, just note that it uses a sophisticated procedure for penalizing overly complex splines or splines that tend to over-fit the data. In the simplest cases, like the one we are currently investigating, this spline is sufficient.

So much for the model formula. Now, let's turn to the entire `gam()` function call. It looks like this:

```{r}

patch_gam <- gam(
  time ~ s(distance, bs = "tp"),
  family = gaussian(link = "identity"),
  data = patch
)

summary(patch_gam)

```

Notice that the summary provides all the familiar diagnostics of a GLM, including coefficient estimates for the parameteric terms (in this case, that's just the intercept), the R-squared value, and the deviance explained. It also provides a measure of the expected degrees of freedom (EDF) for the smooth terms, as well as a test of its significance. 

How do we interpret EDF? Here are three general rules-of-thumb:

1. If EDF = 1, then the hypothesized relationship between the variable and the response is linear.
2. If 1 < EDF < 2, then the hypothesized relationship is _weakly_ non-linear.
3. If EDF > 2, then the hypothesized relationship is _strongly_ non-linear.

## GAM Prediction {#GAM_Prediction}

As with GLMs, GAM has an exponential family distribution for the error term and a link function that makes the response linear in the predictors. To generate predictions on the level of the response, we must, therefore, use the inverse link function, which we can extract in the familar way with:

```{r}
#| eval = FALSE

inverse_link <- family(<<gam>>)$linkinv

```

In the case of the model we just created, however, we used the Gaussian (or normal) distribution with the identity link, so there's no real transformation happening. We can just run predict and call it good. For most cases, it is also probably overkill to try and do this all by hand. Though it's important that you understand that it's an issue! 

At any rate, you are probably better off relying on another R package to handle this. There are two ways to do this. First, the `mgcv` package provides a generic `plot()` function specifically to visualize marginal responses in GAM models. Simply pass it the model object and it will generate the plot for you.

```{r}

plot(patch_gam)

```

This will give you the correct shape of the response curve, but it won't be on the scale of the response. It's actually centered on zero for reasons having to do with the constraints imposed on the spline, but that's getting too far into the weeds. For now, I would recommend that you use this method as a rough and ready way to explore the estimated non-linear trend.

For creating publication-quality graphics, I recommend using a package like `ggeffects` because of its ease of use (and consistency). The basic idea here is to generate a table of data, including model estimates, that you can then use to plot the estimated non-linear trend. To do that, we use the `ggpredict()` function, passing it the model object.

```{r}

estimate <- ggpredict(patch_gam, terms = "distance")

# convert to tibble for more useful printing
estimate <- as_tibble(estimate)

estimate

```

By default, this back-transforms the predictions to the response scale (using the inverse link function!) and adds the upper and lower confidence intervals (plus or minus two times the standard error). Thus, we have everything we need to visualize the estimated trend. Let's do that now with `ggplot()`.

```{r}

ggplot() +
  geom_point(
    data = patch,
    aes(distance, time),
    size = 3,
    color = "gray35",
    alpha = 0.75
  ) +
  geom_ribbon(
    data = estimate,
    aes(x, ymin = conf.low, ymax = conf.high),
    fill = "gray75",
    alpha = 0.3
  ) +
  geom_line(
    data = estimate,
    aes(x, predicted),
    color = "#A20000",
    size = 1.3
  ) +
  labs(
    x = "Distance from central-place (km)",
    y = "Time in patch (hours)"
  ) +
  theme_bw(14) +
  theme(
    panel.grid = element_blank()
  )

```

And voila! Our estimated non-linear trend visualized.

## GAM With Interaction {#GAM_With_Interaction}

GAMs are GLMs with splines for each term. These are added together, hence being an additive model. Because of this, GAMs can do everything GLMs can do, but for non-linear trends, too. That includes interactions. Because we are working with splines, however, these interactions are specified somewhat differently as the interactions have to be accounted for each knot-separated region of the covariate. So, instead of adding them as their own additive or multiplicative term in the model formula, we specify them within each call to `s()`. For example, in our data, we have information about the sex of each forager. To test whether these groups have differences in the times they spend in a patch at each distance, we add them to the model formula like so

```{r}
#| eval = FALSE

time ~ s(distance, bs = "tp", by = sex)

```

Note that you do not have to escape the `sex` variable with quote marks.

There's one catch to all this, of course. There always is, isn't there? The `s()` and `gam()` functions want the factor variable as a literal factor in R, not just a character variable as it is by default. That means, we need to convert it to a factor first, otherwise we will get this error:

```{r}
#| error = TRUE

patch_gam_bysex <- gam(
  time ~ s(distance, by = sex), 
  family = gaussian(link = "identity"),
  data = patch
)

```

It can't find the by variable because it's looking for a factor, not a character. To fix this, we will just mutate that variable.

```{r}

patch <- patch %>% mutate(sex = factor(sex))

patch

```

Now it should work.

```{r}

patch_gam_bysex <- gam(
  time ~ s(distance, by = sex), 
  family = gaussian(link = "identity"),
  data = patch
)

summary(patch_gam_bysex)

```

And, lo, it's a significant difference! But, how do we visualize this? The answer to that is still `ggpredict()`, albeit with a subtle change.

```{r}

estimate <- ggpredict(patch_gam_bysex, terms = c("distance", "sex"))

# convert to tibble for more useful printing
estimate <- as_tibble(estimate)

estimate

```

Here, we included "sex" as the second value (after "distance") passed to "terms." You can read this as saying, "predict time in patch as a function of distance _for each sex_." Notice that the group parameter now includes sex for each estimate. 

Now, we can plot this with `ggplot()`, only this time we will pass the `group` variable in the `estimate` table and the `sex` parameter in the `patch` table as aesthetic parameters to `color` and `fill`. Note that I am moving the legend to the top left corner of the plot panel using various arguments passed to `theme()`.

```{r}

ggplot() +
  geom_point(
    data = patch,
    aes(distance, time, color = sex),
    size = 3,
    alpha = 0.75
  ) +
  geom_ribbon(
    data = estimate,
    aes(x, ymin = conf.low, ymax = conf.high, fill = group),
    alpha = 0.1
  ) +
  geom_line(
    data = estimate,
    aes(x, predicted, color = group),
    size = 1.3
  ) +
  scale_color_viridis(
    name = NULL,
    discrete = TRUE
  ) +
  scale_fill_viridis(
    name = NULL,
    discrete = TRUE
  ) +
  labs(
    x = "Distance from central-place (km)",
    y = "Time in patch (hours)"
  ) +
  theme_bw(14) +
  theme(
    legend.position = c(0.03, 0.97), 
    legend.justification = c("left", "top"),
    legend.text = element_text(size = 16),
    panel.grid = element_blank()
  )

```

I'll leave it to you to decide what differences in costs and benefits might be driving this difference in patch-residence time and distance in the face of predation threat.

## Homework {#Homework}

No homework this week!
